\documentclass{ximera}

\input{../../preamble.tex}

\title{Logarithmic}

\begin{document}

\begin{abstract}
reciprocal
\end{abstract}
\maketitle












\textbf{\textcolor{red!70!darkgray}{$\blacktriangleright$ A Peek ahead to Calculus}}





\textbf{\textcolor{blue!55!black}{In Calculus}}, we will see all of the rules for obtaining the derivative of any funciton for ourselves. \\

Right now we know three rules:

\begin{itemize}
	\item if $f(x) = a \, x^2 + b \, x + c$, then $f'(x) = 2a \, x + b$
	\item if $f(x) = a \, x + b$, then $f'(x) = a$
	\item if $f(x) = a$, then $f'(x) = 0$
\end{itemize}



When we get to Calculus, we'll be able to ``differentiate'' any function we want. \\

Right now, you would be given the derivative of a function, which you could then use to analyze the function. \\



\begin{example}

When analyzing $g(t) = t - \ln(t)$, you might be given the derivative.


\[
g'(t) = 1 + \frac{1}{t} 
\]


We could then use this derivative to identify critical numbers and decide where the function $g$ is increasing and decreasing.




\textbf{Critical Numbers} \\


\textbf{Note:}. The domain of $g$ is $(0, \infty)$. \\



To find critical numbers, we need to solve $g'(t) = 0$. \\


\[
g'(t) = 0
\]


\[
1 - \frac{1}{t}  = 0
\]


\[
1  = \frac{1}{t}
\]





$\blacktriangleright$ $1  = \frac{1}{t}$, when $t = 1$. \\



$g$ has $1$ as its only critical number. \\




\textbf{Behavior}: Increasing and Decreasing \\



The sign of the derivative will tell us where $g$ is increasing and decreasing. \\


\textbf{\textcolor{blue!55!black}{$\blacktriangleright$}} Where is $1  = \frac{1}{t}$ positive and negative?





\[
g'(t) < 0 \, \text{ on } \, (0, 1)
\]


\[
g'(t) > 0 \, \text{ on } \, (1, \infty)
\]





\[
g(t)  \, \text{ decreases on } \, \left( -\infty, \frac{5}{12} \right)(0, 1)
\]


\[
g(t)  \, \text{ increases on } \, (1, \infty)
\]



Since $g$ is continuous, this tells us that there is a local (possibly global) minimum at $1$.


\end{example}














\begin{idea}


We know logarithmic functions as an inverse to exponential functions. 



\begin{center}


$\log_A(B)$ is the thing you raise $A$ to, to get $B$. \\


\end{center}

In particular, $\ln(x)$ is the inverse to $e^x$, meaning their composition gives the identiy function.



\[
e^{\ln(x)} = x
\]


\[
\ln\left( e^x \right) = x
\]

If we were in Calculus, the \textbf{chain rule} will tell us how their derivatives relate to each other.




\[
\frac{d}{dx} \ln(x) = \frac{1}{x}
\]



\end{idea}


We need some more experience with limits to see why this is true. Expereince which Calculus will give us. \\


For now, you would need to be given such derivatives.  You don't know how to get these derivative yet.











\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Rate of Change]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/rateOfChange/examples/exampleList}

\end{center}






\end{document}
