\documentclass{ximera}

\input{../../preamble.tex}

\title{Rate Of Change}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}
\maketitle







As we have seen, there are a couple of viewpoints to functions that drive our investigations.

$\blacktriangleright$ One is a \textbf{\textcolor{purple!85!blue}{point-wise}} view.  What is the function value at each domain number? \\


$\blacktriangleright$ The other is a \textbf{\textcolor{purple!85!blue}{behavioral}} view.  How are the function values changing? \\



We use these viewpoints when analyzing individual functions.  We use these viewpoints when comparing two functions. When comparing functions, it is not enough to compare their values at each domain number.  There are too many.  We need to move faster.  We need to predict what is going to happen. That is where rates-of-change come in. \\




We would like rates-of-change on intervals.  We would also like a pointwise viewpoint of rates-of-change. \\



A rate-of-change at a point seems like a contradiction, but fits in with the story of tangent lines very well.













\subsection{Learning Outcomes}


\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item investigate rate of change.

\end{itemize}
\end{sectionOutcomes}














\begin{center}
\textbf{\textcolor{green!50!black}{ooooo=-=-=-=-=-=-=-=-=-=-=-=-=ooOoo=-=-=-=-=-=-=-=-=-=-=-=-=ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Rate of Change]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/rateOfChange/examples/exampleList}

\end{center}







\end{document}
