\documentclass{ximera}

\input{../../preamble.tex}

\title{Function Forms}

\begin{document}

\begin{abstract}
%
\end{abstract}
\maketitle



Our goal in Calculus is anlayzing functions.

The very first step in analyzing a funciton is identifying the type of function you have. We have a preference to our categories. \\

We first decide if we have an elementary function. If not, then we decide if we are looking at one our four traditional operations.  If not, then it must be a composition.\\



\textbf{\textcolor{blue!55!black}{Choice 1:}}
\begin{itemize}
\item Constant
\item Linear
\item Quadratic
\item Polynomial
\item Rational
\item Exponential
\item Logarithmic
\item Root or Radical
\item Absolute Value
\item Trigonometric
\end{itemize}



\textbf{\textcolor{blue!55!black}{Choice 2:}}
\begin{itemize}
\item Sum
\item Difference
\item Product
\item Quotient
\end{itemize}



\textbf{\textcolor{blue!55!black}{Choice 3:}}
\begin{itemize}
\item Composition
\end{itemize}




Of course, it could also be a piecewise defined function.  Those are usually easy to spot.
\begin{itemize}
\item Piecewise Defined
\end{itemize}







\subsection*{Learning Outcomes}


\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item identify function forms.
\end{itemize}
\end{sectionOutcomes}












\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Function Forms]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/functionForm/examples/exampleList}

\end{center}




\end{document}
