\documentclass{ximera}

\input{../../preamble.tex}

\author{Jim Talamo}
\license{Creative Commons 3.0 By-NC}


\outcome{Set up an integral or sum of integrals that gives the volume of a solid with known cross sections.}
\outcome{Explore how types cross sections affects the volume.}

\begin{document}
\begin{exercise}


The region $R$ in the first quadrant is bounded by $y=\frac{12}{x}$, $x=1$ and $x=3$.  This exercise considers three solids whose base is $R$, but whose cross sections differ from each other. 





For Solid 1, cross sections through the solid taken parallel to the $y$-axis are squares.

For Solid 2, cross sections through the solid taken parallel to the $y$-axis are equilateral triangles.

For Solid 3, cross sections through the solid taken parallel to the $y$-axis are semicircles.

\begin{exercise}
Using geometric reasoning only (i.e. without performing any calculations):

Which solid has the greatest volume?
\begin{multipleChoice}
\choice[correct]{Solid 1}
\choice{Solid 2}
\choice{Solid 3}
\end{multipleChoice}

Which solid has the least volume?
\begin{multipleChoice}
\choice{Solid 1}
\choice{Solid 2}
\choice[correct]{Solid 3}
\end{multipleChoice}

\begin{hint}
How do the areas of a square of side length $s$, and semicircle with diameter $s$, and an equilateral triangle with side length $s$ compare?  How can this be used to compare the volumes of the three given solids?
\end{hint}

\end{exercise}

\begin{exercise}
To verify your observations quantitatively, calculate the volume of each solid.

The volume of Solid 1 is $\answer{96}$ cubic units.
	
The volume of Solid 2 is $\answer{24 \sqrt{3}}$ cubic units.

The volume of Solid 3 is $\answer{12 \pi }$ cubic units.

\end{exercise}


\end{exercise}

\end{document}