\documentclass{ximera}

\input{../../preamble.tex}

\title{Function Algebra}

\begin{document}

\begin{abstract}

\end{abstract}
\maketitle



With the caveat that domains need to be aligned, we have many operations on functions that make up an arithmetic on functions.




$\blacktriangleright$ \textbf{Addition} 

The sum of two functions is again a function. The inverse of addition is subtracton. 

The additive identity is $Zero(x) = 0$ for all real numbers.




$\blacktriangleright$ \textbf{Mulitplication} 

The product of two functions is again a function. The inverse of mulitplication is division. 

The mulitplicative identity is $One(x) = 1$ for all real numbers.



$\blacktriangleright$ \textbf{Composition} 

Composition is a new operation on functions.

The composition of two functions again a function. 

The identity function, $I(x) = x$ for all $x$, is the identity element with respect to composition.



\[   f \circ I = f    \, \text{ and } \, I \circ f = f        \]



The inverse of $f$ with respect to composition is another function whose composition with $f$ produces the identity function.

The symbol for the inverse of $f$ is $f^{-1}$.

\[   f \circ f^{-1} = I    \, \text{ and } \, f^{-1} \circ f = I       \]






\begin{example} Inverse Function


Let $H(t) = 3t - 5$. \\

Then $H^{-1}(x)$ is a function such that $H \circ H^{-1} - I$ \\


We are looking for a function, $H^{-1}$, such that $H(H^{-1})(k) = 3 H^{-1}(k) - 5 = k$ \\



$H$ is linear and $I$ is linear, $H^{-1}$ should be linear.

Let's start building.  $H^{-1}(x) should look like $a \cdot x + b$.  $The leading coefficient of $H^{-1}$, $a$, should be $\frac{1}{3}$.


$H^{-1}(x) =\frac{1}{3} x + b$


That would gets us $H(H^{-1})(k) = 3 (\frac{1}{3}x + b) - 5 = x + 3b - 5 $


We need $3b - 5 = 0$ or $b = \frac{5}{3}$.


$H^{-1}(x) =\frac{1}{3} x + \frac{5}{3}$



\end{example}






























\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item work with function arithmetic.
\item function algebra.
\item .
\item .
\item .
\end{itemize}
\end{sectionOutcomes}

\end{document}
