\documentclass{ximera}

\input{../../preamble.tex}

\title{Polar Graphs}

\begin{document}

\begin{abstract}
%%%
\end{abstract}
\maketitle





We have two ways of describing positions on the Cartesian Plane.


$\blacktriangleright$ \textbf{Rectangular Coordinates} provide left/right and up/down measurements.





$\blacktriangleright$ \textbf{Polar Coordinates} provide direction and radius measurements.




We can describe Complex numbers with either coordinate system.


We can describe curves with either coordinate system, however, some curves are easier to describe with one coordinate system or the other.




\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item use polar coordinates to describe curves.
\item convert between polar and Cartesian coordinates.
\item convert between the Cartesian and polar reprentation of a curve.
\item determine whether different polar representations represent the same point in the $(x,y)$-plane.
\item use the Cartesian to polar method to plot polar graphs.
\item understand the difference between a curve and the choice of coordinates used to describe the curve
\end{itemize}
\end{sectionOutcomes}











\begin{center}
\textbf{\textcolor{green!50!black}{ooooo=-=-=-=-=-=-=-=-=-=-=-=-=ooOoo=-=-=-=-=-=-=-=-=-=-=-=-=ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Polar Graphs]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/polarGraphing/examples/exampleList}

\end{center}







\end{document}

