\documentclass{ximera}

\input{../../preamble.tex}

\title{Absolute Value}

\begin{document}

\begin{abstract}
traits
\end{abstract}
\maketitle









\subsection*{The Core Absolute Value Function}

The core absolute value function is a piecewsie defined function.


\[
|k| = 
\begin{cases}
  -k & \text{if $k<0$,}\\
   k & \text{if $k\ge 0$}.
\end{cases}
\]








Graph of $y = |k|$.


\begin{image}
\begin{tikzpicture} 
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$k$, ylabel=$y$,
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=2, penColor, smooth, samples=200, domain=(-9:9),<->] {abs(x)};
        

  \end{axis}
\end{tikzpicture}
\end{image}



The symbol for the core absolute value function are two vertical bars surrounding the domain value.

The domain is $(-\infty, \infty)$.

Its only zero is $0$.

It is continuous.

It is unbounded positively in both directions.

It decreases on $(-\infty, 0]$ and increases on $[0, \infty)$.

It has a global minimum of $0$ at $0$ and no global maximum.  This is also the only local extreme number.

Its range is $[0, \infty)$.



All other absolute value functions are compositions of linear functions with the Core absolute value function.












\subsection*{General Absolute Value Functions}

Absolute value functions are those functions that can be represented by formulas of the form


\[
A \, |B \, x + C| + D  \, \text{ with }  \, A \ne 0, B \ne 0
\]



These can be viewed as compositions.




\begin{itemize}
\item Let $L_{out}(v) = A \, v + D$ be a linear function.
\item Let $AV(t) = |t|$ be the Core absolute value function.
\item Let $L_{in}(u) = B \, u + C$ be a linear function.
\end{itemize}

Then, our general absolute value function can be expressed as a composition.


\[
 A \, |B \, x + C| + D = (L_{out} \circ AV \circ L_{in})(x)
\]



Now we can use the Chain Rule to establish whether an absolute value function increases or decreases.













\begin{definition} \textbf{\textcolor{green!50!black}{Absolute Functions}}

Absolute functions are those functions that \textbf{\textcolor{purple!85!blue}{can}} be represented by formulas of the form


\[      AV(x) = A \cdot | B \, x + C| + D   \]

where $A$, $B$, $C$, and $D$ are real numbers, with $A \ne B$ and $B \ne 0$ is a nonzero real number.


\end{definition}


\textbf{Note:}  In the template for absolute value functions, There is a leading coefficient for the function and there is a leading coefficient for both linear functions inside the absolute value bars and outside. 








\begin{example}  Absolute Value

Graph of $y = P(t) = -3|2t+4|+5$.


\begin{image}
\begin{tikzpicture} 
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$t$, ylabel=$y$,
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=2, penColor, smooth, samples=200, domain=(-4.33:0.33),<->] {-3*abs(2*x+4)+5};
        

  \end{axis}
\end{tikzpicture}
\end{image}

The formula:

\[
P(t) = -3|2t+4|+5 = 
\begin{cases}
  (-3)(-(2t+4))+5 & \text{if $t < \answer{-2}$,}\\
   -3(2t+4)+5 & \text{if $t \ge \answer{-2}$}.
\end{cases}
\]



\textbf{Domain:} The domain of every absolute value function is $(-\infty, \infty)$.

\textbf{Zeros:}

\begin{center}
$-3|2t+4|+5 = 0$

$|2t+4| = \frac{5}{3}$

Either $2t+4 = -\frac{5}{3}$ or $2t+4 = \frac{5}{3}$

Either $t = \frac{1}{2} \left( -\frac{5}{3}-4 \right)$ or $t = \frac{1}{2} \left(\frac{5}{3} - 4 \right)$

\end{center}





\textbf{Continuity:} Absolute value functions are continuous.

\textbf{End-Behavior:} The leading coefficient is negative, therefore $P$ becomes unbounded negatively in either direction.

\textbf{Behavior:} 

We can view $P$ as a composition:



\begin{itemize}
\item Let $AV(x) = |x|$ be the Core absolute value function.
\item Let $L_{in}(u) = 2 \, u + 4$, an increasing linear function.
\item Let $L_{out}(v) = -3 \, v + 5$, a decreasinglinear function.
\end{itemize}


\[ P(t) = -3|2t+4|+5 = (L_{out} \circ AV \circ L_{out})(t)\]


$|x|$ switches from decreasing to increasing at $0$, so we need to locate where $L_{in} = 0$.


$L_{in}(v) = 2 \, u + 4 = 0$ at $v = -2$



On $\left( -\infty,  \frac{5}{3} \right)$ the values of $L_{in} < 0$ and the core absolute value function is decreasing on $(-\infty, 0)$. 

On $\left( -\infty,  \frac{5}{3} \right)$, the Chain Rule gives

\[ dec \circ dec \circ inc = inc \]





On $\left( \frac{5}{3}, \infty \right)$ the values of $L_{in} > 0$ and the Core absolute value function is increasing on $(0, \infty)$. 

On $\left( \frac{5}{3}, \infty \right)$, the Chain Rule gives

\[ dec \circ inc \circ inc = dec \]


Just Checking: This agrees with the graph.


\textbf{Global Maximum and Minimum:}

Since the leading coefficient is negative, $P$ has no global minimum value.

$P$ switches from increasing to decreasing at $-2$, which means that $P(-2) = 5$ is global maximum value.


\textbf{Local Maximums and Minimums:}  $P(-2) = 5$ is the only local maximum value. There are no local minimum values.

\textbf{Range:} $P$ is continuous.  $5$ is its global maximum value. $P$ is unbounded negatively.  That gives a range of $(-\infty, 5]$.


















\end{example}

















\begin{summary}\textbf{\textcolor{red!70!yellow}{Absolute Value Analysis}}


\textbf{Domain:} The domain of every absolute function is $(-\infty, \infty)$.

\textbf{Zeros:} Absolulte value functions can have 2 zeros, 1 zero, or no zeros.

\textbf{Continuity:} Absolute value functions are continuous.

\textbf{End-Behavior:} In both directions absolute values become unbounded with the same sign.  The leading coefficient gives the sign.






\textbf{Behavior (Increasing and Decreasing):} Absolute value functions switch their behavior. The critical number is the real number that makes the inside equal to $0$.

The Chain Rule can tell you which side is increasing and which side is decreasing.

\textbf{Global Maximum and Minimum:}  Absolute value functions have either a global maximum or a global minimum.

\textbf{Local Maximums and Minimums:} The global extrema is also the only local extrema.

\textbf{Range:} The range of an absolute value function is either $(-\infty, D]$ or it is $[D, \infty)$.



\end{summary}




















\subsection*{Reading Coeffients}



\textbf{Behavior}


Formulas for absolute value functions look like

\[     abs(x) =    A  | B \, x + C | + D           \]

where $A \ne 0$, $B \ne 0$, $C$, and $D$ are real numbers.





The critical number for an absolute value function is the domain number that makes the inside of the absolute value bars equal to $0$.


\[
B \, x + C = 0
\]


The absolute value function increases on one side of the critical number and decreases on the other.  The sign of the leading coefficient determines which way it goes. 




\begin{itemize}
  \item If $A > 0$, then decreasing switching to increasing.
  \item If $A < 0$, then increasing switching to decreasing.
\end{itemize}










\textbf{End-Behavior}


The end-behavior is the same on both sides for an absolute value function. The sign of the leading coefficient determines which way it goes.


If the leading coefficent is positive, then 
\[
\lim_{x \to -\infty} abs(x) = \infty 
\]
 
\[
\lim_{x \to \infty} abs(x) = \infty 
\]







If the leading coefficent is negative, then 
\[
\lim_{x \to -\infty} abs(x) = -\infty 
\]
 
\[
\lim_{x \to \infty} abs(x) = -\infty 
\]












\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Analysis]{https://ximera.osu.edu/csccmathematics/precalculus/precalculus/radicalFunctions/examples/exampleList}

\end{center}





\end{document}
