\documentclass{ximera}

\input{../../preamble.tex}

\title{Inverse Trig Functions}

\begin{document}

\begin{abstract}
%
\end{abstract}
\maketitle




Sine, cosine, and tangent are functions. They connect angle measurements (degrees and radians) with ratios (numbers). Unfortunately, they are not one-to-one functions and do not have inverse functions. \\

But, we would like them to.\\

In order to get inverse functions, we need to make sine, cosine, and tangent into one-to-one functions.  To accomplish this, we'll simply restrict their domains. \\


This story will begin with a review of inverse functions in general.



We would like to analyze all of their characteristics and features. \\



\begin{itemize}
\item \textbf{\textcolor{red!80!black}{Domain}} 
\item \textbf{\textcolor{red!80!black}{Zeros}} 
\item \textbf{\textcolor{red!80!black}{Continuity}} 
  \begin{itemize}
     \item \textbf{\textcolor{purple!85!blue}{discontinuities}} 
     \item \textbf{\textcolor{purple!85!blue}{singularities}} 
  \end{itemize}
\item \textbf{\textcolor{red!80!black}{End-Behavior}} 
\item \textbf{\textcolor{red!80!black}{Behavior}} 
  \begin{itemize}
     \item \textbf{\textcolor{purple!85!blue}{intervals where increasing}} 
     \item \textbf{\textcolor{purple!85!blue}{intervals where decreasing}} 
  \end{itemize}
\item \textbf{\textcolor{red!80!black}{Global Maximum and Minimum}} 
\item \textbf{\textcolor{red!80!black}{Local Maximums and Minimums}} 
\item \textbf{\textcolor{red!80!black}{Range}} 
\item \textbf{\textcolor{blue!55!black}{...and we would like a nice graph}} 
\end{itemize}












\subsection*{Learning Outcomes}

\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item review inverse functions.
\item investigate inverse trigonometric functions.
\end{itemize}
\end{sectionOutcomes}









\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Inverse Trig Functions]{https://ximera.osu.edu/csccmathematics/precalculus/precalculus/inverseTrigFunctions/examples/exampleList}

\end{center}









\end{document}
