\documentclass{ximera}

\input{../../preamble.tex}

\title{Reciprocals}

\begin{document}

\begin{abstract}
zeros to singularities
\end{abstract}
\maketitle



\begin{idea} \textbf{\textcolor{blue!55!black}{Reciprocals}}


If $f(x)$ is a function, then its reciprocal is $\frac{1}{f(x)}$.

Reciprocal functions have ``opposite'' characteristics.




\begin{itemize}
\item where $f(x)$ is increasing, $\frac{1}{f(x)}$ is decreasing.
\item where $f(x)$ is decreasing, $\frac{1}{f(x)}$ is increasing.
\end{itemize}


\end{idea}






\subsection*{$\frac{1}{\sin(\theta)}$}


The reciprocal of sine is called cosecant, which is abbreviated as $\csc$.


\[
\csc(\theta) = \frac{1}{\sin(\theta)}
\]


Since $\sin(\theta)$ is periodic with a period of $2 \pi$, $\csc(\theta)$ is also periodic with a period of $2 \pi$.



\begin{itemize}
\item zeros of $\sin(\theta)$ become singularities of $\csc(\theta)$.
\item $\sin(\theta)$ has no singularities, so $\csc(\theta)$ has no zeros.
\end{itemize}




\begin{itemize}
\item $\sin(\theta)$ is increasing on $\left( 0,\frac{\pi}{2} \right)$ and $\left( \frac{3\pi}{2}, 2 \pi \right)$, which means that $\csc(\theta)$ is decreasing on $\left( 0,\frac{\pi}{2} \right)$ and $\left( \frac{3\pi}{2}, 2 \pi \right)$.
\item $\sin(\theta)$ is decreasing on $\left( \frac{\pi}{2}, \frac{3\pi}{2} \right)$, which means that $\csc(\theta)$ is increasing on $\left( \frac{\pi}{2}, \frac{3\pi}{2} \right)$.
\end{itemize}



The behavior tells us that $\csc(\theta)$ has a local minimum at $\frac{\pi}{2}$ and a local maximum at $\frac{3\pi}{2}$





Since $\lim\limits_{\theta \to 0} \sin(\theta) = 0$, we know that $\csc(\theta)$ is unbounded near $0$.  $0$ is a singularity of $\csc(\theta)$.

We also know that $\sin(\theta) > 0$ when $\theta$ is slightly positive.

That tells us that

\[
\lim\limits_{\theta \to 0^+} \csc(\theta) = \infty
\]



We also know that $\sin(\theta) < 0$ when $\theta$ is slightly negative.

That tells us that

\[
\lim\limits_{\theta \to 0^-} \csc(\theta) = -\infty
\]








\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{ithaiyrlee}{400}{300}
\end{center}






















\subsection*{$\frac{1}{\cos(\theta)}$}


The reciprocal of cosine is called secant, which is abbreviated as $\sec$.


\[
\sec(\theta) = \frac{1}{\cos(\theta)}
\]


Since $\cos(\theta)$ is periodic with a period of $2 \pi$, $\sec(\theta)$ is also periodic with a period of $2 \pi$.



\begin{itemize}
\item zeros of $\cos(\theta)$ become singularities of $\sec(\theta)$.
\end{itemize}




\begin{itemize}
\item $\cos(\theta)$ is decreasing on $(0, \pi)$, which means that $\sec(\theta)$ is increasing on $(0, \pi)$.
\item $\cos(\theta)$ is increasing on $(\pi, 2 \pi)$, which means that $\sec(\theta)$ is decreasing on $(\pi, 2 \pi)$.
\end{itemize}



The behavior tells us that $\sec(\theta)$ has a local minimum at $0$ and a local maximum at $\pi$.





Since $\lim\limits_{\theta \to \tfrac{\pi}{2}} \cos(\theta) = 0$, we know that $\sec(\theta)$ is unbounded near $\frac{\pi}{2}$.  $\frac{\pi}{2}$ is a singularity of $\csc(\theta)$.

We also know that $\cos(\theta) > 0$ when $\theta$ is slightly less than $\frac{\pi}{2}$.

That tells us that

\[
\lim\limits_{\theta \to \frac{\pi}{2}^-} \sec(\theta) = \infty
\]



We also know that $\cos(\theta) < 0$ when $\theta$ is slightly greater than $\frac{\pi}{2}$.

That tells us that

\[
\lim\limits_{\theta \to \frac{\pi}{2}^+} \sec(\theta) = -\infty
\]












\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{rdruru0ayb}{400}{300}
\end{center}

























\subsection*{$\frac{1}{\tan(\theta)}$}


The reciprocal of tangent is called cotangent, which is abbreviated as $\cot$.


\[
\cot(\theta) = \frac{1}{\tan(\theta)}
\]


Since $\tan(\theta)$ is periodic with a period of $\pi$, $\cot(\theta)$ is also periodic with a period of $\pi$.



\begin{itemize}
\item zeros of $\tan(\theta)$ become singularities of $\cot(\theta)$.
\end{itemize}




\begin{itemize}
\item $\tan(\theta)$ is an increasing function.
\item $\cot(\theta)$ is a decreasing function.
\end{itemize}


$\tan(\theta)$ has no maximums or minimums, so neither does $\cot(\theta)$.



Since $\cot(\theta)$ is a decreasing function we know that

\[
\lim\limits_{\theta \to 0^+} \cot(\theta) = \infty
\]




\[
\lim\limits_{\theta \to 0^-} \cot(\theta) = -\infty
\]












\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{vp4b0dqz9b}{400}{300}
\end{center}


Reciprocal functions have ``opposite'' or ``complementary'' characteristics.




\begin{warning} Reverse vs. Reciprocal


The reverse of a function and the reciprocal of a function are not the same thing.

They are completety and totally difference functions.


The \textbf{reverse} of a function switches the domain and range.


The \textbf{reciprocal} of a function uses the same domain, just taking the reciprocal of the function values.




\end{warning}



















\subsection*{Algebra}




Sine and cosine are related through the unit circle.

\[
(\sin(\theta))^2 + (\cos(\theta))^2 = 1
\]



If we divide both sides by $(\cos(\theta))^2$, we will get a similar algebraic relationship for $\tan(\theta)$ and $\sec(\theta)$.




\[
\frac{(\sin(\theta))^2 + (\cos(\theta))^2}{(\cos(\theta))^2} = \frac{1}{(\cos(\theta))^2}
\]





\[
\frac{(\sin(\theta))^2}{(\cos(\theta))^2} + \frac{(\cos(\theta))^2}{(\cos(\theta))^2} = \frac{1}{(\cos(\theta))^2}
\]





\[
(\tan(\theta))^2 + 1 = (\sec(\theta))^2
\]


And, 


\[
(\tan(\theta))^2 = (\sec(\theta))^2 - 1
\]




\textbf{\textcolor{red!70!black}{Looking Ahead...}}



Algebra is our thinking tool.  It provides ways to change the way we are looking at things. With a new perspective, we might see relationships and get new ideas.



This means we are constantly looking for ``forms'' - expressions that look like other relationships, that might help us change our perspective.


One perspective we would like to change is addition or subtraction inside a square root.

Since square roots do not distribute over addition or subtraction, we have littel useful algebra.  However, trigonometric relationships help us change our perspective.  They help us see addition and subtraction of squares as just squares.




\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ Forms}}


\[
(\sin(\theta))^2 = 1 - (\cos(\theta))^2
\]




\[
(\sec(\theta))^2 = (\tan(\theta))^2 + 1 
\]




\[
(\tan(\theta))^2 = (\sec(\theta))^2 - 1
\]





These relationships help us view expressions of the form $A^2 + B^2$ as just $C^2$.






















\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Unit Circle Functions]{https://ximera.osu.edu/csccmathematics/precalculus/precalculus/theUnitCircle/examples/exampleList}

\end{center}



\end{document}

