\documentclass{ximera}

\input{../../preamble.tex}

\title{Expected Rate}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}
\maketitle





This course is an investigation into the Elementary functions.  One of our goals is simply to build a library of the Elementary functions, their properties, and how we represent them.  From there, we can build more complex functions from these building blocks and compare them to each other. Our first two library functions are linear and quadratic. \\



\subsection*{Linear}

Our first exploration is into linear functions. The distinguishing characteristic about linear functions is that they have a constant growth rate or constant rate of change.  \\




Linear functions relate two measurements, which change proportionally.  That is when one measurement changes, the other changes by a constant multiple of the first change.  This constant multiple acts as a conversion factor.  We call it a \textbf{\textcolor{purple!85!blue}{rate}}. \\



This rate is constant throughout a linear function.





In other words, if you calculate the rate of change from any pair in a linear function to any other pair, you always get the same result.

Let $f$ be a linear function.  Let $(a, f(a))$ and $(b, f(b))$ any two pairs in the function.

The rate of change from $a$ to $b$ is 

\[
\frac{f(b) - f(a)}{b-a}
\]


For a linear function, this calculation gives the same result, no matter which two distinct domain numbers you choose.













From this, we discover that linear functions are those functions that can be described as

\[  L(d) = A \cdot d + B \]

where $A$ and $B$ are real numbers.


The graphs of linear functions are lines and we would like to be able to work back and forth between the algebra and geometry.













\subsection*{Quadratic}


Secondly, we will investigate quadratic functions. \\


Quadratic functions describe movement under the influence of gravity, commonly refered to as projectile motion. 





From this we can show that quadratic functions are those functions that can be described as

\[  Q(t) = A \cdot t^2 + B \cdot t + C \]

where $A$, $B$, and $C$ are real numbers. \\



The graphs of quadratice functions are parabolas and we would like to be able to work back and forth between the algebra and geometry. \\



Unlike linear functions, the rate of change of a quadratic is not constant. it changes depending on where in the domain you are measuring.  However, linear and quadratic functions are connected through rates. \\

As we will see, the rate of change of a quadratic function, changes proportionally (linearly). \\


THis also has a reverse perspective.  The accumulation of a linear change results in a quadratic amount.



This forwards and backwards relationship is the beginning of Calculus.





























\subsection*{Learning Outcomes}



\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item examine constant growth rates.
\item produce graphs of linear functions.
\item produce formulas from graphs.
\item explore quadratic functions.
\item draw parabolas
\item create tangent lines.
\end{itemize}
\end{sectionOutcomes}
















\pdfOnly{
\begin{center}
more examples can be found in the online edition.
\end{center}
}




\begin{onlineOnly}
\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Expected Rate]{https://ximera.osu.edu/csccmathematics/precalculus/precalculus/expectedRate/examples/exampleList}

\end{center}
\end{onlineOnly}






\end{document}
