\documentclass{ximera}

\input{../../preamble.tex}

\title{Analyzing Functions}

\begin{document}

\begin{abstract}
%
\end{abstract}
\maketitle






What do we want when analyzing a function?

\begin{itemize}
     \item \textbf{\textcolor{red!70!black}{Domain}} 
     \item \textbf{\textcolor{red!70!black}{Zeros}} 
     \item \textbf{\textcolor{red!70!black}{Continuity}} \\
     \text{    }$\circ$ \textbf{\textcolor{purple!85!blue}{discontinuities}}\\
     \text{    }$\circ$ \textbf{\textcolor{purple!85!blue}{singularities}}\\
     \item \textbf{\textcolor{red!70!black}{End-Behavior}} 
     \item \textbf{\textcolor{red!70!black}{Behavior}} \\
     \text{    }$\circ$ \textbf{\textcolor{purple!85!blue}{intervals where increasing}}\\
     \text{    }$\circ$ \textbf{\textcolor{purple!85!blue}{intervals where decreasing}}\\
     \item \textbf{\textcolor{red!70!black}{Global Maximum and Minimum}} 
     \item \textbf{\textcolor{red!70!black}{Local Maximums and Minimums}} 
     \item \textbf{\textcolor{red!70!black}{Range}} 
     \item \textbf{\textcolor{blue!55!black}{...and we would like a nice graph}} 
\end{itemize}







We want a nice, informative graph, meaning...
\begin{itemize}
	\item labeled axes
	\item intercepts
	\item endpoints
	\item vertical asymptotes
	\item horizontal asymptotes
	\item arrows
\end{itemize}


A graph is simply a communication tool.  It is inherently inaccurate by the very fact that it is drawn.  Algebra and functional reasoning are the only tools for exactness, which is our end goal. \\



$\blacktriangleright$ \textbf{\textcolor{red!70!black}{Reasoning:}} Reasoning is a logical explanation that describes our conclusions, how we arrived at those conclusions, and why we think those conclusions are correct. \\

Analysis is not a list of conclusions. We are not looking for such a list. \\

We are certainly not looking for such a list based on viewing a graph. \\

We are looking for the thought process that arrived at the list of conclusions. \\





\subsection*{Learning Outcomes}


\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item analyze functions algebraically from their formula.
\end{itemize}
\end{sectionOutcomes}


















\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Analyzing Functions]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/analyzingFunctions/examples/exampleList}

\end{center}







\end{document}


