\documentclass{ximera}

\input{../../preamble.tex}

\title{Function Behavior}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}
\maketitle




We have been examining function characteristics near individual numbers. We have noticed that, for the most part, all of the ``action'' happens within some interval.  Action includes zeros, discontinuities, singularities, maximums, and minimums.  $[-10, 10]$  is a very popular interval to investigate, because of technology.  Of course, functions can have zeros outside this interval.  

Perhaps we should be investigating the interval $[-1000, 1000]$, $[-50000, 50000]$, $[-750000, 750000]$, or $[-1000000, 1000000]$. Each of our elementary functions has a finite interval which holds all of its interesting places.  Even periodic functions have an interval containing a full single wave, which then describes the whole function.

Eventually, outside some interval, the function doesn't do anything surprising.  It settles down into a simple pattern.  We call this eventual pattern the \textbf{\textcolor{purple!85!blue}{end-behavior}}.



What are the simple patterns on $(-\infty, -a) \cup (a, \infty)$?  When is $a$ ``big enough''?








\subsection*{Learning Outcomes}


\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item examine endbehavior.
\item use limit notation.
\item examine horizontal asymptotes.
\end{itemize}
\end{sectionOutcomes}










\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Function Behavior]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/functionBehavior/examples/exampleList}

\end{center}






\end{document}
