\documentclass{ximera}

\input{../../preamble.tex}

\title{Set Symbols}

\begin{document}

\begin{abstract}
vocabulary
\end{abstract}
\maketitle




Mathematics is a language.  It uses symbols and vocabulary and language and pronunciation. \\




\subsubsection*{Sets}



We will be investigating functions that are defined on sets. Naturally, we have symbols for sets. \\



\textbf{\textcolor{blue!55!black}{Membership:}}  The symbol $\in$ means ``is a member of''.


\[
7 \in \{ 4, 5, 7, 8 \}
\]

\begin{center}

$7$ is a member of the set $\{ 4, 5, 7, 8 \}$.

\end{center}

People will also say ``$7$ is an element of'' $\{ 4, 5, 7, 8 \}$. \\

People will also say ``$7$ is in'' $\{ 4, 5, 7, 8 \}$. \\



\textbf{\textcolor{blue!55!black}{Subset:}}  The symbol $\subset$ means ``is a proper subset of''.


\[
\{ 4, 5 \} \subset \{ 4, 5, 7, 8 \}
\]

\begin{center}

$\{ 4, 5 \}$ is a subset of the set $\{ 4, 5, 7, 8 \}$.

\end{center}


Every member of $\{ 4, 5 \}$ is also a member of $\{ 4, 5, 7, 8 \}$. \\


\textbf{Proper} means that the subset is not equal to the larger set. \\


The symbol $\subseteq$ means ``is a subset of''.  This allows the possibility of the subset being equal to the larger set.

\[
\{ 4, 5, 7, 8 \} \subseteq \{ 4, 5, 7, 8 \}
\]









\textbf{\textcolor{blue!55!black}{Union:}}  The symbol $\cup$ stands for ``union''.


The union of two sets is another set.  The union contains all of the members of the two original sets.

\[
\{ 1, 2, 3 \} \cup \{ 4, 5, 7, 8 \} = \{ 1, 2, 3, 4, 5, 7, 8 \}
\]








\textbf{\textcolor{blue!55!black}{Intersection:}}  The symbol $\cap$ stands for ``intersection''.

The intersection of two sets is another set.  The intersection contains all of the members shared by the two original sets.

\[
\{ 1, 2, 3, 4, 5 \} \cap \{ 4, 5, 7, 8 \} = \{ 4, 5 \}
\]








\textbf{\textcolor{blue!55!black}{Empty Set:}}  The symbol $\emptyset$ stands for ``the empty set''.

The empty set is a set.  It just contains no members.














\subsection*{Numbers}

We have some standard sets of numbers and they have special symbols. \\



\begin{itemize}

\item $\mathbb{N}$ : the natural numbers
\item $\mathbb{Z}$ : the integers
\item $\mathbb{Q}$ : the rational numbers
\item $\mathbb{R}$ : the real numbers
\item $\mathbb{C}$ : the complex numbers
\end{itemize}

















\subsection*{Small}



As we move toward Calculus, our attention will focus on ``close''... a lot! \\


The word \textbf{instantaneous} will describe our ``close'' measurements. \\


So, we will use our symbols, notation, and language to help us talk about ``close''. \\


We have two symbols from the Greek language that we traditionally use to mean ``a small positive number''. \\



\begin{notation} \textbf{\textcolor{red!80!black}{Small}} 


\textbf{\textcolor{blue!55!black}{epsilon:}}  $\epsilon$ \\

\textbf{\textcolor{blue!55!black}{delta:}}  $\delta$ \\


\end{notation}

$\epsilon$ and $\delta$ are used to mean ``a very small positive number''. \\  


$\epsilon$ and $\delta$ are how we talk algebraically about ``close''. \\





\begin{warning}


$\epsilon$ and $\delta$ are not specific numbers with specific numeric values, like $\pi$. \\


$\epsilon$ and $\delta$ are used to represent very small positive numbers, but not specific small positive numbers. \\


So, $\epsilon$ and $\delta$ are used as constants - small positive constants.


When we use $\epsilon$ and $\delta$, we are talking about numbers smaller than $0.000000000001$. 


\end{warning}




\begin{idea}

\textbf{\textcolor{blue!55!black}{$\blacktriangleright$}}  ``small'' means ``close'' to $0$. \\


A number can be small and negative or a number can be small and positive. Small just means really close to $0$. \\


\textbf{\textcolor{blue!55!black}{$\blacktriangleright$}}  ``big'' or ``large'' means ``far way from'' to $0$. \\


A number can be big and negative or a number can be big and positive. Big just means far awy from $0$. \\


\end{idea}


``Big'' and ``small'' are size words.  They describe how close a number is to $0$. \\

Separate from this are ``greater than'' and ``less than''.  These are position words. \\

\begin{idea}


\textbf{\textcolor{blue!55!black}{$\blacktriangleright$}} ``lesser'' or ``less than'' mean to the left of on the number line. These words describe realtive position.\\


\textbf{\textcolor{blue!55!black}{$\blacktriangleright$}} ``Greater'' or ``less than'' mean to the right of on the number line. These words describe realtive position.\\



\begin{example}


The number $-7$ is less than the number $3$, but $3$ is smaller than $-7$.



\end{example}



\begin{example}


The number $-9$ is less than the number $2$, but $-9$ is bigger than $2$.



\end{example}


\end{idea}

















\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Real-Valued Functions]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/realValued/examples/exampleList}

\end{center}



\end{document}
