\documentclass{ximera}

\input{../../preamble.tex}

\title{Piecewise Composition}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}
\maketitle



\section*{Objects}


\begin{quote}
Functions are packages containing three sets: a domain, a range, and a set of pairs.
\end{quote}

In this view, functions are objects.  They consist of pieces. They have features, characteristics, and attributes.  As objects, they can be examined, measured, and involved in operations.


We can also link these objects as a chain by gluing together the range of one function with the domain of another function.  In this way, we produce a new function called the \textbf{composition}.












\section*{Machines}



We can also view functions as machines, so-called black-boxes.  They take input from the domain, process it, and return an output from the range.  For each input, the process produces exactly one output.



This output could easily be fed into another function, processed by this second function, and producing a final output. 



Before chaining whole functions together to form new functions, in this section, we will simply follow individual numbers through the chaining process and see where they end up.















Functions, like numbers, are a system of things (functions) and rules (operations).  

\begin{itemize}
\item the sum of two functions is another function.
\item the difference of two functions is another function.
\item the product of two functions is another function.
\item the quotient of two functions is another function.
\end{itemize}

Of course, we always have to pay some attention to domains.



The usual attention to domains includes adopting the intersection of domains to be the new domain, and avoiding domain numbers which make a quotient's denominator equal to $0$, argument of a square root negative, or argument of a logarithm $0$ or negative.



In addition to these four familiar operations, functions have another operation: \textbf{\textcolor{purple!85!blue}{composition}}.

We have already peeked at this. We took a single output number of one function and used it as the input to anther function.  We can extend this idea to the whole domain of a function.


The \textbf{\textcolor{purple!85!blue}{composition of F and G}}, symbolized by $F \circ G$ or $F(G)$, is a new function.  Its domain comes from the domain of $G$. Its range comes from the range of $F$.



In this section, we will restrict our study to the composition of linear functions.















\subsection*{Learning Outcomes}


\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item trace values of one function as domain value for another.
\item identify implied domains.
\item create formulas for compositions.
\item create graphs for compositions.
\end{itemize}
\end{sectionOutcomes}











\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Pointwise Composition]{https://ximera.osu.edu/csccmathematics/precalculus/precalculus/compositionPiecewise/examples/exampleList}

\end{center}





\end{document}
