\documentclass{ximera}

\input{../../preamble.tex}

\title{Bridge}

\begin{document}

\begin{abstract}
exp, trig, hyp
\end{abstract}
\maketitle





What can we do with Euler's Formula?



\[ e^{i t} = \cos(t) + i \, \sin(t) \]

Substituting $-t$ for $t$ gives $e^{i (-t)} = \cos(-t) + i \, \sin(-t) = \cos(t) - i \, \sin(t)$, because $\cos(t)$ is an even function and $\sin(t)$ is an odd function.



That gives us 

\[   e^{i t}  + e^{-i t} = 2 \cos(t)    \]


\[   \frac{e^{i t}  + e^{-i t}}{2} = \cos(t)    \]


It also gives us 



\[   e^{i t}  - e^{-i t} = 2 i \sin(t)    \]


\[   \frac{e^{i t}  - e^{-i t}}{2 i} = \sin(t)    \]


This tells us that the geometry of right triangles can be expressed through exponential functions.

Our right triangles come from the unit circle, which are the solutions to the equation $x^2 + y^2 = 1$.


There is also a unit hyperbola.  It consists of the points that satisfy the equation $x^2 - y^2 = 1$.

We can define two functions from the two coordinates of points on the unit hyperbola.  We call these hyperbolic sine ($\sinh$) and hyperbolic cosine ($\cosh$).


They can also be expressed using exponentials.


\[  \cosh(t)  = \frac{e^t + e^{-t}}{2}      \,   \text{ and } \,  \sinh(t)  = \frac{e^t - e^{-t}}{2}     \]



The geometry of the circle and the geometry of the hyperbola are connected through complex numbers.

\[  \cos(t) =  \frac{e^{i t} + e^{-i t}}{2} = \cosh(i \, t)      \]




\[  \sin(t) =  \frac{e^{i t} - e^{-i t}}{2 i} = -i \sinh(i \, t)      \]






If we work with Complex numbers, then trigonometric and hyperbolic functions are expressible in terms of exponential functions.  They are different ways of expressing the same structure.






















\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of the Complex Bridge]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/complexBridge/examples/exampleList}

\end{center}









\end{document}
