\documentclass{ximera}

\input{../../preamble.tex}

\title{Euler's Formula}

\begin{document}

\begin{abstract}
complex exponentials
\end{abstract}
\maketitle






\subsection*{An Adventure}




One of the biggest surprises in mathematics is the fact that trigonometric functions and exponential functions are the same building blocks, provided we use Complex numbers. 


Euler's Formula is the bridge:



\[    e^{i \, t} = \cos(t) + i \sin(t)    \]



This is very weird.  It claims that exponential functions are trigonometric functions are two perspectives on the same thing.  Our previous mathematcis courses have never even hinted at such a connection.  That is because the bridge  travels through the Complex Numbers.



With this bridge, we now have direct connections between


\begin{itemize}
\item percentage growth
\item exponential functions
\item trigonometric functions
\item triangles
\item circles
\end{itemize}

This would also connect inverse trigonometric functions and logarithms and soon hyperbolas and hyperbolic trigonometric functions.  This is crazy!


\begin{center}
\textbf{\textcolor{red!70!black}{Everything is Connected!!!}}
\end{center}






It would be nice to show that Euler's Formula is true.  However, we will need Calculus to dot our i's and cross our t's and leave no doubt.   


But, we can understand the basic idea. 


We just have to remember properties of $\cos(t) + i \sin(t)$ that we have seen.




\textbf{$\blacktriangleright$ Multiplication}


$\cos(t) + i \sin(t)$ is a Complex number on the unit circle and we have seen that multiplication is accomplished by adding the angles.



\[ (\cos(\theta_1) + i \sin(\theta_1)) \cdot (\cos(\theta_2) + i \sin(\theta_2))  = \cos(\theta_1 + \theta_2) + i \sin(\theta_1 + \theta_2) \]


Just like $A^B \cdot A^C = A^{B+C}$




\textbf{$\blacktriangleright$ Quotients}


$\cos(t) + i \sin(t)$ is a Complex number on the unit circle and we have seen that division is accomplished by subtracting the angles.


\[  \frac{\cos(\theta_1) + i \sin(\theta_1)}{\cos(\theta_2) + i \sin(\theta_2)} = \cos(\theta_1 - \theta_2) + i \sin(\theta_1 - \theta_2) \]


Just like $\frac{A^B}{A^C} = A^{B-C}$





\textbf{$\blacktriangleright 1$ }

\[  \cos(0) + i \sin(0)  = 1\]


Just like $r^0 = 1$


The arithmetic of $\cos(t) + i \sin(t)$  is the same as the exponential arithmetic.

And, since every exponential function can be written with base $e$, we have an odd conclusion.


\begin{conclusion} \textbf{\textcolor{green!25!black}{Exponential}}

$\cos(t) + i \sin(t)$ can somehow be represented as $e^{A + B \, i}$.


\end{conclusion}



What else can we deduce?


\textbf{$\blacktriangleright$ Unit Circle}


We know that every product of numbers of the form $\cos(t) + i \sin(t)$ is located on the unit circle, meaning its modulus is $1$.


If $\cos(t) + i \sin(t) = e^{A + B \, i}$, then the modulus of $e^{A + B \, i}$ equals $1$, even for $B = 0$.

The modulus of $e^{A + 0 \, i} = e^A$ is $1$.  But, that can only happen if $A=0$.



\begin{conclusion} \textbf{\textcolor{green!25!black}{Exponential}} 

$\cos(t) + i \sin(t)$ can somehow be represented as $e^{B \, i}$.


\end{conclusion}


We have it whittled down quite a bit.  One side is in terms of $t$ and the other side is in terms of $B$.

It is not that much of a stretch of the imagination to hypothesize that $B = 1$.








\begin{theorem} \textbf{\textcolor{-red!75!green}{Euler's Formula}} 



\[ \cos(t) + i \sin(t) = e^{t \, i} \]


\end{theorem}



\subsection*{$\mathbb{C}$}



Every Complex number in $\mathbb{C}$ can be witten as a product of a number on the unit circle and a positive real number.

The number on the unit circle gives the direction and the real number gives the distance from the origin.


\begin{itemize}
\item Every Complex number on the unit circle (direction) can be given in the form $\cos(t) + i \sin(t)$.
\item Every positive real number can be given in the form $e^r$.
\end{itemize}




\[ R \, (\cos(\theta) + i \sin(\theta)) = e^r  \, (\cos(\theta) + i \sin(\theta)) = e^r \cdot e^{i \, t} = e^{r + i \, t}\]



\begin{theorem} \textbf{\textcolor{-red!75!green}{$\mathbb{C}$}} 


Every nonzero Complex number can be written in the exponential form 

\[ e^{r + \theta \, i} \]


\end{theorem}









































\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Complex Bridge]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/complexBridge/examples/exampleList}

\end{center}







\end{document}
