\documentclass{ximera}

\input{../../../preamble.tex}


\author{Lee Wayand}

\begin{document}
\begin{exercise}



Completely analyze $y(x) = \left(\frac{4}{3}\right)^x - 3$ with its natural domain.



\textbf{\textcolor{blue!55!black}{Domain:}}  \\

$y$ is a shifted exponential function, therefore its natural domain is $(-\infty, \infty)$.




\textbf{\textcolor{blue!55!black}{Continuity:}}  \\

$y$ is a shifted exponential function, therefore it is continuous on its domain and has no discontinuities.


$y$ is a shifted exponential function, therefore it has no singularities.




\textbf{\textcolor{blue!55!black}{End-Behavior:}}  \\



$y$ is a shifted exponential function. The leading coefficient is $1$, which is positive.  The base is $\frac{4}{3}$, which is greater than $1$.  


$\blacktriangleright$ As $x$ becomes big and negative, then the exponent becomes big and negative.  We have a number greater than $1$ raised to big negative numbers.  That will approach $0$ and then we subtract $3$

\[
\lim\limits_{x \to -\infty} y(x) = -3
\]


$\blacktriangleright$ As $x$ becomes big and positive, then the exponent becomes big and positive.  We have a number greater than $1$ raised to big positive numbers. That becomes unbounded positively.

\[
\lim\limits_{x \to \infty} y(x) = \infty
\]




\textbf{\textcolor{blue!55!black}{Behavior:}}  \\

$y$ is a shifted exponential function. The leading coefficient is $1$, which is positive.  The base is $\frac{4}{3}$, which is greater than $1$.  Therefore, $y$ is an increasing function.

Shifted exponential functions have no global or local maximums or minimums.




\textbf{\textcolor{blue!55!black}{Zeros:}}  \\


Since we have a continuous exponential function with negative and positive values, it will have a single zero.


The zero will occur when 

\[
\left(\frac{4}{3}\right)^x = 3
\]

$x$ is the thing you raise $\frac{4}{3}$ to, to get $3$.

\[
x = log_{\tfrac{4}{3}}(3)
\]




This all agrees with the graph.


\begin{image}
\begin{tikzpicture} 
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel=$y$, 
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=1, gray, dashed,domain=(-9:9),<->] ({x},{-3});
          \addplot [line width=2, penColor, smooth,samples=200,domain=(-9:8.5),<->] {(1.333)^x - 3};
          

           

  \end{axis}
\end{tikzpicture}
\end{image}















\end{exercise}
\end{document}