\documentclass{ximera}

\input{../../../preamble.tex}


\author{Lee Wayand}

\begin{document}
\begin{exercise}



Completely analyze $y(k) = -2 \ln(-k+5)$ with its natural domain.



\textbf{\textcolor{blue!55!black}{Domain:}}  \\

$y$ is a logarithmic function, therefore its natural domain contains those numbers that make the inside positive. 

\[
-k + 5 > 0
\]

The natural domain is $(-\infty, 5)$.




\textbf{\textcolor{blue!55!black}{Continuity:}}  \\

$y$ is a logarithmic function, therefore it is continuous on its domain and has no discontinuities.


$y$ is a logarithmic function, therefore it has a singularity, when the inside is $0$, which is when $k = 5$.




\textbf{\textcolor{blue!55!black}{Behavior Near a Singularity:}}  \\



$\blacktriangleright$ $y$ is a logarithmic function. The leading coefficient is $-2$, which is negative.  When the domain gets near the singularity $ln(-k+5)$ becomes unbounded negatively.  Since the leading coefficient is negative, we have

\[
\lim\limits_{k \to 5^-} y(k) = \infty
\]





\textbf{\textcolor{blue!55!black}{End-Behavior:}}  \\


$\blacktriangleright$ As $x$ becomes big and negative, then the inside becomes big and positive.  $ln(big positive)$  is big and positive.  Since the leading coefficient is negative, we have

\[
\lim\limits_{k \to \infty} y(x) = -\infty
\]








\textbf{\textcolor{blue!55!black}{Behavior:}}  \\

$y$ is a logarithm function, therefore, is either increasing or decreasing.  The leading coefficient is $-2$, which is negative.  


Logarithmic functions have no global or local maximums or minimums. Since the extreme behavior is $\infty$ to $-\infty$, we know that $y$ is a decreasing function.




\textbf{\textcolor{blue!55!black}{Zeros:}}  \\


Since we have a continuous logarithmic function with negative and positive values, it will have a single zero.


The zero will occur when 

\[
-2 \ln(-k+5) = 0
\]

\[
\ln(-k+5) = 0
\]



\[
1 = -k + 5
\]


\[
k = 4
\]



This all agrees with the graph.


\begin{image}
\begin{tikzpicture} 
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$k$, ylabel=$y$, 
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=1, gray, dashed,domain=(-9:9),<->] ({5},{x});
          \addplot [line width=2, penColor, smooth,samples=200,domain=(-9:4.99),<->] {-2 * ln(-x + 5)};
          

           

  \end{axis}
\end{tikzpicture}
\end{image}















\end{exercise}
\end{document}