\documentclass{ximera}

\input{../../preamble.tex}



\title{Cooling}

\begin{document}
\begin{abstract}
  Newton's Law
\end{abstract}
\maketitle




Newton's Law of Cooling states



\begin{quote}


The rate of heat loss of a body is directly proportional to the difference in the temperatures between the body and its environment. 
\end{quote}



Newton formulated this relation as 

\[
\dot{Q} = h \, A \, (T(t) - T_{env})
\]


where


\begin{itemize}
\item $\dot{Q}$ is the rate of heat transfer out of the body.
\item $h$ is the  heat transfer coefficient.
\item $A$ is the heat transfer surface area.
\item $T$ is the temperature of the object's surface.
\item $T_{env}$ is the temperature of the environment.
\end{itemize}



In modern notation, we have

\[
\frac{dT}{dt} = r \, (T_{env} - T(t))
\]

where $r$ is the coefficient of heat transfer.\\





The derivative (rate of change) of temperature is proportional to the difference betweent the current temperature and the surrounding temperature.


\begin{observation}


Our story is a cooling story, therefore the temperature of the body is higher than the surrounding temperature.  That makes $T_{env} - T(t) < 0$, which means our derivatiev is negative, which means the temperature is decreasing, i.e. cooling.
\end{observation}



\begin{observation}


As the body cools, $T$ becomes closer to $T_{env}$, which means  $T_{env} - T(t)$ gets closer to $0$.  Therefore, $\frac{dT}{dt}$ gets closer to $0$.  The body cools slower as it cools.
\end{observation}



\begin{observation}


The body cools, however, it can't cool below the surrounding temperature. Therefore, we have

\[
\lim\limits_{t \to \infty} T(t) = T_{env}
\]

\end{observation}








\begin{example} Law of Cooling 



A freshly brewed pot of tea is removed from the stove at $200^{\circ}$.  It is set on the table where the room temperature is $72^{\circ}$ and the heat transfer coefficient is $0.025$. \\


Newton's Law of Cooling says

\[
\frac{dT}{dt} = 0.025 \, (72 - T(t))
\]


where $t$ is measured in minutes.


$\blacktriangleright$ \textbf{Frrst Minute} \\

The tea has an initial temperature of $200^{\circ}$, which means the temperature is changing approximately at a rate of 


\[
\frac{dT}{dt} = 0.025 \, (72 - 200) = 3.2 \, \text{degrees per minute}
\]


After $1$ minute the tea is at approximately $196.8^{\circ}$.




\end{example}




















\end{document}
