\documentclass{ximera}

\input{../../preamble.tex}



\title{Rates}

\begin{document}
\begin{abstract}
  rate dependent on function
\end{abstract}
\maketitle

A \textit{differential equation}\index{differential equation} is
simply an equation with a derivative in it. Here is an example:
\[
a \cdot f^{\prime\prime}(x) + b \cdot f^{\prime}(x) + c \cdot f(x) = g(x). 
\]
\begin{question}
  What is a differential equation?
  \begin{multipleChoice}
    \choice{An equation that you take the derivative of.}
    \choice[correct]{An equation that relates the rate of a function to the function value.}
    \choice{It is a formula for the slope of a tangent line at a given point.}  
  \end{multipleChoice}
\end{question}








When a mathematician solves a differential equation, they are finding
\textit{functions} satisfying the equation.
\begin{example}
  Which of the following functions solve the differential equation
  \[
  f^{(4)}(x) = f(x)?
  \]

    Remember: $f^{(4)}$ is the fourth derivative of $f$.

    It turns out that the complete solution to this differential equation
    is $c_1\sin(x)+c_2\cos(x)+c_3e^x+c_4e^{-x}$.  In other words, every
    solution to this differential equation can be written in this form
    (for example $f(x) =
    \sin(x)+3\cos(x)-7e^x+\pi e^{-x}$ is a solution).  Proving that these
    are \textbf{all} of the solutions is beyond the scope of this course.

\end{example}


The differential equation above is an example of a fourth \textbf{order}
differential equation, because the highest derivative in the equation
is a ``fourth'' derivative. In general the highest derivative in a
differential equation is the order.


Differential equations are one of the most practical objects of
mathematical study.  They appear constantly in every field of science
and engineering.  They are a powerful way to model many diverse
situations.





\section{Modeling with differential equations}

Setting up differential equations is a skill to be acquired. However,
you can try your hand with our next question.

\begin{question}
  Imagine that a glass of water has initial temperature $5^\circ
  \mathrm{C}$, and that the ambient temperature is $22^\circ \mathrm{C}$.
  The water will warm up over time.  Assume that the rate of change in
  the temperature of the water is directly proportional to the
  difference between the current water temperature and the ambient
  temperature.  Which of the following differential equations 
  describes this relationship for the function $H(t)$ which measures the temperature of
  the water with respect to time?
  \begin{multipleChoice}
    \choice{$H^{\prime} = 5+\frac{H}{22}$}
    \choice[correct]{$H^{\prime} = k(22-H)$ for some $k>0$}
    \choice{$H^{\prime} = k(H-22)$ for some $k>0$}
    \choice{$H^{\prime} = k(5-H)$ for some $k>0$}
    \choice{$H^{\prime} = k(H-5)$ for some $k>0$}
  \end{multipleChoice}
  \begin{hint}
    This is just a straight translation job.  ``The rate of change in
    the temperature of the water'' is $H^{\prime}$.  ``Directly proportional
    to'' means that it is equal to some constant (say $k$) times
    whatever it is proportional to.  ``The difference between the
    current water temperature and the ambient temperature" is either
    $22-H$ or $H-22$, since $H$ is the temperature of the water and
    $22$ is the ambient temperature.  Think about which we should
    choose before looking at the next hint.  Will it be $H^{\prime}=k(22-H)$
    or $H^{\prime}=k(H-22)$ where $k>0$?
  \end{hint}
  \begin{hint}
    Since the temperature of the water is increasing over time, we
    want $H^{\prime}>0$.  Since the temperature will be increasing (and it is
    reasonable to assume it never surpasses the ambient temperature!)
    $22-H$ is positive.  So we can conclude that $H^{\prime} = k(22-H)$ for
    some $k>0$.
  \end{hint}
  
  \begin{feedback}
    The differential equation does not involve the number $5$.  If we
    wanted to incorporate that piece of data into our model we could
    ask ``Which solution(s) to this differential equation satisfy
    $H(0) = 5$?''  This is known as an \textbf{initial value problem}.  %%%
  \end{feedback}
\end{question}
We say that $H$ \textbf{satisfies}  $y^{\prime} = k(22-y)$ for some $k>0$. \\

Sometimes the rates in question are constant. 



\begin{question}
  One can approximate the force of gravity as constant near the Earth.
  So the acceleration of a falling object is a constant $g>0$.  If
  $h(t)$ is the height of an object at time $t$, which differential
  equation describes $h$?
  \begin{multipleChoice}
    \choice{$h^{\prime} = g$}
    \choice{$h=-g$}
    \choice[correct]{$h^{\prime\prime}=-g$}
    \choice{$h^{\prime\prime} = g$}
    \choice{$h^{\prime\prime} = -gy$}
  \end{multipleChoice}
  
  \begin{hint}
    The acceleration of an object is the second derivative of its
    position, so the differential equation should say the second
    derivative, $h^{\prime\prime}$, is constant.  Should it be a positive of
    negative constant?
  \end{hint}
  \begin{hint}
    A falling object will fall quicker and quicker, so the second
    derivative of its height should be negative.  Thus $h^{\prime\prime}=-g$ is the
    correct answer.
  \end{hint}	
\end{question}
We say that $h$ \textbf{satisfies}  $y^{\prime\prime}=-g$. \\


















\end{document}
