\documentclass{ximera}

\input{../../preamble.tex}

\title{Differential Equations}

\begin{document}

\begin{abstract}
%%%
\end{abstract}
\maketitle




We have seen functions described in two algebraic ways: formulas and equations.


Formulas are special equations where the function (dependent variable) is isolated by itself on one side of the equation.  The formula provides a procedure whereby you begin with a domain number, perform calculations with it, and produce the function value.

We have already seen instances where a formula for a function leads to a formula for its derivative. 

$\blacktriangleright$ If $f(x) = a \, x^2  + b \, x + c$, then $f'(x) = 2a \, x + b$

$\blacktriangleright$ If $g(t) = A \, \sin(t) + B$, , then $g'(t) = A \, \cos(t)$


When the function is described explicitly, then so is the derivative.


We have also seen functions defined implicitly.


$f(x)$ might be described by $x \cos(f) - \sqrt(x + f) = 0$.  There is no way to solve for $f$ in terms of $x$. It cannot be separated.  Same for its derivative. There is no way to separate $f'(x)$ from $f(x)$.

The equation describing $f'(x)$ includes $f(x)$ and there is no way around it.


We call these equations involving the function and its derivative \textbf{Differential Equations}.

It turns out that Differential Equations are the way scientists and engineers model the world.










\subsection{Learning Outcomes}


\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item examine differential equations and approximate function values.
\end{itemize}
\end{sectionOutcomes}








\begin{center}
\textbf{\textcolor{green!50!black}{ooooo=-=-=-=-=-=-=-=-=-=-=-=-=ooOoo=-=-=-=-=-=-=-=-=-=-=-=-=ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Differential Equations]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/differentialEquations/examples/exampleList}

\end{center}






\end{document}
