\documentclass{ximera}

\input{../../preamble.tex}

\title{Up and Down}

\begin{document}

\begin{abstract}
stretch the range
\end{abstract}
\maketitle











Let $D(y)$ be a function with its domain.

Let $K(r)$ be defined as $K(r) = 2 D(r)$ with its induced domain.


Then the domain of $K$ is

\begin{multipleChoice}
\choice {the domain of $D$ shifted left $7$}
\choice {the domain of $D$ shifted right $7$}
\choice[correct] {the same as for $D$}
\end{multipleChoice}













\begin{example} Stretching Vertically




Graph of $y = m(f)$.

\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$f$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
	\addplot [draw=penColor,very thick,smooth,domain=(-7:-4)] {-x-6};
	\addplot [draw=penColor,very thick,smooth,domain=(-2:1)] {x-3};
	\addplot [draw=penColor,very thick,smooth,domain=(1:7)] {-x+5};

	\addplot[color=penColor,only marks,mark=*] coordinates{(-3,2)}; 
	
	\addplot[color=penColor,only marks,mark=*] coordinates{(-7,1)}; 
	\addplot[color=penColor,fill=white,only marks,mark=*] coordinates{(-4,-2)}; 
	\addplot[color=penColor,only marks,mark=*] coordinates{(-2,-5)}; 
	\addplot[color=penColor,only marks,mark=*] coordinates{(1,-2)}; 
	\addplot[color=penColor,only marks,mark=*] coordinates{(1,4)}; 
	\addplot[color=penColor,fill=white,only marks,mark=*] coordinates{(7,-2)}; 


    \end{axis}
\end{tikzpicture}
\end{image}


$m$ two zeros: $-6$ and $5$. These are represented by intercepts in the left and right line segments. \\
$m$ has a global minimum represent by the left endpoint of the middle line segment. \\
$m$ has a global miaximum represent by the left endpoint of the right line segment.





Graph of $z = P(t) = 2 m(t)$.

\begin{image}
\begin{tikzpicture}
	\begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$f$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
	\addplot [draw=penColor,very thick,smooth,domain=(-7:-4)] {2*(-x-6)};
  \addplot [draw=penColor,very thick,smooth,domain=(-2:1)] {2*(x-3)};
  \addplot [draw=penColor,very thick,smooth,domain=(1:7)] {2*(-x+5)};

  \addplot[color=penColor,only marks,mark=*] coordinates{(-3,4)}; 
  
  \addplot[color=penColor,only marks,mark=*] coordinates{(-7,2)}; 
  \addplot[color=penColor,fill=white,only marks,mark=*] coordinates{(-4,-4)}; 
  \addplot[color=penColor,only marks,mark=*] coordinates{(-2,-10)}; 
  \addplot[color=penColor,only marks,mark=*] coordinates{(1,-4)}; 
  \addplot[color=penColor,only marks,mark=*] coordinates{(1,8)}; 
  \addplot[color=penColor,fill=white,only marks,mark=*] coordinates{(7,-4)}; 


    \end{axis}
\end{tikzpicture}
\end{image}




\begin{itemize}

\item The domain of $P$ is $[-7,-4) \cup \{3\} \cup [-2,7)$, equal to the domain of $m$. $2$ was outside the domain parentheses in the formula for $m$.
\item $P$ two zeros: $-6$ and $5$. These are represented by intercepts in the left and right line segments. $2 \cdot 0 = 0$. \\
\item $P$ has a global minimum represent by the left endpoint of the middle line segment. \\
\item $P$ has a global miaximum represent by the left endpoint of the right line segment.

\end{itemize}


\end{example}


Multiply a function by a positive constant stetches everything.  Since stretching $0$ still gives $0$, the intercepts do no change.  Everythig is stetched from the intercepts.


















Multiplying by a negative constant does the same thing, but also reflects the graph about the horizontal axis.





\begin{example} Reflecting Vertically




Graph of $y = m(f)$.

\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$f$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
  \addplot [draw=penColor,very thick,smooth,domain=(-7:-4)] {-x-6};
  \addplot [draw=penColor,very thick,smooth,domain=(-2:1)] {x-3};
  \addplot [draw=penColor,very thick,smooth,domain=(1:7)] {-x+5};

  \addplot[color=penColor,only marks,mark=*] coordinates{(-3,2)}; 
  
  \addplot[color=penColor,only marks,mark=*] coordinates{(-7,1)}; 
  \addplot[color=penColor,fill=white,only marks,mark=*] coordinates{(-4,-2)}; 
  \addplot[color=penColor,only marks,mark=*] coordinates{(-2,-5)}; 
  \addplot[color=penColor,only marks,mark=*] coordinates{(1,-2)}; 
  \addplot[color=penColor,only marks,mark=*] coordinates{(1,4)}; 
  \addplot[color=penColor,fill=white,only marks,mark=*] coordinates{(7,-2)}; 


    \end{axis}
\end{tikzpicture}
\end{image}


$m$ two zeros: $-6$ and $5$. These are represented by intercepts in the left and right line segments. 
$m$ has a global minimum represent by the left endpoint of the middle line segment. 
$m$ has a global maximum represent by the left endpoint of the right line segment.





Graph of $z = B(h) = -2 m(h)$.

\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$h$, ylabel=$z$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
  \addplot [draw=penColor,very thick,smooth,domain=(-7:-4)] {2*x+12};
  \addplot [draw=penColor,very thick,smooth,domain=(-2:1)] {-2*x+6};
  \addplot [draw=penColor,very thick,smooth,domain=(1:7)] {2*x-10};

  \addplot[color=penColor,only marks,mark=*] coordinates{(-3,-4)}; 
  
  \addplot[color=penColor,only marks,mark=*] coordinates{(-7,-2)}; 
  \addplot[color=penColor,fill=white,only marks,mark=*] coordinates{(-4,4)}; 
  \addplot[color=penColor,only marks,mark=*] coordinates{(-2,10)}; 
  \addplot[color=penColor,only marks,mark=*] coordinates{(1,4)}; 
  \addplot[color=penColor,only marks,mark=*] coordinates{(1,-8)}; 
  \addplot[color=penColor,fill=white,only marks,mark=*] coordinates{(7,4)}; 


    \end{axis}
\end{tikzpicture}
\end{image}


The shape has not chnaged.  It has just been flipped over vertically, about the horizontal axis.

\begin{itemize}

\item The domain of $B$ is $[-7,-4) \cup \{3\} \cup [-2,7)$, equal to the domain of $m$. $-2$ was outside the domain parentheses in the formula for $m$.
\item $B$ two zeros: $-6$ and $5$. These are represented by intercepts in the left and right line segments. $-2 \cdot 0 = 0$. \\
\item $B$ has a global minimum.  It is represented by the flipped highest dot in the graph of $m$.\\
\item $B$ has a global maximum.  It is represented by the flipped lowest dot in the graph of $m$.

\end{itemize}


\end{example}
















\begin{example} Sine



Graph of $y = 3 sin(\theta)$.

\begin{image}
\begin{tikzpicture} 
  \begin{axis}[
            domain=-10:10, ymax=3.5, xmax=10, ymin=-3.5, xmin=-10,
            xtick={-6.28, -3.14, 3.14, 6.28}, 
            xticklabels={$-2\pi$, $-\pi$, $\pi$, $2\pi$},
            axis lines =center,  xlabel={$\theta$}, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          	\addplot [line width=2, penColor, smooth,samples=200,domain=(-9:9), <->] {3*sin(deg(x))};

           

  \end{axis}
\end{tikzpicture}
\end{image}


The location of zeros, maximums, and minimums inside the domain have not changed.  The function values have been multiplied by $3$. Therefore, the maximum and minimum values of the funciton have changed. But their positions have not.

\begin{itemize}
\item The zeros of $sin(\theta)$ are all integer multiples of $\pi$.
\item The maximum value is $1$ and it occurs at:  $\left\{     \frac{(4k+1)\pi}{2} \, | \, k \in \textbf{Z}     \right\}$
\item The minimum value is $-1$ and it occurs at:  $\left\{    \frac{(4k+3)\pi}{2} \, | \, k \in \textbf{Z}     \right\}$
\end{itemize}













\end{example}






Neither shifting nor stretching (or compressing) changes the shape of the graph.  The extreme features may chnage their values, but they remain relatively in the same position.  The shape may be reflected about one (or both) of the axis, but the shape is the same, just draw in reverse from the original.





\section{Together}

We can apply horizontal and vertical stretches together as well.





\begin{example}  Shifting

Let $C(w) = \frac{1}{2}|3w|$.


\begin{image}
\begin{tikzpicture} 
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$w$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=2, penColor, smooth, samples=200, domain=(-5:5),<->] {0.5 * abs(3*x)};
        

  \end{axis}
\end{tikzpicture}
\end{image}

\end{example}

The graph of $|w|$ has been compressed horizontally by a facotr of $3$ and compressed vertically by a factor of $\frac{1}{2}$.




We have already noticed that mulitplication and addition on the inside of the formula's parentheses affect the domain.  They also have graphical effects which seem backwards from the arithmetic.  This is because the formula inside the parenthese does not represent the domain numbers of the original function.  We have to set the original variable equal to this new domain expression and solve for the new variable.  That tells us what happens to theoriginal variable.  When we solve, all of the arithmetic is reversed and that is what we see graphically.

This is not the case for multiplication and addition on the outside of the parentheses.  The value of the formula is the value of the function, which is the heights of the dots on the graph.  The multiplication and addition is applied directly to this value.  Therefore, the graphical effects follow the arithmetic.


\begin{center}

Graphs follow domain transformations backwards or in reverse as the arithmetic.


\end{center}


\begin{center}

Graphs follow range transformations exactly as the arithmetic.


\end{center}




\end{document}
