\documentclass{ximera}

\input{../../preamble.tex}

\title{Logarithmic}

\begin{document}

\begin{abstract}
inverse
\end{abstract}
\maketitle


\section*{Core Logarithmic Functions}




The expression $\log_a(b)$ was defined to be the number that you raise $a$ to, to get $b$.

\[   a^{\log_r(b)} = b  \]




We made a Core logarithmic function from this by fixing the base:  $L(x) = \log_r(x)$. 


This function is the ``reverse'' of $r^x$.  Meaning that if $(A, B)$ is a pair in the $L(x) = \log_r(x)$ function, then $(B, A)$ is a pair in the $r^x$ function. Their domains and ranges are swapped. 





\begin{itemize}

\item The domain of an exponential function is all real numbers. The range of a logarithmic function is all real numbers.  


\item The range of an exponential function is all positive real number. The domain of a logarithmic function is all positive real numbers.

\item The graphs of exponential functions have a horizontal asymptote. The graphs of logarithmic functions have a vertical asymptote. 

\end{itemize}






Here is the graph of $y = L(x) = \log_2(x)$.

\begin{image}
\begin{tikzpicture} 
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=2, penColor, smooth,samples=300,domain=(0.001:9),<->] {ln(x)/ln(2)};
          \addplot [line width=1, gray, dashed,domain=(-9:9),<->] ({0},{x});

           

  \end{axis}
\end{tikzpicture}
\end{image}






Here is the graph of $y = E(x) = 2^x$.

\begin{image}
\begin{tikzpicture} 
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=2, penColor, smooth,samples=200,domain=(-9:3.2),<->] {2^x};
          \addplot [line width=1, gray, dashed,domain=(-9:9),<->] ({x},{0});

           

  \end{axis}
\end{tikzpicture}
\end{image}


The graphs of exponential functions (and shifted exponential functions) have a horizontal asymptote. 


The graphs of logarithmic functions have a vertical asymptote. 




We get general logarithmic functions by composing with linear functions.













\section*{General Logarithmic Functions}


Logarithmic functions are functions taht can bre represented by formulas of the form


\[  A \log_r(B \, x + C) + D  \]



These templates can be viewed as a composition.





\begin{itemize}
\item Let $\log_r(t)$ be a Core logarithmic function.
\item Let $L_{in}(u) = B \, u + C$ be a linear function.
\item Let $L_{out}(v) = A \, v$ be a linear function.
\end{itemize}

Then, our general logarithmic function can be expressed as a composition.


\[
 A \log_r(B \, x + C) + D = L_{out}(v) \circ \log_r(t) \circ L_{in}(x)
\]



Now we can use the Chain Rule to establish whether a logarithmic function increases or decreases.



















\begin{example} Logarithmic Function



Here is the graph of $y = L(x) = -3 \log_2(5-x)+1$.

\begin{image}
\begin{tikzpicture} 
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel=$y$,
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=1, gray, dashed,domain=(-9:9),<->] ({5},{x});
          
          \addplot [line width=2, penColor, smooth,samples=200,domain=(-5:4.8),<->] {-3*ln((5-x)/ln(2)+1};
          

           

  \end{axis}
\end{tikzpicture}
\end{image}




Analysis:

\textbf{Domain:} The inside must be positive. This tells us that the domain is $(-\infty, 5)$.

\textbf{Zero:} 

\begin{align*}
-3 \log_2(5-x)+1 &= 0 \\
\log_2(5-x)      &= \frac{1}{3} \\
5-x              &= 2^{\tfrac{1}{3}} \\
5-2^{\tfrac{1}{3}}    &= x 
\end{align*}

Just Checking:  $5-2^{\tfrac{1}{3}} \approx 3.74 \in (-\infty, 5)$ and agrees with the graph.


\textbf{Continuity:} Logarithmic functions are continuous.

\textbf{Behavior:}

We can view $L(x) = -3 \log_2(5-x)+1$ as a composition of three functions.



\begin{itemize}
\item Let $\log_2(t)$, is an increasing Core logarithmic function.
\item Let $L_{in}(u) = 5 - u$, is a decreasing linear function.
\item Let $L_{out}(v) = -3 \, v + 1$, is a decreasing linear function.
\end{itemize}

Then, our general logarithmic function can be expressed as a composition.


\[
 L(x) = -3 \log_2(5-x)+1 = L_{out}(v) \circ \log_2(t) \circ L_{in}(x)
\]


The Chain Rule tells us that $dec \circ inc \circ dec = inc$

$L(x)$ is an increasing function.

Just Checking: This agrees with the graph.





\textbf{End-Beavior:} $L$ only has end-behavior in only one direction.

The domain of $L$ is $(-\infty, 5)$ and $L$ is an increasing logarithmic function.  The gives

\[ \lim\limits_{x \to -\infty} L(x) = -\infty \]







\textbf{Singularity Beavior:} 


The domain of $L$ is $(-\infty, 5)$ and $L$ is an increasing logarithmic function.  The gives

\[ \lim\limits_{x \to 5^-} L(x) = \infty \]





\textbf{Global Maximum and Minimum:}  Logarithmic functions do not have a global maximum or global minimum.

\textbf{Local Maximums and Minimums:} Logarithmic functions do not have local maximums or minimums.

\textbf{Range:} The range of any logarithmic function is either $(-\infty, \infty)$.




\end{example}











\subsection*{Behavior}



Logarithmic functions either increase or they decrease.

The domain and behavior can tell you which.


This can also be summed up by comparing the base, the leading coefficient, and the leading coefficient of the linear inside function.



Our general template for logarithmic functions looks like

\[
log(x) = A \cdot \log_r(B \, x + C) + D
\]



If the base $r > 1$, then

\begin{itemize}
  \item $A > 0$ and $B > 0$ gives an increasing logarithmic function.
  \item $A < 0$ and $B > 0$ gives a decreasing logarithmic function.
  \item $A > 0$ and $B < 0$ gives a decreasing logarithmic function.
  \item $A < 0$ and $B < 0$ gives an increasing logarithmic function.
\end{itemize}





If the base $r < 1$, then everything is reversed (per the Chain Rule)

\begin{itemize}
  \item $A > 0$ and $B > 0$ gives an decreasing logarithmic function.
  \item $A < 0$ and $B > 0$ gives a increasing logarithmic function.
  \item $A > 0$ and $B < 0$ gives a increasing logarithmic function.
  \item $A < 0$ and $B < 0$ gives an decreasing logarithmic function.
\end{itemize}


$\blacktriangleright$ When the leading coefficients are the same sign, then the behavior of the logarithmic function is the same as the Core function. 

$\blacktriangleright$ When the leading coefficients are different signs, then the behavior of the logarithmic function is opposite the Core function.






\subsection*{Mental Models}



Just like with exponential functions, it is nice to have a mental model in your head of a basic logarithmic function an its characteristics. 



And, like with exponential functions, we like $e$ as the base. 


This gives us a basic basic logarithmic function: $\ln(x)$ 



And, then three other alternative choices, if you prefer. 


\[
\ln(x) \, \text{ or } \, \ln(-x) \, \text{ or } \, -\ln(x) \, \text{ or } \, -\ln(-x) 
\]



\begin{image}
\begin{tikzpicture}
    \begin{axis}[name = basictop, 
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel=$y$, 
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=2, penColor, smooth,samples=300,domain=(0.02:9),<->] {ln(x)};
          \addplot [line width=1, gray, dashed,domain=(-9:9),<->] ({0},{x});
    \end{axis}
    \begin{axis}[at={(basictop.outer east)},anchor=outer west, 
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel=$y$, 
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=2, penColor, smooth,samples=200,domain=(-9:-0.02),<->] {ln(-x)};
          \addplot [line width=1, gray, dashed,domain=(-9:9),<->] ({0},{x});
    \end{axis}




\end{tikzpicture}
\end{image}






\begin{image}
\begin{tikzpicture}
    \begin{axis}[name = basictop, 
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel=$y$, 
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=2, penColor, smooth,samples=300,domain=(0.02:9),<->] {-ln(x)};
          \addplot [line width=1, gray, dashed,domain=(-9:9),<->] ({0},{x});
    \end{axis}
    \begin{axis}[at={(basictop.outer east)},anchor=outer west, 
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel=$y$, 
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=2, penColor, smooth,samples=200,domain=(-9:-0.02),<->] {-ln(-x)};
          \addplot [line width=1, gray, dashed,domain=(-9:9),<->] ({0},{x});
    \end{axis}




\end{tikzpicture}
\end{image}













\begin{summary}\textbf{\textcolor{red!70!yellow}{Logarithmic Analysis}}


\textbf{Domain:} The domain of a logarithmic funciton is all real numebrs that make the inside function positive.

\textbf{Zeros:} Every logarithmic funciton has exatly one zero.

\textbf{Continuity:} All logarithmic functions are continuous.

\textbf{End-Behavior:} Logarithmic functions have end-behavior in only one direction.  This behavior is also unbounded behavior, either positive or negative.

In the other direction is the singularity. The singularity behavior is opposite of the end-behavior.


\textbf{Behavior (Increasing and Decreasing):} A logarithmic function is either increasing or it is decreasing.  The Chain Rule can tell you which.

\textbf{Global Maximum and Minimum:} Logarithmic funcitins do not have a global maximum or minium.

\textbf{Local Maximums and Minimums:}  Logarithmic functions do not have local maximums or miniums.

\textbf{Range:} The range of a logarithmic function is always $(-\infty, \infty)$.



\end{summary}







\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Analysis]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/libraryAnalysis2/examples/exampleList}

\end{center}




\end{document}
