\documentclass{ximera}

\input{../../preamble.tex}

\title{Basic Forms}

\begin{document}

\begin{abstract}
base e
\end{abstract}
\maketitle



Algebra allows us to represent the same function with different formulas.

Exponential functions have several ``forms'', depending on how you like to think about them. Our two favorites are

\[     A \cdot r^x = a \cdot r^{b \, y + c}   \]



\textbf{\textcolor{blue!55!black}{A Different Perspective}} 


Exponential functions, are those functions which \textbf{\textcolor{red!70!black}{CAN}} be represented by formulas of the form $A \cdot r^x$ with $A \ne 0$.  


We can decide whether the function is increasing or decreasing by the value of $r$ and the sign of $A$. 



\begin{itemize}
\item $A > 0$ and $r > 1$ : increasing positive function
\item $A < 0$ and $r > 1$ : decreasing negative function  
\item $A > 0$ and $r < 1$ : decreasing positive function
\item $A < 0$ and $r < 1$ : increasing negative function  
\end{itemize}



\textbf{On the other hand,} we have the algebra rule $\frac{1}{b} = b^{-1}$.  We could think of the base of the exponential formula as always being greater than $1$, and use positive and negative exponents to switch between increasing and decreasing functions. 


Exponential functions, are those functions which \textbf{\textcolor{red!70!black}{CAN}} be represented by formulas of the form $A \cdot r^{B \, x}$, where $r > 1$ and $A \ne 0$.  


In this case, we would have the following behaviors: 


\begin{itemize}
\item $A > 0$ and $B > 0$ : increasing positive function
\item $A > 0$ and $B < 0$ : decreasing positive function
\item $A < 0$ and $B > 0$ : decreasing negative function  
\item $A < 0$ and $B < 0$ : increasing negative function  
\end{itemize}

We would decide function behavior (increasing or decreasing) by the signs of \textbf{both} leading coefficients, $A$ and $B$.

\textbf{\textcolor{red!70!darkgray}{$\blacktriangleright$}} If $A$ and $B$ are the same sign, then we have an increasing function. 

\textbf{\textcolor{red!70!darkgray}{$\blacktriangleright$}} If $A$ and $B$ are different signs, then we have a decreasing function. 








\section*{(e)}



Algebra always us to change our exponential functions to any base.  We might as well always choose bases greater than $1$.


If this is the case, then we might as well use $e$ as our base. 


\begin{explanation} \textbf{\textcolor{red!75!green}{Base $e$}} 


If our formula looks like. $A \cdot r^{B \, x}$ and $r>1$, then we can use a little algebra to rewrite our formula.



Since we are only using positive bases, we know that $r > 0$.  Since $r$ is a positive real number, we know that $r$ can be written as a power of $e$.


\[
r = e^{\ln(r)}
\]



We can rewrite our formula as



\[
a \cdot r^{b \, x} = a \cdot \left( e^{\ln(r)} \right)^{b \, x} = a \cdot  e^{\ln(r) \cdot (b \, x)} = a \cdot  e^{(\ln(r) \cdot b) \, x} = a \cdot  e^{B \, x}
\]


\end{explanation}




\textbf{\textcolor{red!70!darkgray}{$\blacktriangleright$}}  Exponential functions, are those functions which \textbf{\textcolor{red!70!black}{CAN}} be represented by formulas of the form $A \cdot e^{B \, x}$.  






In this model, our basic forms to memorize would be






\begin{image}
\begin{tikzpicture}
   \begin{axis}[name = leftgraph, 
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel={$e^x$},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=1, gray, dashed,samples=200,domain=(-10:10),<->] {0};

          \addplot [line width=2, penColor, smooth,samples=100,domain=(-9:8), <->] {1.3^x};
          \addplot [color=penColor,only marks,mark=*] coordinates{(0,1)};

           \addplot [line width=1, violet, dashed, domain=(-10:10)] ({10},{x});
           \addplot [line width=1, violet, dashed, domain=(-10:10)] ({-10},{x});
           \addplot [line width=1, violet, dashed, domain=(-10:10)] ({x},{10});
           \addplot [line width=1, violet, dashed, domain=(-10:10)] ({x},{-10});

           

  \end{axis}
  \begin{axis}[at={(leftgraph.outer east)},anchor=outer west, 
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel={$-e^x$},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=1, gray, dashed,samples=200,domain=(-10:10),<->] {0};

          \addplot [line width=2, penColor, smooth,samples=100,domain=(-9:8),<->] {-(1.3^x)};
          \addplot [color=penColor,only marks,mark=*] coordinates{(0,-1)};

           

  \end{axis}
\end{tikzpicture}
\end{image}









\begin{image}
\begin{tikzpicture}
   \begin{axis}[name = leftgraph, 
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel={$e^{-x}$},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=1, gray, dashed,samples=200,domain=(-10:10),<->] {0};

          \addplot [line width=2, penColor, smooth,samples=200,domain=(-8:9), <->] {1.3^(-x)};
          \addplot [color=penColor,only marks,mark=*] coordinates{(0,1)};
           

  \end{axis}
  \begin{axis}[at={(leftgraph.outer east)},anchor=outer west, 
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel={$-e^{-x}$},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=1, gray, dashed,samples=200,domain=(-10:10),<->] {0};

          \addplot [line width=2, penColor, smooth,samples=200,domain=(-8:9),<->] {-(1.3^(-x)};
          \addplot [color=penColor,only marks,mark=*] coordinates{(0,-1)};
           

  \end{axis}
\end{tikzpicture}
\end{image}





And finally, \textbf{\textcolor{red!70!darkgray}{$\blacktriangleright$}}  exponential functions, are those functions which \textbf{\textcolor{red!70!black}{CAN}} be represented by formulas of the form $A \cdot e^{B \, x + C}$.  

It just depends on how you want to think about them.














\subsection*{Pick Your Form}

You should pick your own basic exponential form that you like and understand.  

Then, you can change anything given to you into that form.









\begin{example} Base Greater than $1$


If you like bases that are greater than $1$, then you can change any formula for an exponential or shifted exponential function into a formula with a base greater than $1$. 


Given the formula 

\[
F(x) = 8 \cdot \left( \frac{2}{3} \right)^{2 \, x + 7}
\]



Rewrite 

\[
\left( \frac{2}{3} \right) = \left( \frac{3}{2} \right)^{-1}
\]


This replacement gives 



\[
F(x) = 8 \cdot \left( \frac{2}{3} \right)^{2 \, x + 7} = 8 \cdot \left( \left( \frac{3}{2} \right)^{-1} \right)^{2 \, x + 7}
\]

\[
F(x) = 8 \cdot \left( \frac{3}{2} \right)^{-2 \, x - 7}
\]


Now, we have a base greater than $1$.  The negative sign has been multiplied through the exponent.


\end{example}



Or, you might pick a particular number greater than $1$ as the base you like.  $e$ is a very popular choice, because it shows up quite frequently in mathematics, like Calculus.










\begin{example} Base $e$

If you prefer $e$ as your base...


Given the formula 

\[
G(t) = -4 \cdot 3^{8 \, t - 5}
\]



Rewrite 

\[
3 = e^{\ln(3)}
\]


This replacement gives 



\[
G(t) = -4 \cdot 3^{8 \, t - 5} = -4 \cdot \left( e^{\ln(3)} \right)^{8 \, t - 5}
\]

\[
G(t) = -4 \cdot e^{8 \ln(3) \, t - 5 \ln(3)}
\]


We have a new formnula for our function that uses base $e$.


\end{example}




If you pick a base you like and change formulas to use that base, then you have a better chance of quickly deciding on function characteristics. 



\textbf{\textcolor{blue!55!black}{Core Form:}}  $e^x$ 



\textbf{\textcolor{blue!55!black}{Basic Forms:}}  $e^x$, $e^{-x}$, $-e^x$, $-e^{-x}$ 


Now increasing and decreasing become questions about the signs of the two leading coefficents - the leading coefficient of the entire formula and the leading coefficient of the exponent, which is exactly what the Chain Rule tells us.














\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Exponential Functions]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/expFunctions/examples/exampleList}

\end{center}









\end{document}
