\documentclass{ximera}

\input{../../preamble.tex}

\title{Shifted Theory}

\begin{document}

\begin{abstract}
the shifted story
\end{abstract}
\maketitle





We have two levels of exponential formulas: the basic template and then the general template, which includes shifts and stretches.  All of these are based on our original idea of \textbf{\textcolor{purple!85!blue}{constant percentage growth}}.


Constant percentage growth means that the value  of our function grows by an equal percentage over equal size intervals in the domain.





\begin{example}  Shifted Exponential Function \\



Analyze   $f(x) = \frac{1}{3} \cdot \, 2^{x+5} - 7$ \\



We'll begin by thinking about the graph to help our analysis. \\



\begin{explanation}

Compared to a basic exponential function graph, where the horizontal axis is the horizontal asymptote, here, the horizontal asymptote has been moved down $\answer{7}$ units, because the constant term in the formula is $-7$.



The ``inside'' (the argument) of the formula for $f(x)$, representing the domain, is $x+5$.  This equals $0$, when $x=-5$.  The exponent is positive for $x>-5$, and because the base is $2 > 1$, this is the direction (right) of unbounded growth.  Therefore, the other direction (left) is where the horizontal asymptote is in effect.  Since the leading coefficient is $\frac{1}{3} > 0$, the unbounded growth is positive.

At $x=-5$, we have our one anchor point for the graph.  The point is $\left(-5, \frac{1}{3} - 7 \right)$, which is $\frac{1}{3}$ above the horizontal asymptote, $y = -7$.


Graph of $y = f(x)$.

\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel=$y$, grid = major,
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
          	xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
          	ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=1, gray, dashed,samples=200,domain=(-10:10),<->] {-7};

      	  \addplot [line width=2, penColor, smooth,samples=200,domain=(-10:0.6),<->] {0.33 * 2^(x+5)-7};

      	  \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(-5,-6.66)};

          


 

  \end{axis}
\end{tikzpicture}
\end{image}



\end{explanation}


Now, for our function analysis.



\begin{explanation}



\textbf{Domain:} The domain of every shifted exponential function is $(-\infty, \infty)$. \\



\textbf{Zeros:}  We see that the graph has a horizontal intercept, which means the function has a zero. \\


$f(x) = \frac{1}{3} \cdot \, 2^{x+5} - 7 = 0$ \\


\begin{align*}
\frac{1}{3} \cdot \, 2^{x+5} - 7 & = 0 \\
\frac{1}{3} \cdot \, 2^{x+5} & = 7 \\
2^{x+5} & = 21 \\
\answer{x+5} & = \log_2{21} \\
x & = \log_2{21} - 5
\end{align*}

\textbf{Note:}  $\log_2{21} - 5 \approx -0.607$, which agrees with the graph. \\


\textbf{Continuity:}  Shifted exponential functions are continuous.  \\



\textbf{End-Behavior:} The negative direction is the direction that makes the exponent negative.  That makes this the direction that the function's value levels off to the constant term, $-7$.


\[
\lim\limits_{x \to -\infty} f(x) = -7
\]




The other direction is where the shifted exponential function is uunbounded.  Since the leading coefficient of $f$ is positive, $f$ is unbounded positively. \\



\[
\lim\limits_{x \to \infty} f(x) = \infty
\]









\textbf{Behavior:} The leading coefficient of $f$ is $\frac{1}{3}$, which is positive and the leading coefficient of the argument is $1$, which is also positive.  Since these are the same sign, we know that $f$ is an increasing shifted exponential function.



\textbf{Global Extrema:} Shifted exponential functions do not have global maximums or minimums. \\


\textbf{Local Extrema:} Shifted exponentials function do not have local maximums or minimums. \\



\textbf{Range:} The end-behavior tells us that the range is $(-7, \infty)$.



\end{explanation}

\end{example}


The shape of the graph looks like that of the basic exponential graph. \\


\begin{itemize}
\item The horizontal asymptote moved.
\item The strategic point moved.
\end{itemize}

















$\blacktriangleright$ \textbf{\textcolor{blue!55!black}{What happened to stretching?}} \\



Basic exponential functions swallow up stretching and compressing in the leading coefficient and base.


\begin{itemize}
\item If we multiply on the outside, by $b$, for a vertical stretch, then we have

\[
b \cdot (a \cdot r^x) = (a \cdot b) \cdot r^x
\]

We obtain a new leading coefficient, which tells us about the sign of the function values. \\






\item If we multiply on the inside, by $b$, for a horizontal stretch, then our exponential rules give us

\[
a \cdot r^{b \cdot x} = a \cdot (r^b)^x = a \cdot R^x
\]

We obtain a new base, which tells us about the growth of the function.


\end{itemize}





Similarly, horizontal stretching can also be absorbed by the leading coefficent.

 \[
 a \cdot r^{x+c} = a \cdot r^x \cdot r^c = (a \cdot r^c) \cdot r^x
\]




We have a choice.  For a shifted exponential function, we can think about stretching or we can apply a little algebra and return to a basic exponential function.




















\begin{example} Another View



$f(x) = \frac{1}{3} \, 2^{x+5} - 7$ is the function in the last example.  We could take advantage of exponential algebra to simplify the formula.   \\

\[ 
f(x) = \frac{1}{3} \, 2^{x+5} - 7 = \frac{1}{3} \, 2^x \cdot 2^5 - 7 = \frac{32}{3} \, 2^x  - 7
\]


We now have a basic exponential function shifted down $7$.  The new base is still $2$ and the new leading coefficient is $\frac{32}{3}$.





By doing this, we change our strategic point to $\left( 0, \frac{32}{3} - 7 \right) = \left( 0, \frac{11}{3} \right)$.




\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel=$y$, grid = major,
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=1, gray, dashed,samples=200,domain=(-10:10),<->] {-7};

          \addplot [line width=2, penColor, smooth,samples=200,domain=(-10:0.6),<->] {0.33 * 2^(x+5)-7};

          \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(0,3.66)};

          


 

  \end{axis}
\end{tikzpicture}
\end{image}




\end{example}






















\begin{example} Altering Algebraically



Let $G(t) = 4 \cdot 3^{2t-1} - 5$.  

We could take advantage of exponential algebra to simplify the exponent.   \\

\[ 
G(t) = 4 \cdot 3^{2t-1} - 5 = 4 \cdot 3^{2t} \cdot \frac{1}{3} - 5 = \frac{4}{3} \cdot (3^2)^t - 5  = \frac{4}{3} \cdot 9^t - 5
\]


We now have a basic exponential function shifted down $\answer{5}$.  The new base is $\answer{9}$ and the new leading coefficient is $\answer{\frac{4}{3}}$.




The leading coefficient is $\frac{4}{3} > 0$ and the base is $9 > 1$. Therefore, $G(t)$ is an increasing function.  The function values will be positive, once they overcome the drop of $5$ to the horizontal asymptote.



$G(t) = \frac{4}{3} \cdot 9^t - 5$ is a transformation of $g(t) = 3^t$.  Let's compare $G(t)$ back to $3^t$.


Since the new base $9$ is greater than the old base of $3$, the function grows faster.  The graph goes up to the right faster and steeper.  This means the graph is compressed horizontally, which is the result of multiplying by $2$ in the exponent.





End-Behavior:
\[
\lim\limits_{t \to -\infty}G(t) = \answer{-5}  
\]

\[
\lim\limits_{t \to \infty}G(t) = \answer{\infty}
\]




\end{example}










\begin{observation} \textbf{\textcolor{blue!55!black}{Big Bases : $r > 1$}}     \\



When multiplying by $b$ on the inside of an exponential function, for a horizontal stretch or compression, 


if $1 < b$, then we are compressing horizontally.  This works out in the algebra. When $1 < b$, then $r < r^b$. The base becomes larger, which means the function grows faster, which means the graph goes up faster, which means the graph has been compressed horizontally. \\



if $0 < b < 1$, then we are stretching horizontally.  This works out in the algebra. When $0 < b < 1$, then $r^b < r$. The base becomes smaller, which means the function grows slower, which means the graph goes up slower, which means the graph has been stretched horizontally.


\end{observation}









\begin{example}  Shifted Exponential Function



Analyze   $B(t) = -2 \, \left( \frac{2}{3} \right)^{3-t} + 4$ \\


\begin{explanation}


\textbf{\textcolor{purple!85!blue}{Formula dissection:}}  \\


$\blacktriangleright$  the base: $\frac{2}{3} < 1$\\
$\blacktriangleright$  the exponent: $3-t$ is positive for large negative $t$. \\
$\blacktriangleright$  the exponent: $3-t$ is negative for large positive $t$. \\
$\blacktriangleright$  the leading coefficient is negative. \\


Together, these tell us that $B(t)$ settles down for large negative $t$ and that $B(t)$ becomes unbounded when $t$ is large and positive.



This is a transformed version of the basic exponential function template $\left( \frac{2}{3} \right)^{-t} = \left( \answer{\frac{3}{2}} \right)^t$.  



When $t < 0$, then $-t > 0$ and we get  $\left( \frac{2}{3} \right)^{positive}$ and the basic exponential portion is becoming smaller, approaching $0$.  





\[ \lim\limits_{t \to -\infty} \left( \frac{2}{3} \right)^{-t}  = 0 \]



When $t > 0$, then $-t < 0$ and we get  $\left( \frac{2}{3} \right)^{negative} = \left( \frac{3}{2} \right)^{positive}$ and $B(t)$ is becoming unbounded.  



\[ \lim\limits_{t \to \infty} \left( \frac{2}{3} \right)^{-t}  = \lim\limits_{t \to \infty} \left( \frac{3}{2} \right)^t  = \infty \]




Exponential growth to the right and decay to the left.






Since $a = -2 < 0$, these smaller/larger values of $-2 \, \left( \frac{2}{3} \right)^{-t}$ are smaller/larger negative values.



We also have two shifts. One from the $3$ in the exponent and the second from the constant term, $4$:




$\blacktriangleright$ \textbf{Vertical Shift}


Adding $4$ to the outside shifts the graph vertically up $4$.  The asymptote is $y = 4$ and 

\[ \lim\limits_{t \to -\infty} B(t) = 4 \]



$\blacktriangleright$ \textbf{Horizontal Shift}

Our exponent here is $3 - t$.  Our function's exponent is zero, $3-t=0$ when $t=3$. Our one anchor point is shifted over to $3$.  Multipying by $-2$, means the dot is $2$ away(below) from the horizontal asymptote, which is now $y=4$.  Our anchor point is $(3, 2)$.








Graph of $y = B(t)$.

\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$t$, ylabel=$y$, grid = major,
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
          	xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
          	ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=1, gray, dashed,samples=200,domain=(-10:10),<->] {4};

      	  \addplot [line width=2, penColor, smooth,samples=200,domain=(-10:7.5),<->] {-2 * (0.666^(3-x)) + 4};

      	  \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(3,2)};

          


  \end{axis}
\end{tikzpicture}
\end{image}




Our analysis tells us that 

\begin{itemize}
\item The natural or implied domain of $B$ is $\mathbb{R}$.
\item $B$ is always decreasing.
\item $B$ has no maximums or minimums.
\item $\lim\limits_{t \to -\infty} B(t) = 4$
\item $\lim\limits_{t \to \infty} B(t) = -\infty$
\end{itemize}


\end{explanation}

\end{example}





In the example above, we could have algebraically moved the horizontal shift to the leading coefficient.



\[
-2 \, \left( \frac{2}{3} \right)^{3-t} + 4
\]


\[
-2 \, \left( \frac{2}{3} \right)^{3} \cdot \left( \frac{2}{3} \right)^{-t}+ 4
\]


\[
-2 \, \left( \frac{8}{27} \right)  \cdot \left( \frac{2}{3} \right)^{-t}+ 4
\]


\[
\left( \frac{-16}{27} \right)  \cdot \left( \frac{2}{3} \right)^{-t}+ 4
\]


We could have further transfered the negative sign in the exponent to the base by reciprocating the base.

\[
\left( \frac{-16}{27} \right)  \cdot \left( \frac{3}{2} \right)^t+ 4
\]



Algebra provides many tools for modifying the representing formula and altering how we think about the behavior. \\








\begin{example}  Shifted Exponential Function



Analyze   $K(f) = 3^{5-f} - 5$ \\

\begin{question}
 

The base is \wordChoice{\choice{less than}\choice[correct]{greater than}} $1$.
\end{question}
\begin{question}


The exponent gets big and positive when $f$ gets big and \wordChoice{\choice{positive}\choice[correct]{negative}}.
\end{question}
\begin{question}


The graph will become unbounded to the \wordChoice{\choice{right}\choice[correct]{left}}.\\
\end{question}

\begin{question}


The horizontal asymptote is $y = \answer{-5}$.
\end{question}

\begin{question}


The graph will approach the asymptote to the \wordChoice{\choice[correct]{right}\choice{left}}.\\
\end{question}
\begin{question}


Our one strategic anchor point moves to $\left(\answer{5}, \answer{-4}\right)$.
\end{question}
\begin{question}


The graph will become unbounded \wordChoice{\choice[correct]{up}\choice{down}}.\\
\end{question}




Graph of $y = K(f)$.

\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$f$, ylabel=$y$, grid = major,
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
          	xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
          	ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=1, gray, dashed,samples=200,domain=(-10:10),<->] {-5};

      	  \addplot [line width=2, penColor, smooth,samples=200,domain=(2.75:10),<->] {3^(5-x) - 5};

      	  \addplot[color=penColor,fill=penColor,only marks,mark=*] coordinates{(5,-4)};

 

  \end{axis}
\end{tikzpicture}
\end{image}





Our analysis tells us that

\begin{itemize}
\item The natural or implied domain of $K$ is $\mathbb{R}$.
\item $K$ is always decreasing.
\item $K$ has no maximums or minimums.
\item $\lim\limits_{f \to -\infty} K(f) = \infty$
\item $\lim\limits_{f \to \infty} K(f) = -5$
\end{itemize}


\end{example}





















\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Exponential Functions]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/expFunctions/examples/exampleList}

\end{center}









\end{document}
