\documentclass{ximera}

\input{../../preamble.tex}

\title{Complex Multiplication}

\begin{document}

\begin{abstract}
products
\end{abstract}
\maketitle


Multiplication of complex numbers looks weird.





\begin{definition} \textbf{\textcolor{green!50!black}{Complex Multiplication}}


\[    (a + b \, i) \cdot (c + d \, i) = (ac-bd) + (ad+bc) \, i           \]

\end{definition}


It is defined this way to maintain the distributive property, which we want to keep from the real numbers.  The distributive property is vital to our algebra.



$\blacktriangleright$ Division requires a little discussion.



With real numbers, we have stories that give meaning to division.


$10 \div 5$ has two meanings.  This expresion might represent how many groups are needed to split a group of $10$ marbles into groups of size $5$.  That would be $2$ groups.  It could also represent how many marbles would be in each group, if $10$ marbles were split up into $5$ equal groups.  That would be $2$ marbles per group.


But, even real numbers stretch this story.  $10 \div \sqrt{2}$ ?  What does it mean to split a pile of marbles into $\sqrt{2}$ piles?

This marble story is also going to break with complex numbers.  Fortunately, the real numbers have a different view of division, which we can bring with us over to complex numbers.





Just from a strictly algebaic viewpoint, $10 \div 5$ represents the number that you multiply $5$ by, to get $10$.  In this way, we can think about division in terms of multiplication.  This thought leads to reciprocals and fractions.  Luckily, every real number, except $0$ has a reciprocal.  




\textbf{Note:}  $10 \div 0$ has no meaning in real numbers, because there is no real number that you can mulitply by $0$ and get $10$. \\









What about complex numbers?  Do they have reciprocals?  Does each complex number have a partner complex number such that their product equals $1$?



 $\blacktriangleright$ Can we solve $1 = (2 + 3 \, i) \cdot (a + b \, i)  $?



\begin{example} Multiplicative Inverse


Solve $1 = (2 + 3 \, i) \cdot (a + b \, i)  $?


\begin{explanation}

\[   1 = (2 + 3 \, i) \cdot (a + b \, i)  =  (2a - 3b) + (2b+3a) \, i      \]






We need $1 = \answer{2a - 3b}$ and $0 = \answer{2b+3a}$.


The second  equation tells us that $\frac{-3a}{2} = b$.  Substituting this into the first equation tells us that $1 = 2a - 3 \cdot \frac{-3a}{2}$.  Now we can solve for $a$.



\begin{align*}
1     & = 2a - 3 \left( \frac{-3a}{2} \right)  \\
      & =   2a + \answer{\frac{9a}{2}}   \\
      & =    \frac{13a}{2}   \\
  \frac{2}{13}    & =  a
\end{align*}




With that, we can get a value for $b$.   


\[ b = \frac{-3a}{2} = \frac{2}{13}  \cdot \frac{-3}{2} = -\frac{3}{13} \]

Let's check:



\[    (2 + 3 \, i) \cdot \left(\frac{2}{13} - \frac{3}{13} \, i \right)     =   \left( 2 \cdot  \frac{2}{13} - 3 \cdot \left(- \frac{3}{13}\right)\right)  + \left( 3 \cdot \frac{2}{13} - 2 \cdot \frac{3}{13} \right) \, i = \frac{4+9}{13} + \frac{6-6}{13} \, i = 1  \]



\end{explanation}



\end{example}


Does this process work for all nonzero complex numbers? \\


$\blacktriangleright$  We know it works for $r + 0 \, i$ with $r \ne 0$, because these are real numbers.


\[   (r + 0 \, i) \cdot \left(\frac{1}{r} + 0 \, i \right) = 1        \]







$\blacktriangleright$  Similarly, it works for $0 + r \, i$ with $r \ne 0$.


\[   (0 + r \, i) \cdot \left(0 - \frac{1}{r} \, i \right) = 1        \]





$\blacktriangleright$ Now for complex numbers where neither component is $0$


\begin{explanation}


Given real numbers $A \ne 0$ and $B \ne 0$, solve the following for $x$ and $y$.


\[       1 = (A + B \, i) \cdot (x + y \, i)            \]



\begin{align*}
1          & = (A + B \, i) \cdot (x + y \, i)      \\
           & = (Ax-By) + (Bx+Ay) \, i
\end{align*}


For this to work, we need $\answer{A x - B y} = 1$ and $\answer{B x + A y} = 0$.



The second equation tells us that $x = \answer{-\frac{A y}{B}}$, which is ok, since $B \ne 0$ in this case.

Substituting that into the first equation, for $x$, gives us


\[   A \cdot \left(-\frac{A y}{B}\right) - B y = 1     \]


\[   -\frac{A^2 y}{B} - B y = 1     \]

\[   -\frac{A^2 y}{B} - \frac{B^2}{B} y = 1     \]


\[   -\frac{A^2 + B^2}{B} y = 1     \]


\[  y = \frac{-B}{A^2 + B^2}     \]

This is ok, since $A^2 + B^2 \ne 0$, because neither $A$ nor $B$ equals $0$.



This gives us $x = -\frac{A y}{B} = -\frac{A}{B} \cdot \frac{-B}{A^2 + B^2} = \frac{A}{A^2 + B^2} $.







\[       (A + B \, i) \cdot \left( \frac{A}{A^2 + B^2} - \frac{B}{A^2 + B^2} \, i \right) = 1            \]



\end{explanation}












$\blacktriangleright$  Every nonzero complex number has a reciprocal.  We will use this to define division as multiplication by the reciprocal.





\section*{Complex Conjugate}


The \textbf{modulus} of a complex number is its distance from the origin or $0$.  The symbol for the modulus of a complex number, $z$, is $|z|$.  \\

If $z = a + b \, i$, then

\[    |z| = |a + b \, i| = \sqrt{a^2 + b^2}          \]



This is helpful, because a sum of two squares factors over the Complex numbers.



\[   (a + b \, i) \cdot    (a - b \, i)  = a^2 + b^2   \]



$\blacktriangleright$  $a - b \, i$ is called the \textbf{complex conjugate} of $a + b \, i$ and it will help us with division, by converting complex denominators to real numbers.




\[   \frac{1}{a + b \, i} =   \frac{1}{a + b \, i} \cdot 1 = \frac{1}{a + b \, i} \cdot \frac{a - b \, i}{a - b \, i}  =  \frac{a - b \, i}{(a + b \, i)(a - b \, i)} =   \frac{a - b \, i}{a^2 + b^2}  \]


\[  \frac{1}{a + b \, i}   =  \frac{a}{a^2 + b^2} - \frac{b}{a^2 + b^2} \, i      \]


We can use the complex conjugate of the denominator to get us back to standard form.















\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Complex Arithmetic]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/complexArithmetic/examples/exampleList}

\end{center}


\end{document}
