\documentclass{ximera}

\input{../../preamble.tex}

\title{Visual Features}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}
\maketitle






\begin{warning}


This section talks about deciphering function information from the graph of a function, \textbf{\textcolor{red!80!black}{which is not what we want}}.  Our goal is exactness, which can only be achieved through algebra and functional reasoning.\\

However, students have extensive experience with graphs, which makes graphs a good starting point. \\

Our plan is to use graphs to understand the ideas of function analysis and let that guide our algebraic analysis, \textbf{\textcolor{red!80!black}{which is what we want}}.

\end{warning}





Our word for investigating a function is \textbf{\textcolor{blue!55!black}{analysis}}. \\

We have an official list of function characteristics we would like to describe in our analysis. \\






\begin{itemize}
\item \textbf{\textcolor{blue!55!black}{Domain}}
\item \textbf{\textcolor{blue!55!black}{Zeros}}
\item \textbf{\textcolor{blue!55!black}{Continuity}}
\item \textbf{\textcolor{blue!55!black}{End-Behavior}}
\item \textbf{\textcolor{blue!55!black}{Behavior}}
\item \textbf{\textcolor{blue!55!black}{Global Extrema}}
\item \textbf{\textcolor{blue!55!black}{Local Extrema}}
\item \textbf{\textcolor{blue!55!black}{Range}}
\end{itemize}


In this section, we will explore the first four items on this list.









\subsection*{Learning Outcomes}

\begin{sectionOutcomes}
In this section, students will...

\begin{itemize}
\item identify domain.
\item determine zeros.
\item describe end-behavior.
\item explore discontinuities and singularities.
\end{itemize}
\end{sectionOutcomes}























\pdfOnly{
\begin{center}
more examples can be found in the online edition.
\end{center}
}




\begin{onlineOnly}
\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Visual Features]{https://ximera.osu.edu/csccmathematics/precalculus/precalculus/visualFeatures/examples/exampleList}

\end{center}
\end{onlineOnly}





\end{document}
