\documentclass{ximera}

\input{../../preamble.tex}

\title{Analyzing}

\begin{document}

\begin{abstract}
describe everything
\end{abstract}
\maketitle







$\blacktriangleright$ \textbf{\textcolor{red!80!black}{Reasoning:}} Reasoning is a logical explanation that describes our conclusions, how we arrived at those conclusions, and why we think those conclusions are correct. \\

Analysis is not a list of conclusions. We are not looking for such a list. \\

We are looking for the thought process that arrived at the list of conclusions. \\











\begin{example}

\textbf{\textcolor{purple!85!blue}{Completely analyze}} \\


\[   R(t) = \frac{(t+3)(t+1)}{(t-2)(t-3)} \]

with 

\[   R'(t) = -\frac{9 t^2 -6 t -39}{(t-2)^2 (t-3)^2} \]



$R(t)$ is not a composition. It is a rational function. \\


Here is a quick DESMOS graph.


\begin{center}
\desmos{mgkklcsmqr}{400}{300}
\end{center}


There is something wrong with this graph.  $2$ and $3$ make the denominator $0$, but the numbers in the interval $(2, 3)$ are in the domain.  There should be a piece of the graph there that is not showing up.


Let's evalaute at $2.5$ and see what type of values we are getting. \\

$R(2.5) = -77$


\begin{center}
\desmos{8baq5avdvr}{400}{300}
\end{center}


There is the other part of the graph. \\



From the graph, we are anticipating two singularities, two critical numbers, switching between increasing and decreasing, and end-behavior which is not unbounded. \\







































\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ Domain}} 

$R(t)$ is a rational function.  Its domain is all real numbers except the zeros of the denominator. \\

The domain is $(-\infty, 2) \cup (2,3) \cup (3, \infty)$  \\














\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ Zeros}} 


$R(t)$ is a rational function.  Its zeros are the zeros of the numerator except those that are also zeros of the denominator. \\


The zeros of $R$ are $-3$ and $-1$. \\






\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ Continuity}} 


$R(t)$ is a rational function.  It is continuous. \\


$R$ does have singularities.  We need the behavior around them.   Both $2$ and $3$ are zeros of the denominator and not zeros of the numerator.  This makes them asymptotic singularities. \\


We knw know that $R$ is unbounded as $t$ approaches $2$ and $3$.  We need to figure out the sign. \\


\begin{explanation}


For $\lim\limits_{t \to 2^-}R(t)$,  $t < 2$, which gives

\begin{itemize}
\item $t + 3 > 0$
\item $t + 1 > 0$
\item $t - 2 < 0$
\item $t - 3 < 0$
\end{itemize}


The sign of $R$ is $\frac{pos \cdot pos}{neg \cdot neg} = positive$ \\



\[
\lim\limits_{t \to 2^-}R(t) = \infty
\]






For $\lim\limits_{t \to 2^+}R(t)$,  $t > 2$, which gives

\begin{itemize}
\item $t + 3 > 0$
\item $t + 1 > 0$
\item $t - 2 > 0$
\item $t - 3 < 0$
\end{itemize}


The sign of $R$ is $\frac{pos \cdot pos}{pos \cdot neg} = negative$ \\



\[
\lim\limits_{t \to 2^+}R(t) = -\infty
\]





For $\lim\limits_{t \to 3^-}R(t)$,  $t < 3$, which gives

\begin{itemize}
\item $t + 3 > 0$
\item $t + 1 > 0$
\item $t - 2 > 0$
\item $t - 3 < 0$
\end{itemize}


The sign of $R$ is $\frac{pos \cdot pos}{pos \cdot neg} = negative$ \\



\[
\lim\limits_{t \to 3^-}R(t) = -\infty
\]







For $\lim\limits_{t \to 3^+}R(t)$,  $t > 3$, which gives

\begin{itemize}
\item $t + 3 > 0$
\item $t + 1 > 0$
\item $t - 2 > 0$
\item $t - 3 > 0$
\end{itemize}


The sign of $R$ is $\frac{pos \cdot pos}{pos \cdot pos} = positive$ \\



\[
\lim\limits_{t \to 3^+}R(t) = \infty
\]








\end{explanation}



This agrees with the graph.






\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ End-Behavior}} 

$R(t)$ is a rational function. The degree of the numerator and the degree of the denomiator are equal.  The end-behavior is the quotient is the leading coefficients.



\[  
\lim\limits_{t \to -\infty}R(t) = \frac{1}{1} = 1
\]




\[  
\lim\limits_{t \to \infty}R(t) = \frac{1}{1} = 1
\]











\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ Behavior}} 
\textbf{Rate-of-Change}  
\textbf{Increasing and Decreasing}   



The sign of $R'(t)$ will give us the behavior of $R(t)$. \\

To get the sign of $R'(t)$, we will factor it. \\

The denominator is factored.  The numerator is a quadratic function.  To factor it, we need its zeros, which we can get from the quadratic formula. \\

But, first we might as well factor out $3$ in the numerator. \\







\[   
R'(t) = -\frac{9 t^2 -6 t -39}{(t-2)^2 (t-3)^2} = -\frac{3 (3 t^2 - 2 t - 13)}{(t-2)^2 (t-3)^2}
\]




$3 (3 t^2 - 2 t - 13) = 0$ \\

$3 t^2 - 2 t - 13 = 0$ 



\[
t = \frac{2 \pm \sqrt{(-2)^2 - 4(3)(-13)}}{2(3)} = \frac{2 \pm \sqrt{160}}{6} = \frac{2 \pm 4 \sqrt{10}}{6} = \frac{1 \pm 2 \sqrt{10}}{3}
\]




Two critical numbers: $\frac{1 - 2 \sqrt{10}}{3}$ and $\frac{1 + 2 \sqrt{10}}{3}$ \\




Quick Check:  \\


\[
\frac{1 - 2 \sqrt{10}}{3} \approx -1.774851773
\]


\[
\frac{1 + 2 \sqrt{10}}{3} \approx 2.44151844
\]


These agree with the graph, so we feel like our algebra is correct. \\



We can now factor $R'(t)$ \\








\[   
R'(t) = -\frac{3 \left(t - \left(\frac{1 - 2 \sqrt{10}}{3} \right) \right) \left(t - \left(\frac{1 + 2 \sqrt{10}}{3} \right) \right) }{ (t-2)^2 (t-3)^2 } 
\]


Both $\frac{1 - 2 \sqrt{10}}{3}$ and $\frac{1 - 2 \sqrt{10}}{3}$ have odd multiplicities, so $R'$ will change signs across them. \\




Both $2$ and $3$ have even multiplicities, so $R'$ will not change signs across them. \\



We need the order of $\frac{1 - 2 \sqrt{10}}{3}$, $\frac{1 - 2 \sqrt{10}}{3}$, $2$, and $3$.   \\



\begin{explanation}

That graph makes us think that 

\[
\frac{1 - 2 \sqrt{10}}{3} < 2 < \frac{1 + 2 \sqrt{10}}{3} < 3
\]


Let's show that $\frac{1 - 2 \sqrt{10}}{3} < 2$. \\

This true because $\frac{1 - 2 \sqrt{10}}{3} < 0$, since $1 < 2 \sqrt{10}$. \\



Next, we'll show that $2 < \frac{1 + 2 \sqrt{10}}{3}$. \\



This is true if $6 < 1 + 2 \sqrt{10}$, which is true since $3 < \sqrt{10}$. \\



Finally, we need to show that $\frac{1 + 2 \sqrt{10}}{3} < 3$ \\


This is true if $1 + 2 \sqrt{10} < 9$, which is true since $\sqrt{10} < 4$.





\end{explanation}




We now know the ordering of the critical numbers and singularities. \\


\[
\frac{1 - 2 \sqrt{10}}{3} < 2 < \frac{1 + 2 \sqrt{10}}{3} < 3
\]





On $\left( -\infty, \frac{1 - 2 \sqrt{10}}{3} \right)$, 



\begin{itemize}
\item $t < \frac{1 - 2 \sqrt{10}}{3}$ and $\left( t - \frac{1 - 2 \sqrt{10}}{3} \right) < 0$
\item $t < 2$ and $(t - 2) < 0$
\item $t < \frac{1 + 2 \sqrt{10}}{3}$ and $\left( t - \frac{1 + 2 \sqrt{10}}{3} \right) < 0$
\item $t < 3$ and $(t - 3) < 0$
\end{itemize}

Note: $(t-2)$ and $(t-3)$ are squared in the denominator, making the denominator positive. \\

\[   
R'(t) = -\frac{3 \left(t - \left(\frac{1 - 2 \sqrt{10}}{3} \right) \right) \left(t - \left(\frac{1 + 2 \sqrt{10}}{3} \right) \right) }{ (t-2)^2 (t-3)^2 }  = neg \cdot \frac{pos \cdot neg \cdot neg}{pos \cdot pos} = neg
\]


$R$ is decreasing. \\













On $\left( \frac{1 - 2 \sqrt{10}}{3}, 2 \right)$, 


$\frac{1 - 2 \sqrt{10}}{3}$ has an odd multiplicity, so the sign of $R'$ switches, $R'(t) > 0$. \\




Or, 




\begin{itemize}
\item $t > \frac{1 - 2 \sqrt{10}}{3}$ and $\left( t - \frac{1 - 2 \sqrt{10}}{3} \right) > 0$
\item $t < 2$ and $(t - 2) < 0$
\item $t < \frac{1 + 2 \sqrt{10}}{3}$ and $\left( t - \frac{1 + 2 \sqrt{10}}{3} \right) < 0$
\item $t < 3$ and $(t - 3) < 0$
\end{itemize}

Note: $(t-2)$ and $(t-3)$ are squared in the denominator, making the denominator positive. \\



\[   
R'(t) = -\frac{3 \left(t - \left(\frac{1 - 2 \sqrt{10}}{3} \right) \right) \left(t - \left(\frac{1 + 2 \sqrt{10}}{3} \right) \right) }{ (t-2)^2 (t-3)^2 }  = neg \cdot \frac{pos \cdot pos \cdot neg}{pos \cdot pos} = pos
\]


$R$ is increasing. \\











On $\left( 2, \frac{1 + 2 \sqrt{10}}{3} \right)$, 


$2$ has an even multiplicity, so the sign of $R'$ does not switch sign, $R'(t) > 0$. \\


$R$ is increasing. \\






On $\left( \frac{1 + 2 \sqrt{10}}{3}, 3 \right)$, 


$\frac{1 + 2 \sqrt{10}}{3}$ has an odd multiplicity, so the sign of $R'$ switches sign, $R'(t) < 0$. \\


$R$ is decreasing. \\









On $(3, \infty)$, 


$3$ has an even multiplicity, so the sign of $R'$ does not switch sign, $R'(t) < 0$. \\


$R$ is decreasing. \\














\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ Global (Extrema) Maximum and Minimum}} 




From the behavior around the singularities, we hknow


\[
\lim\limits_{t \to 3^-}R(t) = -\infty
\]



\[
\lim\limits_{t \to 3^+}R(t) = \infty
\]


$R$ has no global maximum or minimum.



















\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ Local (Extrema) Maximum and Minimum}} 


We have two critical numbers.


On $\left( -\infty, \frac{1 - 2 \sqrt{10}}{3} \right)$, $R$ is decreasing. \\


On $\left( \frac{1 - 2 \sqrt{10}}{3} \right), 2$, $R$ is increasing. \\


$R$ has a local minimum value at $\frac{1 - 2 \sqrt{10}}{3}$.  \\






On $\left( 2, \frac{1 + 2 \sqrt{10}}{3} \right)$, $R$ is increasing. \\


On $\left( \frac{1 + 2 \sqrt{10}}{3} \right), 3$, $R$ is decreasing. \\


$R$ has a local maximum value at $\frac{1 + 2 \sqrt{10}}{3}$.  \\


















\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ Range}} 




The only thing we need to know is which is greater 

\[
R \left( \frac{1 - 2 \sqrt{10}}{3} \right) \, \text{ or } \, R \left( \frac{1 + 2 \sqrt{10}}{3} \right)
\]



THis will tell us if there is a gap in the range. \\
Hint: We can see a gap in the graph. \\

Instead of comparing these two function values directly, it might be easy to compare them to other numbers forcing a gap.


Like, we might show that 


\[
R \left( \frac{1 - 2 \sqrt{10}}{3} \right) > -10 > R \left( \frac{1 + 2 \sqrt{10}}{3} \right)
\]


\[
R\left( \frac{1 - 2 \sqrt{10}}{3}   \right) = \frac{\left( \frac{1 - 2 \sqrt{10}}{3} + 3 \right) \left( \frac{1 - 2 \sqrt{10}}{3} + 1 \right)}{\left( \frac{1 - 2 \sqrt{10}}{3} - 2 \right)  \left( \frac{1 - 2 \sqrt{10}}{3} - 3 \right)}
\]




\[
= \frac{\left( \frac{10 - 2 \sqrt{10}}{3}  \right) \left( \frac{4 - 2 \sqrt{10}}{3} \right)}{\left( \frac{-3 - 2 \sqrt{10}}{3} \right)  \left( \frac{-8 - 2 \sqrt{10}}{3} \right)}
\]




\[
= \frac{ (10 - 2 \sqrt{10}) (4 - 2 \sqrt{10}) }{ (-3 - 2 \sqrt{10}) (-8 - 2 \sqrt{10}) }
\]



\[
= \frac{ 80 - 28 \sqrt{10} }{ 64 + 22\sqrt{10} }
\]




\[
= \frac{ 80 - 28 \sqrt{10} }{ 64 + 22\sqrt{10} } \cdot \frac{ 64 - \sqrt{10} }{ 64 - 22\sqrt{10} } = \frac{11280 - 3552 \sqrt{10}}{-744}
\]



Is this greater than $-10$ ? \\


\[
\frac{11280 - 3552 \sqrt{10}}{-744} > -10
\]

\[
11280 - 3552 \sqrt{10} < -10 (-744)
\]


\[
11280 - 7440 < 3552 \sqrt{10}
\]


\[
3840 < 3552 \sqrt{10}
\]

This is certainly true, since  $3552 \cdot 2 = 7104 < 3552 \sqrt{10}$


Similarly,  $-10 > R \left( \frac{1 + 2 \sqrt{10}}{3} \right)$




The range has a gap.








The range of $R$ is 

\[
\left( -\infty, R \left( \frac{1 - 2 \sqrt{10}}{3} \right) \right) \cup \left( R \left( \frac{1 + 2 \sqrt{10}}{3} \right), \infty \right)
\]








\begin{center}

\textbf{\textcolor{red!70!black}{Everything agrees with the graph.}}


\end{center}






\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Analyzing Functions]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/analyzingFunctions/examples/exampleList}

\end{center}








\end{document}
