\documentclass{ximera}

\input{../../preamble.tex}

\title{Relations}

\begin{document}

\begin{abstract}

\end{abstract}
\maketitle




We can hardly think for a minute without asking a question.  All day long we think in terms of questions. What are we doing when we ask questions? How can they be that intrical to our view of the world? [ Two questions already. ] \\




\section{Anatomy of a Question}


What is the underlying assumption of a question?


\begin{observation} \textbf{\textcolor{blue!75!black}{Connecting Information}}  

\begin{itemize} 
\item What ingredients are needed to make chili?
\item Who was in House of Wax (1953)?
\item What is the speed limit on Route 3?
\item What is Camila's social security number?
\item What ice cream flavors does Kevin like?
\item What is $32^\circ F$ in Celsius?
\item Could you provide a list of good mystery books and their authors?
\item Did you pass Chemistry?
\item How high is Mt. Everest?
\item Who are Timothy's nieces?
\item What are the Post Office's hours?
\end{itemize}

$\blacktriangleright$We understand our world through connections between things. \\


We might view this as \textbf{\textcolor{purple!85!blue}{characteristics}} or \textbf{\textcolor{purple!85!blue}{attributes}} of something. 
\begin{itemize}
\item Mt Everest is $29,029 ft$ tall.
\end{itemize}


We might experience this as an \textbf{\textcolor{purple!85!blue}{if-then}} statement.
\begin{itemize}
\item If you are driving on Route 3, then do not go over 35 mph.
\end{itemize}



We might experience this as \textbf{\textcolor{purple!85!blue}{cause and effect}}.
\begin{itemize}
\item The Post Office is closed, because it is 9:00pm.
\end{itemize}

\end{observation}


All of our questions assume the same structure.  \\


\begin{quote} 
\textbf{\textcolor{green!50!black}{Questions assume there are two sets or collections of things and that the items from each set are connected or associated with items in the other collection.}} 
\end{quote}


\begin{example} Movies and Actors \\
There is a set or collection of movies and a set or collection of actors.  A movie is associated with an actor if the actor appeared in the movie.
\end{example}

\begin{example} SSN \\
There is a set of 9-digit numbers and a collection of U.S. citizens.  A number is connected to a person if the person was issued the number as their social security number.
\end{example}

\begin{example} Hours of Operation \\
There is a collection of Post Offices and a collection of times.  A time is connected to a Post Office if the Post Office is open for buiness during those hours.
\end{example}

There is certainly a pragmatic side to questions.


But, also, questions express our curiosity into the structure of how collections of things are connected. Mathematics is one way in which we describe such structures. Our mathematical word for these connected collections is a \textbf{\textcolor{blue!75!black}{relation}}. \\


This section explores \textbf{\textcolor{blue!75!black}{relations}} and how we communicate about them. \\










\subsection{Learning Outcomes}

\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item view a relation as a structured package.
\item encode relation pairings as written ordered pairs.
\item decipher written ordered pairs into relation pairings.
\item trace connections back-and-forth between sets.
\end{itemize}
\end{sectionOutcomes}







\begin{center}
\textbf{\textcolor{green!50!black}{ooooo=-=-=-=-=-=-=-=-=-=-=-=-=ooOoo=-=-=-=-=-=-=-=-=-=-=-=-=ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Relations]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/relations/examples/exampleList}

\end{center}







\end{document}
