\documentclass{ximera}

\input{../../preamble.tex}

\title{Categories}

\begin{document}

\begin{abstract}
elementary library
\end{abstract}
\maketitle




Success at analyzing functions rests entirely on our ability to recognize what type of function we have.  Categorizing functions is the quickest and best way of gaining information about our function.

Our library of elementary functions gives us a wonerful filter to begin idenitfying functions.  \\



\begin{template}  \textbf{\textcolor{blue!55!black}{Constant Functions}} \\



Constant funtions are functions which give the same value for all domain numbers. \\


They can be represented by formulas that look like


\[ f(x) = C  \,  \text{ for all } x  \, \text{ in the domain. }   \]



\end{template}



The natural domain of a constant function is all real numbers.  However, we can always state a restricted domain.




\begin{warning}   \textbf{\textcolor{red!80!black}{can}}  \\

All of our definitions for the elementary functions use the word \textbf{\textcolor{red!80!black}{can}}.


\begin{example}

The function representated by 

\[ f(x) = \frac{x^2 - 1}{x + 1} - x  \,  \text{ on } \,   [0, \infty)   \]

is a constant function.   $-1$ is the only function value.


\end{example}





\begin{example}

The function representated by 

\[ C(\theta) = (\cos(\theta))^2 + (\sin(\theta))^2   \,  \text{ on } \,   (-\infty, \infty)   \]

is a constant function.   $1$ is the only function value.


\end{example}




Constant functions are disguised with many formulas.  They \textbf{\textcolor{red!80!black}{can}} be written in the form $f(x) = C$.



\end{warning}




Graphs of constant functions are horizontal lines.
\[ f(x) = 3\]





\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel={$y=f(x)$}, grid = major, grid style={dashed},
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            yticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$}, 
            xticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
            
      \addplot [line width=2, penColor, smooth,samples=200,domain=(-9:9),<->] {3};




  \end{axis}
\end{tikzpicture}
\end{image}

















\begin{template}  \textbf{\textcolor{blue!55!black}{Linear Functions}} \\



Linear funtions are functions which exhibit a constant rate of change. \\


They can be represented by formulas that look like


\[ L(x) = A \, x + B  \,  \text{ for all } x  \, \text{ in the domain. }   \]



\end{template}



The natural domain of a linear function is all real numbers.  However, we can always state a restricted domain.







Graphs of linear functions are lines. 
(This make constant functions special linear functions)
\[ f(x) = 2x - 6\]





\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel={$y=f(x)$}, grid = major, grid style={dashed},
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            yticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$}, 
            xticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
            
      \addplot [line width=2, penColor, smooth,samples=200,domain=(-1.5:7),<->] {2*x-6};




  \end{axis}
\end{tikzpicture}
\end{image}






Although the definition quotes the slope-intercept form ($m x + b$), this is really not very useful.  In most situations, we do not have the intercept.  Instead, it is more common to have a slope and a random point or just two random points.



\begin{itemize}

\item \textbf{\textcolor{blue!55!black}{Point-Slope:}}  : $(y - y_0) = m (x - x_0)$

\item \textbf{\textcolor{blue!55!black}{Two Points:}}  : $(y - y_0) = \frac{y_1 - y_0}{x_1 - x_0} \, (x - x_0)$

\end{itemize}


All three of these forms are equivalent.  They can all be rearranged into the form of the others.  They are all reflecting the geometry of lines.  


On the other hand, we have seen that the analysis of functions often begins with zeros.  Products of factors is a preferred form.  With that in mind, linear functions can be written in factored form.


\[  f(x) = a (x - z_0)  \]

Where $a$ is the leading coefficient and $z_0$ is the zero of the function.





\begin{itemize}

\item $a = m$, the rate of change of the function is the slope of the line.

\item $z_0$ is the zero of the linear function, which is visually encoded as the intercept $(z_0, 0)$.

\end{itemize}



\textbf{Note:} Constant functions are linear functions, because they can be written as $L(x) = 0 x + C$. \\












\begin{template}  \textbf{\textcolor{blue!55!black}{Quadratic Functions}} \\



Quadratic funtions are the next level of polynomial. \\


They can be represented by formulas that look like


\[ f(x) = A \, x^2 + B \, x + C  \,  \text{ for all } x  \, \text{ in the domain. }   \]



\end{template}



The natural domain of a quadratic function is all real numbers.  However, we can always state a restricted domain.







Graphs of quadratic functions are parabolas. 
\[ f(x) = x^2 + x - 6\]





\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel={$y=f(x)$}, grid = major, grid style={dashed},
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            yticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$}, 
            xticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
            
      \addplot [line width=2, penColor, smooth,samples=200,domain=(-4:3),<->] {x^2 + x -6};




  \end{axis}
\end{tikzpicture}
\end{image}






Although the definition quotes the standard form ($A \, x^2 + B \, x + C$), there are more useful forms. 





\begin{itemize}

\item \textbf{\textcolor{blue!55!black}{Vertex Form:}}  $A \, (x - h)^2 + k$ 

\item \textbf{\textcolor{blue!55!black}{Factored Form:}}  $A \, (x - r_1)(x - r_2)$ 

\end{itemize}








\subsection*{Polynomial Functions}


\begin{template}  \textbf{\textcolor{blue!55!black}{Polynomial Functions}} \\



Constant, Linear, and Quadratic funtions are examples of polynomials. \\


Polynomials can be represented by formulas that look like


\[ f(x) = a_n x^n + a_{n-1} x^{n-1} + \cdots + a_1 x + a_0  \,  \text{ for all } x  \, \text{ in the domain. }   \]



\end{template}




The natural domain of a polynomial functions is all real numbers.  However, we can always state a restricted domain.




Although the definition quotes the standard form for a polynomial, the factored form is better suited for our analysis.

\[
p(x) = a (x - r_1)^{e_1} (x - r_2)^{e_2} \cdots (x - r_{k-1})^{e_{k-1}} (x - r_k)^{e_k}  
\]


\begin{itemize}
\item $r_1, r_2, \cdots r_k$ are the unique roots (zeros) of the polynomial.
\item $e_1, e_2, \cdots e_k$ are the exponents of the factors. They are called \textit{multiplicities}.
\end{itemize}



\begin{example}

If is difficult to tell anything from the standard form of this polynomial.

\[
x^8 - 22 x^7 +145 x^6 - 6 x^5 - 2997 x^4 + 6774 x^3 + 7931 x^2 - 15386 x - 13720
\]



Our analysis will benefit greatly from the factored form.


\[
(x+4) (x+1)^2 (x-2) (x-5) (x-7)^3
\]




\end{example}




















\subsection*{Rational Functions}




\begin{template}  \textbf{\textcolor{blue!55!black}{Rational Functions}} \\



Rational functions are quotients of polynomials. \\


Rational functions can be represented by formulas that look like


\[ 
f(x) = \frac{ a_n x^n + a_{n-1} x^{n-1} + \cdots + a_1 x + a_0 } {b_m x^m + b_{m-1} x^{m-1} + \cdots + b_1 x + b_0}
\]


\[  \text{ for all } x  \, \text{ in the domain. }   \]



\end{template}




The natural domain of a polynomial functions is all real numbers except the roots (zeros) of the polynomial in the denominator.  However, we can always state a restricted domain.








Although the definition quotes the standard form for a polynomial, the factored form is better suited for our analysis.

\[
p(x) = \frac{ a (x - r_1)^{e_1} (x - r_2)^{e_2} \cdots  (x - r_k)^{e_k} } { b (x - s_1)^{f_1} (x - s_2)^{f_2} \cdots  (x - s_h)^{f_h} }
\]








\begin{itemize}
\item The zeros (roots) of the polynomial in the numerator are the zeros of the rational function.
\item The zeros (roots) of the polynomial in the denominator are the singularities of the rational function.
\end{itemize}





















\subsection*{Exponential Functions}










\begin{template}  \textbf{\textcolor{blue!55!black}{Exponential Functions}} \\


Exponential functions are characterized by a constant \textit{percentage} growth rate. \\


We have several thoughts about exponential functions.

First, we have the basic expnential function, which has a formula like, $f(x) = r^x$ with $r > 0$. 
This give us a basic feeling for the behavior of exponential functions.



for example $f(x) = e^x$:


\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel={$y=f(x)$}, grid = major, grid style={dashed},
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            yticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$}, 
            xticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
            
      \addplot [line width=2, penColor, smooth,samples=200,domain=(-9:2),<->] {2.7^x};
      \addplot [line width=1, black, dashed,samples=200,domain=(-9:9),<->] {0};




  \end{axis}
\end{tikzpicture}
\end{image}















Secondly, we have four basic forms of the basic exponential funciton that illustrate the basic variations of the basic behavior.

\begin{itemize}
\item $f(x) = e^x$ 
\item $f(x) = e^{-x}$ 
\item $f(x) = -e^x$ 
\item $f(x) = -e^{-x}$ 
\end{itemize}






Finally, we include shifting and stretching, which algebraically is described by compositions with linear functions.


\[ f(x) = A \, e^{B \, x + C} + D = L_o \circ E \circ L_i\]

\[
\text{ where } \,  E(y) = e^y   \,  \text{ and } \,   L_o(t) = A \, t + D    \,  \text{ and } \,   L_i(v) = B \, v + C
\]



Strictly speaking, vertical shifts are not allowed for exponential functions.  We call them \textit{shifted exponential functions}.  However, nobody is going to complain if we just called them exponential functions.






\end{template}





The natural domain of an exponential function is all real numbers.  However, we can always state a restricted domain. \\









\subsection*{Logarithmic Functions}








\begin{template}  \textbf{\textcolor{blue!55!black}{Logarithmic Functions}} \\


Logarithmic function are the inverse functions of exponential functions. \\


We have several thoughts about logarithmic functions.

First, we have the basic logarithmic function, which has a formula like, $f(x) = \log_r(x)$ with $r > 0$. 
This give us a basic feeling for the behavior of exponential functions.



for example $f(x) = \ln(x)$:


\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel={$y=f(x)$}, grid = major, grid style={dashed},
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            yticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$}, 
            xticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
            
      \addplot [line width=2, penColor, smooth,samples=200,domain=(0.1:9),<->] {ln(x)};
      \addplot [line width=1, black, dashed,samples=200,domain=(-9:9),<->] ({0}, {x});




  \end{axis}
\end{tikzpicture}
\end{image}















Secondly, we have four basic forms of the basic logarithmic funciton that illustrate the basic variations of the basic behavior.

\begin{itemize}
\item $f(x) = \ln(x)$ 
\item $f(x) = \ln(-x)$ 
\item $f(x) = -\ln(x)$ 
\item $f(x) = -\ln(-x)$ 
\end{itemize}






Finally, we include shifting and stretching, which algebraically is described by compositions with linear functions.


\[ f(x) = A \, \ln(B \, x + C) + D = L_o \circ L \circ L_i\]

\[
\text{ where } \,   L(y) = \ln(y)   \,  \text{ and } \,    L_o(t) = A \, t + D    \,  \text{ and } \,   L_i(v) = B \, v + C
\]








\end{template}





















\subsection*{Absolute Value Functions}








\begin{template}  \textbf{\textcolor{blue!55!black}{Absolute Functions}} \\



Absolute value functions are usually identified by their vertical bars.





\[ f(x) = A | B \, x + C | + D = L_o \circ A \circ L_i\]

\[
\text{ where } \,  A(y) = | y |  \,  \text{ and } \,    L_o(t) = A \, t + D    \,  \text{ and } \,   L_i(v) = B \, v + C
\]







\end{template}






Unfortunately, our algebra doesn't work vrey well with the vertical bars.  Usually, our first algebraic thought is to get rid of the bars.  This is because the absolute value function is really a piecewise defined function.   \\



\textbf{Notice} that all of our forms are defined by composition with linear functions.  \\

When the ``inside'' or ``outside'' is not a linear function, then we do not have these basic elementary functions. (This is usually the case.)















































\subsection*{Operations}





Values of functions are numbers.  So, it is no surprise that we can make new functions by combining old functions with addition, subtraction, multiplication, and division. Thinking ahead to Calculus, it is also very handy to see multiplication by a constant.\\




\begin{warning}  \textbf{\textcolor{red!70!black}{Is}}


When categorizing a formula, our first thought is that the function is an elementary function and we would categorize it from our library. \\

If it is not, then we turn to the structure of the expression where the \textbf{Order of Operations} dictates how we view the algebra. \\

When thinking of the Order of Operations, we are thinking of what the expression actually looks like as it is written.


\begin{center}
These are ``\textbf{\textcolor{red!70!black}{Is}}'' questions.
\end{center}

We are not thinking of how the expression might be changed algebraically.  We are identifying how it is currently written. \\

The Order of Operations instructs us on how to make this decision.

\end{warning}



\subsection*{Constant Multiple}


\begin{template}  \textbf{\textcolor{blue!55!black}{Constant Multiple}} \\


If  $f$ is a function and $A$ is a constant, then $A \, f$ is called a \textbf{\textcolor{green!50!black}{constant multiple}} of $f$. \\

The domain of a constant multiple of a function is equal to the domain of the function. \\


\[ A \cdot f(x)  \]



\end{template}



\begin{observation}

If $f$ belongs to a nice category of functions, then the constant multiple also belongs to the same nice category.

\end{observation}




\begin{example}

$f(x) = 4 \, e^x$ is a constant multiple of $e^x$. \\
$Y(t) = -5 \, \sin(4t + 8)$ is a constant multiple of $\sin(4t + 8)$. \\
$m(k) = 4 (k+1)(k-5)$ is a constant multiple of $(k+1)(k-5)$. \\


$T(n) = 4 (n+1) + 6$ is not a constant multiple of $(n+1) + 6$. \\

\end{example}


















\subsection*{Sums}


\begin{template}  \textbf{\textcolor{blue!55!black}{Sum}} \\


If  $f$ and $g$ are functions, then $f + g$ is called the \textbf{\textcolor{green!50!black}{sum}} of $f$ and $g$. \\

The sum of two functions is defined on the intersection of their two domains. \\


\[ (f + g)(a) = f(a) + g(a)  \]



\end{template}



\begin{warning}

If $f$ or $g$ belongs to a nice category of functions, their sum usually does not.

\end{warning}




\begin{example}

$f(x) = e^x$ is an exponential function. \\
$g(t) = 3 t + 4$ is a linear function. \\

However, $(f + g)(k) = f(k) + g(k) = e^k + 3 k + 4$ is neither exponential nor linear.

\end{example}








\subsection*{Differences}



\begin{template}  \textbf{\textcolor{blue!55!black}{Difference}} \\


If  $f$ and $g$ are functions, then $f - g$ is called the \textbf{\textcolor{green!50!black}{difference}} of $f$ and $g$. \\

The difference of two functions is defined on the intersection of their two domains. \\


\[ (f - g)(a) = f(a) - g(a)  \]



\end{template}



\begin{warning}

If $f$ or $g$ belongs to a nice category of functions, their difference usually does not.

\end{warning}





\begin{example}

$f(x) = e^x$ is an exponential function. \\
$g(t) = 3 t + 4$ is a linear function. \\

However, $(f - g)(k) = f(k) - g(k) = e^k - 3 k - 4$ is neither exponential nor linear.

\end{example}
















\subsection*{Products}



\begin{template}  \textbf{\textcolor{blue!55!black}{Product}} \\


If  $f$ and $g$ are functions, then $f \cdot g$, or just $f g$, is called the \textbf{\textcolor{green!50!black}{product}} of $f$ and $g$. \\

The product of two functions is defined on the intersection of their two domains. \\


\[ (f \cdot g)(a) = f(a) \cdot g(a)  \]



\end{template}



\begin{warning}

If $f$ or $g$ belongs to a nice category of functions, their product usually does not.

\end{warning}





\begin{example}

$f(x) = e^x$ is an exponential function. \\
$g(t) = 3 t + 4$ is a linear function. \\

However, $(f \cdot g)(k) = f(k) \cdot g(k) = e^k (3 k + 4)$ is neither exponential nor linear.

\end{example}























\subsection*{Quotients}



It has been very common to see numbers involved in division: $15 \div 7$. However, functions are never written as division. They are always written as quotients (fractions).




\begin{template}  \textbf{\textcolor{blue!55!black}{Quotients}} \\


If  $f$ and $g$ are functions, then $\frac{f}{g}$ is called the \textbf{\textcolor{green!50!black}{quotient}} of $f$ and $g$. \\

The quotient of two functions is defined on the intersection of their two domains except at the zeros of the function in the denominator. \\


\[ \left(\frac{f}{g}\right)(a) = \frac{f(a)}{g(a)}  \]



\end{template}



\begin{warning}

If $f$ or $g$ belongs to a nice category of functions, their quotient usually does not.

\end{warning}





\begin{example}

$f(x) = e^x$ is an exponential function. \\
$g(t) = 3 t + 4$ is a linear function. \\

However, $\left(\frac{f}{g}\right)(k) = \frac{f(k)}{g(k)} = \frac{e^k}{(3 k + 4)}$ is neither exponential nor linear.

\end{example}











\begin{warning}


\begin{itemize}
\item Just because you see an exponential formula does not mean you have an exponential function. \\ 
\item Just because you see a logarithmic formula does not mean you have a logarithmic function. \\ 
\item Just because you see a quadratic formula does not mean you have a quadratic function. \\ 
\item Just because you see a linear formula does not mean you have a linear function. \\ 
\end{itemize}

\end{warning}







































\subsection*{Piecewise Defined Functions}













We combine numbers together with the operations addition, subtraction, multiplication, and division.  We can combine functions by applying these same operations on the function values.  \\

In this way, we can build new functions from old funcitons. \\

We can also build new functions by gluing together pieces of old functions.



\begin{template} \textbf{\textcolor{blue!55!black}{Piecewise Defined Functions}} \\


Suppose we have several functions with their domains.

\begin{itemize}
\item Function $F$ with domain $D_F$. 
\item Function $G$ with domain $D_G$. 
\item Function $H$ with domain $D_H$. 
\end{itemize}


We can create a new function, $N$, by gluing together pieces of $F$, $G$, and $H$.


First, select disjoint subsets of the three given domains.


\begin{itemize}
\item $S_F \subseteq D_F$. 
\item $S_G \subseteq D_G$. 
\item $S_H \subseteq D_H$. 
\end{itemize}


The domain of $N$ will be the union of these three subsets: $S_F \cup S_G \cup S_H$. \\

When evaluating $N(a)$, use the formula from $F$, $G$, or $H$ that goes with the domain subset holding $a$.


\end{template}

$N$ is called a \textbf{\textcolor{blue!55!black}{piecewise defined function}}. \\




\begin{notation}


The notation for a piecewise defined function must list the formulas to use along with the domain subset corresponding to formula.



\[
N(x) = 
\begin{cases}
  F & S_F     \\
  G & S_G \\
  H & S_H
\end{cases}
\]



The right column lists the subsets of the original domains.  \\
The left column lists the orignal formulas corresponding to the subdomains.

\end{notation}






\begin{example}

Define $N$ as follows.


\[
N(x) = 
\begin{cases}
  2x - 1   &    [-8, 0)     \\
  3 - x^2  &    (0, 2) \\
  \ln(x)    &    (5, 9]
\end{cases}
\]


\begin{explanation}

The domain of $N$ is the union of the given intervals: $[-8, 0) \cup (0,2) \cup (5, 9]$.


To evaluate $N(d)$, first identify to which interval $d$ belongs.  Then, use the corresponding formula to evaluate.







\end{explanation}




To evaluate $N(-4)$, first identify that $-4 \in [-8, 0)$.  That tells us to use the formula $2x - 1$. \\

\[  N(-4) = 2(-4) - 1 = -9  \]





To evaluate $N(1)$, first identify that $1 \in (0, 2)$.  That tells us to use the formula $3 - x^2$. \\

\[  N(1) = 3 - (1)^1 = 2  \]





To evaluate $N(7)$, first identify that $7 \in (5, 9]$.  That tells us to use the formula $\ln(x)$. \\

\[  N(-7) = \ln(7)  \]



\end{example}














\begin{example}

Define $N$ as follows.


\[
N(x) = 
\begin{cases}
  2x - 1   &    [-8, 0)     \\
  3 - x^2  &    (0, 2) \\
  \ln(x)    &    (5, 9]
\end{cases}
\]







\begin{explanation}


To evaluate $\lim\limits_{x \to 0^- N(x)}$, we first notice that we are only considering numbers in the interval $[-8, 0)$.


\[
\lim\limits_{x \to 0^-} N(x) = \lim\limits_{x \to 0^-} 2x - 1 = -1
\]



\end{explanation}









\begin{explanation}


To evaluate $\lim\limits_{x \to 0^+ N(x)}$, we first notice that we are only considering numbers in the interval $(0,2)$.


\[
\lim\limits_{x \to 0^+} N(x) = \lim\limits_{x \to 0^+} 3 - x^2 = 3
\]



\end{explanation}





\end{example}

























\begin{example}

Define $F$ as follows.


\[
F(t) = 
\begin{cases}
  e^x - x   &    [-10, -3)     \\
  6x + 1  &    (-1, 5) \\
  4 \ln(x+2)    &    (5, 9]
\end{cases}
\]




\begin{question}

\[
F(-5) = \answer{e^{-5} + 5}
\]


\end{question}




\begin{question}

\[
F(0) = \answer{1}
\]


\end{question}






\begin{question}

\[
F(7) = \answer{4 \ln(9)}
\]


\end{question}




\end{example}
















































\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Function Forms]{https://ximera.osu.edu/csccmathematics/precalculus/precalculus/composition/examples/exampleList}

\end{center}







\end{document}
