\documentclass{ximera}

\input{../../preamble.tex}

\title{Visual Behavior}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}
\maketitle






\begin{warning}


This section talks about deciphering function information from the graph of a function, \textbf{\textcolor{red!80!black}{which is not what we want}}.  Our goal is exactness, which can only be achieved through algebra and functional reasoning.\\



However, students have extensive experience with graphs, which makes graphs a good starting point. \\

Our plan is to use graphs to understand the ideas of function analysis and let that guide our algebraic analysis, \textbf{\textcolor{red!80!black}{which is what we want}}.

\end{warning}



Functions are packages containing three sets (one of which is a set of pairs), which satisfies one rule.  That is the struture of a function.  In Analysis, the sets are often measurements and the function is how we compare those measurements.

We would like to know things like:

\begin{itemize}
\item When one measurement increases, does the other also increase?
\item When one measurement increases, does the other decrease?
\item When one measurement changes quickly, does the other change quickly or slowly?
\item Which measurements are connected to maximums and minimums in the other measurements?
\item If one measurement is smooth, then is the other smooth or does it have sudden changes?
\end{itemize}


In this section, we will explore this type of \textbf{function behavior} via graphs. Once we gain a graphical familiarity with this type of behavior, we will extend our thinking to include formulas (Algebra).  Our eventual goal is to be exact with our analysis.  This will require both our graphical and algebraic tools working together.









\subsection{Learning Outcomes}

\begin{sectionOutcomes}
In this section, students will...

\begin{itemize}
\item describe end-behavior.
\item explore discontinuities.
\item locate intervals where functions are increasing and decreasing.
\end{itemize}
\end{sectionOutcomes}



















\begin{center}
\textbf{\textcolor{green!50!black}{ooooo=-=-=-=-=-=-=-=-=-=-=-=-=ooOoo=-=-=-=-=-=-=-=-=-=-=-=-=ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Visual Behavior]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/visualBehavior/examples/exampleList}

\end{center}






\end{document}
