\documentclass{ximera}

\input{../../preamble.tex}

\title{Functions}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}
\maketitle





\section*{Anatomy of a Function}


A function is a special type of structure called a \textbf{\textcolor{blue!55!black}{relation}}.  \\



As such, a function is a package containing three sets. 


\begin{itemize}
\item One set is called the \textbf{\textcolor{purple!85!blue}{domain}}. 
\item One set is called the \textbf{\textcolor{purple!85!blue}{codomain}}.  
\item Finally, there is a third set of pairings.  Each pairing associates a member of the domain with a member of the codomain. This third set does not seem to have an official title.
\end{itemize}


While a relation is just two sets of items and some pairings between the two sets, functions satisfy one rule: \\





\begin{condition} \textbf{\textcolor{purple!85!blue}{THE Rule}}  \\

\begin{itemize}
\item Functions are those relations where each domain item is paired with \underline{exactly} one codomain item.
\end{itemize}
\end{condition}


\begin{center}
\textbf{OR}
\end{center}


\begin{condition} \textbf{\textcolor{purple!85!blue}{THE Rule}} \\

\begin{itemize}
\item Functions are those relations where each domain item occurs in \underline{exactly} one pair.
\end{itemize}
\end{condition}




This simple rule makes a world of difference. But, we can do a little better for this course. \\




This course (and Calculus) is a study of measurements and how they compare to each other, which means it is a study of the real numbers.  Therefore, the domains and ranges of our functions will be subsets of the real numbers.  We say that our functions \textbf{map} real numbers to real numbers.


\[
\text{ f : } \mathbb{R} \mapsto \mathbb{R}
\]





With the addition of this one rule and the restriction of our domains and codomains to be sets of real numbers, we are ready to begin our journey towards Calculus.








\subsection*{Learning Outcomes}


\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item use function notation.
\item evaluate functions.
\item solve equations involving functions.
\item analyze functions.
\item focus on numeric functions.
\end{itemize}
\end{sectionOutcomes}









\pdfOnly{
\begin{center}
more examples can be found in the online edition.
\end{center}
}




\begin{onlineOnly}
\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Functions]{https://ximera.osu.edu/csccmathematics/precalculus/precalculus/functions/examples/exampleList}

\end{center}
\end{onlineOnly}







\end{document}
