\documentclass{ximera}

\input{../../preamble.tex}

\title{Composition}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}
\maketitle













Composition is an operation on functions.  It takes two functions and produces a third function.


\[  NewFunction(d) = (Outside \circ Inside)(d) = Outside(Inside(d))    \]



We have explored this idea with linear functions.  Now, we would like to explore the composition of any two functions.






With just a single function, we normally think of running along the real line from left to right connecting these domain numbers with their function values.  It is a very orderly process.  Composition follows the same idea, except the domain of the outside function is now the range of the inside function.

Now the $Inside$ will be providing the domain numbers to the $Outside$ function and it may not be as orderly. The values of the $Inside$ function may increase and decrease, which, in turn, means the domain numbers for $Outside$ may get repeated or run backwards.   The $Inside$ function takes us on a whacky ride through the domain of $Outside$, possibly repeating domain numbers and thus function values as well.


In this section, we will map the whole process.










\subsection{Learning Outcomes}





\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item compose functions.
\end{itemize}
\end{sectionOutcomes}














\begin{center}
\textbf{\textcolor{green!50!black}{ooooo=-=-=-=-=-=-=-=-=-=-=-=-=ooOoo=-=-=-=-=-=-=-=-=-=-=-=-=ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Composition]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/composition/examples/exampleList}

\end{center}







\end{document}
