\documentclass{ximera}

\input{../../preamble.tex}

\title{Tangent Lines}

\begin{document}

\begin{abstract}
zooming in
\end{abstract}
\maketitle


 
 


\subsection*{Tangent Lines}

The story about tangent lines we told involved parabolas, but it extends to any curve.






Suppose we have the cubic function $W(t) = \frac{1}{10} (3x - 2)(x+5)(x+3)$. 

This graph contains the point $(0,-3)$.


The line $y=\frac{29}{10}x-3$  goes through the point.

The curve and line have a special relationship, which we can see by zooming in around the point.






\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{j3rltvwd0l}{400}{300}
\end{center}





\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{el01gfmoiv}{400}{300}
\end{center}





\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{sbfbfh7oqx}{400}{300}
\end{center}






\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{izri7rfpqf}{400}{300}
\end{center}






As you zoom in, the graph slowly looks more and more like the line. 


The line does the best job of approximating the graph at the point $(0,-3)$. 


And, this is the only line that will work for the point $(0,-3)$.  Any other line, besides this one, will have a permanant nonzero angle between the graph and the line.  

The graph and this particular line have the same ``slope'' at the point. 


This line is called the \textbf{\textcolor{blue!55!black}{tangent line}} for $W(t) = \frac{1}{10} (3x - 2)(x+5)(x+3)$ at the point $(0,-3)$.


Each point on the graph of the function has its own unique tangent line with its own slope.




\begin{definition} (from earlier) \textbf{\textcolor{green!50!black}{Tangent Line}}


Suppose $C$ is a curve, like the graph of a function. 

Let $(a, b)$ be a point on the curve.

The line tangent to the curve, $C$, at the point $(a, b)$ is the line that goes through the point $(a, b)$ and does the best job of approximating the curve at the point $(a, b)$.

This line is called a \textbf{tangent line} or a \textbf{tangent} at the point $(a, b)$.




\end{definition}


Since ``slope'' is a characteristic of a line and a curve is not a line, curves do not have a slope as we know that word. 

However, it is easy to see that curves have something that seems a lot like a slope at points. We will adopt the slope of the tangent as the slope of the curve at a point. 

As the cubic curve above demonstrates, curves might have different slopes at different points. 

We want to quantize this idea of a slope for a curve at a point. 

If we can quantize it, then we can build a function for it and use that to measure a rate of change for our curve. This is the first step toward Calculus. It will take us a while to quantize this idea of slope for a curve in a logical tool. A tool we will call the \textbf{derivative}.

The way Calculus gets to this tangent line is through secant lines on either side of the tangent point.








\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{vilqspjxsf}{400}{300}
\end{center}

































































\subsection*{Intervals and Secants}



Suppose we have a quadratic function $Q(x) = a (x - h)^2 + k$, with $a > 0$. 


Then its graph is a parabola.







\begin{image}
\begin{tikzpicture}
     \begin{axis}[
                domain=-5:10, ymax=10, xmax=10, ymin=-5, xmin=-5,
                axis lines =center, xlabel=$x$, ylabel=$y$,
                ytick={-4,-2,2,4,6,8,10},
                xtick={-4,-2,2,4,6,8,10},
                ticklabel style={font=\scriptsize},
                every axis y label/.style={at=(current axis.above origin),anchor=south},
                every axis x label/.style={at=(current axis.right of origin),anchor=west},
                axis on top,
                ]


        \addplot [draw=penColor, very thick, smooth, domain=(-1:7),<->] {0.5*(x-3)^2 + 2};
        \addplot [line width=1, gray, dashed,samples=100,domain=(-9.5:9.5)] ({3},{x});
        


        \addplot [color=penColor,only marks,mark=*] coordinates{(3,2)};
        \node[penColor] at (axis cs:4,1.5) {$(h, k)$};
        %\node[penColor] at (axis cs:5,-9) {$-0.5 x^2 - 5 x + 15.5$};



    \end{axis}
\end{tikzpicture}
\end{image}
We can examine the rate of change over any interval.


Let's examine the interval $[2, 5]$.


$\blacktriangleright$ \textbf{\textcolor{blue!55!black}{Algebraically}}, the rate of change of $Q(x)$ over the interval $[2,5]$ is given by 

\[
\frac{Q(5) - Q(2)}{5 - 2} 
\]



$\blacktriangleright$ \textbf{\textcolor{blue!55!black}{Geometrically}}, this is the slope of the \textbf{secant} line running through the points $(2, Q(2))$ and $(5, Q(5))$.














\begin{image}
\begin{tikzpicture}
     \begin{axis}[
                domain=-5:10, ymax=10, xmax=10, ymin=-5, xmin=-5,
                axis lines =center, xlabel=$x$, ylabel=$y$,
                ytick={-4,-2,2,4,6,8,10},
                xtick={-4,-2,2,4,6,8,10},
                ticklabel style={font=\scriptsize},
                every axis y label/.style={at=(current axis.above origin),anchor=south},
                every axis x label/.style={at=(current axis.right of origin),anchor=west},
                axis on top,
                ]


        \addplot [draw=penColor, very thick, smooth, domain=(-1:7),<->] {0.5*(x-3)^2 + 2};
        \addplot [line width=1, gray, dashed,samples=100,domain=(-9.5:9.5)] ({3},{x});
        

        \addplot [color=penColor2,only marks,mark=*] coordinates{(5,4)};
        \addplot [color=penColor2,only marks,mark=*] coordinates{(2,2.5)};
        
        \addplot [draw=penColor2, very thick, smooth, domain=(-2:9),<->] {0.5*(x-5) + 4};

        \addplot [color=penColor,only marks,mark=*] coordinates{(3,2)};
        \node[penColor] at (axis cs:4,1.5) {$(h, k)$};
        %\node[penColor] at (axis cs:5,-9) {$-0.5 x^2 - 5 x + 15.5$};



    \end{axis}
\end{tikzpicture}
\end{image}

Intervals and secants are algebraic and geometric partners. Secants give a picture of rates of change over intervals.



What if we push the secant over a little bit until it becomes a tangent line?







\begin{image}
\begin{tikzpicture}
     \begin{axis}[
                domain=-5:10, ymax=10, xmax=10, ymin=-5, xmin=-5,
                axis lines =center, xlabel=$x$, ylabel=$y$,
                ytick={-4,-2,2,4,6,8,10},
                xtick={-4,-2,2,4,6,8,10},
                ticklabel style={font=\scriptsize},
                every axis y label/.style={at=(current axis.above origin),anchor=south},
                every axis x label/.style={at=(current axis.right of origin),anchor=west},
                axis on top,
                ]


        \addplot [draw=penColor, very thick, smooth, domain=(-1:7),<->] {0.5*(x-3)^2 + 2};
        \addplot [line width=1, gray, dashed,samples=100,domain=(-9.5:9.5)] ({3},{x});
        

        \addplot [color=penColor2,only marks,mark=*] coordinates{(3.5,2.125)};
        
        \addplot [draw=penColor2, very thick, smooth, domain=(-2:9),<->] {0.5*(x-3.5) + 2.125};

        %\addplot [color=penColor,only marks,mark=*] coordinates{(3,2)};
        %\node[penColor] at (axis cs:4,1.5) {$(h, k)$};
        %\node[penColor] at (axis cs:5,-9) {$-0.5 x^2 - 5 x + 15.5$};



    \end{axis}
\end{tikzpicture}
\end{image}

Now it is a picture of the rate of change over the interval $\left[ \tfrac{7}{2}, \tfrac{7}{2} \right]$. 


How should we interpret this? 


What is the rate of change \textbf{\textcolor{red!90!darkgray}{AT}} a point? 






\begin{idea}

You cannot calculate the rate of change over an interval with $0$ length.  

Therefore, we are inventing an interpretation for the rate of change at a point. 

We are calling our invention the \textit{instantaneous rate of change}.


\end{idea}











\subsection*{Tangent Lines}

Suppose $f$ is a function.  The graph of $f$ is the collection of points whose coordinates are pairs in $f$.  That is, their coordinates look like $(d, f(d))$.

Suppose $a$ and $b$ are two distinct domain numbers of $f$.  Then the line through $(a, f(a))$ and $(b, f(b))$ is called a secant line.  The slope of this secant line equals the rate of change of $f$ over the interval $[a, b]$.

Tangent lines are degenerate secants. Secant lines need two points on the graph of a function.  A tangent line is a secant line where the two points are the same point. Tangent lines are secant lines created from an interval of length $0$.  But tangent lines are still lines.  They still have a slope.


\textbf{\textcolor{red!90!darkgray}{$\blacktriangleright$}} If the slope of a secant line corresponds to the function's rate of change over an interval, then the slope of a tangent line corresponds to the function's rate of change at a single number.


We are calling this slope of the tangent line the \textbf{\textcolor{purple!85!blue}{instantaneous rate of change}} at the domain number corresponding to the tangent point.




\textbf{\textcolor{green!50!black}{Remember Definition: Instantaneous Rate of Change}}  


Let $f$ be a function. Let $a$ be a number in the domain of $f$.

If the graph of $y = f(x)$ has a non-vertical tangent line at the point $(a, f(a))$, then the slope of this tangent line is the \textbf{instantaneous rate of change} of $f$ \textbf{at} a.

















\subsection*{Quadratics}

We'll begin our investigation into instantaneous rate of change with quadratic functions, whose graphs are parabolas (which never have vertical tangent lines). \\


We need a method of obtaining the slope of tangent lines to parabolas.







$\blacktriangleright$ \textbf{Basic Quadratic}


We'll begin with our basic quadratic function: \textbf{\textcolor{purple!85!blue}{$Q(x) = x^2$}} and its graph \textbf{\textcolor{purple!85!blue}{$y = x^2$}}. \\


We are investigating the graph of $y = x^2$, which is a parabola whose vertex is $(0, 0)$. \\

Let's select a domain number of $Q$, call it $x_0$.   This corresponds to the point $(x_0, y_0) = (x_0, (x_0)^2)$ on the graph of $Q$.

Let's add a picture of the tangent line at $(x_0, y_0) = (x_0, (x_0)^2)$ to our graph of $Q$.


\begin{image}
\begin{tikzpicture}
     \begin{axis}[
                domain=-5:5, ymax=10, xmax=5, ymin=-5, xmin=-5,
                axis lines =center, xlabel=$x$, ylabel=$y$,
                ytick={-4,-2,2,4,6,8,10},
                xtick={-4,-2,2,4,6,8,10},
                ticklabel style={font=\scriptsize},
                every axis y label/.style={at=(current axis.above origin),anchor=south},
                every axis x label/.style={at=(current axis.right of origin),anchor=west},
                axis on top,
                ]


        \addplot [draw=penColor, very thick, smooth, domain=(-3:3),<->] {x^2};
        %\addplot [line width=1, gray, dashed,samples=100,domain=(-5:9.5)] ({0},{x});
        

        \addplot [color=penColor2,only marks,mark=*] coordinates{(2,4)};
        \node[penColor2] at (axis cs:3,4) {$(x_0, y_0)$};
        \addplot [draw=penColor2, very thick, smooth, domain=(0.5:3),<->] {4*x - 4};
        


        %\addplot [color=penColor,only marks,mark=*] coordinates{(3,2)};
        %\node[penColor] at (axis cs:2,1.5) {$(h, k)$};




    \end{axis}
\end{tikzpicture}
\end{image}

The tangent line is the graph of a linear function. Let's call this linear function $T$.


$T$ is a linear function. Therefore, the formula for $T$, would look like $T(x) = m(x - x_0) + y_0$, for some $m$.  According to our definition, $m$ is the instantaneous rate of change of $Q$ at $x_0$.  How do we determine its exact value?



\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ We would like a way to obtain the exact value of $m$.}} \\


To get the exact value of $m$, we are going to create a new quadratic function, which will be the difference between the original quadratic function and the linear function.



\begin{explanation}


\[
Q_{new}(x) = Q(x) - T(x) 
\]

\[
Q_{new}(x) = Q(x) - T(x) = x^2 - (m(x - x_0) + y_0)
\]




$Q_{new}(x)$ is a quadratic function and $Q_{new}(x_0) = Q(x_0) - T(x_0) = y_0 - y_0 = 0$. \\


We can also see from the graph above that $Q(x) \geq T(x)$, which means $Q_{new}(x) \geq 0$\\


So, the graph of $Q_{new}(x)$ is a parabola with vertex, $(x_0, 0)$, on the $x$-axis.  The graph of $y = Q(x)$ has slid to the left a distance of $x_0$.  \\


Therefore, the vertex form of $Q_{new}(x)$ is $Q_{new}(x) = (x - x_0)^2 = (x - x_0)(x - x_0)$  \\






\[
Q_{new}(x) = Q(x) - T(x) = x^2 - (m \, (x - x_0) + y_0) 
\]



On the other hand,


\[
Q_{new}(x) = (x - x_0)^2
\]



Therefore, we have

\[
x^2 - (m \, (x - x_0) + y_0) =  (x - x_0)^2
\]


or


\[
x^2 - (m \, (x - x_0) + y_0) =  x^2 - 2 x_0 \, x + x_0^2
\]


or


\[
 -m \, x + m \, x_0 - y_0 =   - 2 \, x_0 \, x + x_0^2
\]


For this to happen, we must have $-m \, x = - 2 \, x_0 \, x$ or \textbf{$m = 2 \, x_0$}. \\


\end{explanation}


At any point $(x_0, y_0)$ on the graph of $y = x^2$, the slope of the tangent line is $2 \, x_0$.  The slope of the tangent line is always twice the $x$-coordinate of the tangent point.




\begin{conclusion} \textbf{\textcolor{green!50!black}{Slope of a Tangent Line}} 


The graph of $y = x^2$ is a parabola.

Let $(x_0, y_0)$ be a point on this parabola.

Then the slope of the tangent line to the parabola at $(x_0, y_0)$ is given by 



\[ \text{slope } \, = \, 2 \, x_0  \]


\end{conclusion}
This for any point on the parabola. \\

What about other parabolas, besides the basic parabola? \\

All parabolas can be obtained from the basic parabola by stretching and shifting. \\



























\subsection*{Stretching}



We can vertically stretch our basic quadratic parabola by multiplying all of the $y$-coordinates by a nonzero constant, $a \ne 0$. \\


The new parabola has an equation of the form $y = a \, x^2$. \\

These are graphs of quadratic functions of the form $Q(x) = a \, x^2$. \\


Let's go through the same process. \\


\begin{explanation}

The tangent line at $(x_0, y_0)$ is the graph of a linear function. Let's call this linear function $T$.


$T$ is a linear function. Therefore, the formula for $T$, would look like $T(x) = m(x - x_0) + y_0$, for some $m$.  According to our definition, $m$ is the instantaneous rate of change of $Q$ at $x_0$.  How do we determine its exact value?


$\blacktriangleright$ We would like a way to obtain the exact value of $m$. \\



To get the exact value of $m$, we are going to create a new quadratic function.



\[
Q_{new}(x) = Q(x) - T(x) = a \, x^2 - (m \, (x - x_0) + y_0)
\]



$Q_{new}(x)$ is a quadratic function and $Q_{new}(x_0) = Q(x_0) - T(x_0) = y_0 - y_0 = 0$. \\

We can also see from the graph that $Q(x) > T(x)$ when $x \ne x_0$, which means $Q_{new}(x) \geq 0$.\\


So, its graph is a parabola with vertex, $(x_0, 0)$, on the $x$-axis.  \\


Therefore, the vertex form of $Q_{new}(x)$ is $Q_{new}(x) = a \, (x - x_0)^2 = (x - x_0)(x - x_0)$  \\






\[
Q_{new}(x) = Q(x) - T(x) = a \, x^2 - (m \, (x - x_0) + y_0) = a \, (x - x_0)^2
\]




\[
a \, x^2 - (m \, (x - x_0) + y_0) = a \, (x - x_0)^2 = a \, x^2 - 2 \, a \, x_0 x + a \, x_0^2
\]



\[
 -m \, x + m \, x_0 - y_0 =   - 2 \, a \, x_0 \, x + a \, x_0^2
\]


For this to happen, we must have $-m \, x = - 2 \, a \, x_0 \, x$ or \textbf{$m = 2 \, a \, x_0$}. \\


\end{explanation}

At any point $(x_0, y_0)$ on the graph of $y = a \, x^2$, the slope of the tangent line is $2 \, a \, x_0$.  The slope of the tangent line is always $2a$ times the $x$-coordinate of the tangent point.










\begin{conclusion} \textbf{\textcolor{green!50!black}{Slope of a Tangent Line}} 


The graph of $y = a \, x^2$ is a parabola.

Let $(x_0, y_0)$ be a point on this parabola.

Then the slope of the tangent line to the parabola at $(x_0, y_0)$ is given by 



\[ \text{slope } \, = \, 2 \, a \, x_0  \]


\end{conclusion}
This for any point on the parabola.




















\subsection*{Shifting}




Beginning with $Q(x) = a \, x^2$ and its parabola $y = a \, x^2$, we can shift horizontally and vertically. \\





Shifting keeps the shape of the parabola, $y = a \, x^2$, intact. \\


The new shifted parabola has an equation of the form $y = a \, (x - h)^2 + k$.

These are graphs of quadratic functions of the form $Q_{new}(x) = a \, (x - h)^2 + k$. \\


If we pick any point, $(x_0, y_0)$, on this parabola, then the slope of the tangent line is the same as the slope of the tangent line for $y = a \, x^2$ at the point $(x_0 - h, y_0 - k)$, because the actual shape of the parabola is not changed by shifting.  \\

From above, we know this slope is \textbf{$2 \, a \, (x_0 - h)$}.






\begin{conclusion} \textbf{\textcolor{green!50!black}{Slope of a Tangent Line}} 


The graph of $y = a \, (x - h)^2 + k$ is a parabola.

Let $(x_0, y_0)$ be any point on this parabola.

Then the slope of the tangent line to the parabola at $(x_0, y_0)$ is given by 



\[ \text{slope } \, = \, 2 \, a \, (x_0 - h)  \]


\end{conclusion}


Every parabola can be described by an equation of the form $y = a \, (x - h)^2 + k$. \\

Our slope of tangent line idea can be applied to every parabola. \\

We have covered every parabola. \\














\begin{theorem} \textbf{\textcolor{green!50!black}{Slope of a Tangent Line (Parabola)}} 


The graph of $y = a (x - h)^2 + k$ is a parabola.

Let $(x_0, y_0)$ be any point on this parabola.

Then the slope of the tangent line to the parabola at $(x_0, y_0)$ is given by 



\[ 2 \, a \, (x_0 - h) \]


\end{theorem}












\textbf{Note:} If we look at the vertex of the parabola $(x_0, y_0) = (h, k)$, then $2 a (x_0 - h) = 2 a (h - h) = 0$ and the slope of the tangent line is $0$, just like we had reasoned before. \\











\section*{iRoC}

Let $Q(x) = a (x - h)^2 + k$ be any quadratic function.

Every point, $(x_0, y_0)$, on the graph of $y = Q(x)$ has a tangent line.

Each tangent line has a slope, which we are calling the instantaneous rate of change of $Q$ at $x_0$.


\textbf{\textcolor{red!90!darkgray}{$\blacktriangleright$}} We can create a new function from this.



\begin{definition} \textbf{\textcolor{green!50!black}{iRoC}}  


Given a quadratic function, $Q(x) = a (x - h)^2 + k$, we define \textbf{the instantaneous rate of change of Q} to be the slope of the tangent line on the graph of $y = Q(x)$ at the point $(x, y)$.


The function $iRoC_Q(x)$ is defined to equal the slope of the tangent line at the point $(x, y)$.


$iRoC_Q(x)$ is a linear function given by 

\[  iRoC_Q(x) = 2 a (x-h) \]

\end{definition}


We now have a function, with a formula, for measuring the instantaneous rate of change of any quadratic function at any domain number. \\


\textbf{\textcolor{red!90!darkgray}{$\blacktriangleright$}} The instantaneous rate of change of the quadratic function, $Q(x) = a \, (x - h)^2 + k$, is the linear function, $iRoC_Q(x) = 2 \, a \, (x-h)$. \\




We can add this to our growing collection. \\ 


\textbf{\textcolor{red!90!darkgray}{$\blacktriangleright$}} The instantaneous rate of change of the linear function, $L(x) = m \, x + b$, is the constant function, $iRoC_L(x) = m$. \\



\textbf{\textcolor{red!90!darkgray}{$\blacktriangleright$}} The instantaneous rate of change of the constant function, $C(x) = c$, is the zero function, $iRoC_C(x) = 0$. \\





\begin{notation} \textbf{\textcolor{blue!55!black}{Language}} 


We are just beginning our investigation of function behavior, so we have given our new function a very descriptive name, \textbf{\textcolor{blue!55!black}{Instantaneous Rate of Change}},  \textbf{\textcolor{blue!55!black}{$iRoC$}}. \\


Calculus uses the name \textbf{\textcolor{blue!55!black}{Derivative}} and it uses a little prime sign as the notation.



\[
iRoC_f(x) = f'(x)
\]


People pronounce $f'(x)$ as ``f prime of $x$''.


\end{notation}

We will use both names.  This will keep the idea fresh in our heads and also give us experience with Calculus notation.






\begin{notation} \textbf{\textcolor{red!70!black}{A Peek Ahead}}


We are beginning our investigation of function analysis at the beginning.  Our formulas have one variable.  But, later, in Calculus, it may not be so clear on what the variable is or what the function is. \\

For this reason, Calculus introduces more notation that is very explicit on this matter.



\[
iRoC_f(x) = f'(x) = \frac{df}{dx}
\]


People pronounce $\frac{df}{dx}$ as ``dfdx''.  They just say the letters.


\end{notation}






















\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of the Derivative]{https://ximera.osu.edu/csccmathematics/precalculus/precalculus/theDerivative/examples/exampleList}

\end{center}





\end{document}




