\documentclass{ximera}

\input{../../preamble.tex}

\title{the Derivative}

\begin{document}

\begin{abstract}
%
\end{abstract}
\maketitle




Calculus will provide a procedure for deriving the derivative of any function.  With the derivative, we will be able to, algebraically, calculate where a function is increasing and decreasing.  This will give us methods for tracking down local maximums and minimums.


Currently we know how to get formulas for the derivatives of a linear function and a quadratic function.  That's it! 

This isn't much, but it gives us enough experience to prepare for Calculus.


In this course, we will add the derivatives of sine and cosine, which will give us a bit more experience heading to Calculus.


Calculus will give us a \textbf{\textcolor{blue!55!black}{procedure}} to obtain formulas for derivatives ourselves.  In this course, we can just be supplied the derivative and use it to help our analysis of functions right now.


Thst's what we'll do in this section.






\subsection*{Learning Outcomes}


\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item analyze functions using the derivative.
\end{itemize}
\end{sectionOutcomes}













\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of the Derivative]{https://ximera.osu.edu/csccmathematics/precalculus/precalculus/theDerivative/examples/exampleList}

\end{center}







\end{document}
