\documentclass{ximera}

\input{../../preamble.tex}

\title{More Function Analysis}

\begin{document}

\begin{abstract}
%
\end{abstract}
\maketitle






What do we want when analyzing a function?



We would like to analyze all of their characteristics and features. 



\begin{itemize}
     \item \textbf{\textcolor{red!70!black}{Domain}} 
     \item \textbf{\textcolor{red!70!black}{Zeros}} 
     \item \textbf{\textcolor{red!70!black}{Continuity}} \\
     \text{    }$\circ$ \textbf{\textcolor{purple!85!blue}{discontinuities}}\\
     \text{    }$\circ$ \textbf{\textcolor{purple!85!blue}{singularities}}\\
     \item \textbf{\textcolor{red!70!black}{End-Behavior}} 
     \item \textbf{\textcolor{red!70!black}{Behavior}} \\
     \text{    }$\circ$ \textbf{\textcolor{purple!85!blue}{intervals where increasing}}\\
     \text{    }$\circ$ \textbf{\textcolor{purple!85!blue}{intervals where decreasing}}\\
     \item \textbf{\textcolor{red!70!black}{Global Maximum and Minimum}} 
     \item \textbf{\textcolor{red!70!black}{Local Maximums and Minimums}} 
     \item \textbf{\textcolor{red!70!black}{Range}} 
     \item \textbf{\textcolor{blue!55!black}{...and we would like a nice graph}} 
\end{itemize}




$\blacktriangleright$ Secondly, we would like a nice graph, including corresponding important points and auxilary graphing items.

\begin{itemize}
	\item intercepts
	\item endpoints and arrows
	\item vertical asymptotes
	\item horizontal asymptotes
\end{itemize}




We often turn to technology for a nice graph and we must remember that it is plotting individual points and may not understand the larger implications. As humans, we possess more information from our algebraic description and can enhance the graph with exaggerations and auxilary graphing items to help the reader understand the function better.





\textbf{\textcolor{red!70!black}{Remember: }}  We are not using our graph as reasoning.  We are not drawing conclusions and pointing to the graph and saying "because of the graph".  We are using the graph to help our algebraic and functional reasoning.


\subsection*{Learning Outcomes}


\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item analyze functions from their formula.
\end{itemize}
\end{sectionOutcomes}











\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of More Function Analysis]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/moreFunctionAnalysis/examples/exampleList}

\end{center}

\end{document}
