\documentclass{ximera}

\input{../../preamble.tex}

\title{Core Functions}

\begin{document}

\begin{abstract}
parents
\end{abstract}
\maketitle




In this course, we will begin studying a collection of functions called the \textbf{Elementary Functions}.

The \textbf{Elementary Functions} are created from an initial collection of ``building block'' functions, which we will call the \textbf{Core Functions}.

The Elementary Functions are created by combining the Core Functions together through addition, subtraction, multiplication, division, composition, and inverses.


The Elementary Functions are sums, differences, products, quotients, compositions, and inverses of Core Functions.

Calculus is largly about analyzing the Elementary Functions.






We will begin with seven Core Functions

\begin{itemize}
\item{Constant Functions: $f(x) = c$}
\item{The Identity Function: $f(x) = x$}
\item{Power Functions: $f(x) = x^n$}
\item{Root Functions: $f(x) = \sqrt[n]{x}$}
\item{The Exponential Function: $f(x) = e^x$}
\item{The Natural Logarithm Function: $f(x) = \ln(x)$}
\item{Sine and Cosine: $f(x) = \sin(x)$ and $f(x) = \cos(x)$}
\item{The Absolute Value Function: $f(x) = |x|$}
\end{itemize}


By forming sums, differences, products, quotients, compositions, and inverses of these Core Functions, we obtain the Elementary Functions.


\begin{itemize}
\item{Constant}
\item{Linear}
\item{Quadratic}
\item{Polynomial}
\item{Rational}
\item{Radical/Root}
\item{Exponential}
\item{Shifted Exponential}
\item{Logarithmic}
\item{Absolute Value}
\item{Trigonometric}
\item{Inverse Trigonometric}
\end{itemize}


Creating the library of Elementray Functions is a long journey. We'll need a lot of experiencing, which will take us right through Calculus.







\subsection*{The Core Functions}




\begin{formula} \textbf{\textcolor{blue!55!black}{Constant Functions}} 

A constant function is a function where all domain numbers are paired with the exact same range number. A constant function only has one value. 


\[   Constant(x) = C      \]

where $C$ is a fixed ``constant'' real number.

\end{formula}


\begin{example}

$\blacktriangleright H(t) = 3$

$\blacktriangleright H(t) = \sqrt{2}$

$\blacktriangleright H(t) = \pi$

$\blacktriangleright H(t) = \frac{3}{4}$

\end{example}

There is a core constant function for every real number.













\begin{formula} \textbf{\textcolor{blue!55!black}{The Identity Function}} 

The is only one Identity function. The Identity function pairs every domain number with itself. The value of the Idenitty function is just the domain value where it is evaluated. 


\[   Id(x) = x      \]



\end{formula}


\begin{example}

$\blacktriangleright Id(3) = 3$

$\blacktriangleright Id(\sqrt{2}) = \sqrt{2}$

$\blacktriangleright Id(\pi) = \pi$

$\blacktriangleright Id \left( \frac{3}{4} \right) = \frac{3}{4}$

\end{example}
















\begin{formula} \textbf{\textcolor{blue!55!black}{Power Functions}} 

Power functions raise the domain number to a power.


\[   Power(x) = x^r      \]

where $r$ is a fixed non-zero real number.


\end{formula}

\textbf{Note:} When $r$ is not an integer, then need to pay attention to the domain.

\begin{example}

$\blacktriangleright Power(x) = x^3 $

$\blacktriangleright Power(x) = x^{-2} $

$\blacktriangleright Power(x) = x^{\sqrt{2}} $

$\blacktriangleright Power(x) = x^{\frac{2}{3}} $

\end{example}

There is a core power function for every real number.

\textbf{Note:} The power is non-zero, because $x^0 = 1$.  That would just give us a constant function.









\begin{formula} \textbf{\textcolor{blue!55!black}{Root Functions}} 

The value of a Root (Radical) function is a root of the domain number.


\[   Root(x) = \sqrt[n]{x}      \]

where $n$ is a fixed positive real number.

\end{formula}

These are often called ``$n^{th}$'' Root Functions.

\textbf{Note:} When $n$ is not a positive integer, then need to pay attention to the domain.



\begin{example}

$\blacktriangleright Root(x) = \sqrt[2]{x} = \sqrt{x}  $

$\blacktriangleright Root(x) = \sqrt[3]{x}  $

$\blacktriangleright Root(x) = \sqrt[\tfrac{3}{4}]{x}  $

$\blacktriangleright Root(x) = \sqrt[\pi]{x}  $

\end{example}

There is a core root or radical function for every real number.

\textbf{Note:} There is a connecting bridge between power functions and root functions, which simplifies the algebra.








By now, you probably believe that the number $\pi$ is an important number in mathematics.  However, there is even a more importnat number, called $e$.

We will see the definition of the number $e$ later, but a good decimal approaximation is $2.718281828$.

We will see that every exponential function can be described in terms of $e$, so $e^x$ is our core exponential function.


\begin{formula} \textbf{\textcolor{blue!55!black}{The Exponential Function}} 




\[   e^x     \]



\end{formula}



Composition is a new operation for functions. By composing two functions, a new function is created.

Sometimes it is possible for the composition of two functions to be the Identity function, which is the identify element for composition. 

This will be important as we build our library of Elementary Functions.

When the composition of two functions turns out to be the Identity function, then we say that the two component functions are ``inverses'' of each other.

The inverse of the Exponential Function is the Natural Logarithm function.









\begin{formula} \textbf{\textcolor{blue!55!black}{The Natural Logarithm Function}} 




\[   \ln(x)    \]



\end{formula}


When we get to exponential and logarithmic functions, our magic sentence will be 

\begin{center}
$\ln(a)$ is the thing you raise $e$ to, to get $a$.
\end{center}





There are plenty of relationships in our world that exhibit a periodic pattern.

Our Core functions for periodic relationships will be Sine and Cosine.







\begin{formula} \textbf{\textcolor{blue!55!black}{Sine and Cosine}} 


\[   \sin(x)    \]

\[   \cos(x)    \]


\end{formula}





Our last Core function is how we algebraically talk about distance.  The Absolute Value function returns the distance between $0$ and the domain number.









\begin{formula} \textbf{\textcolor{blue!55!black}{The Absolute Value Function}} 


\[
|x| = 
\begin{cases}
  -x & \text{if $x<0$,}  \\
  x & \text{if $x\ge 0$}
\end{cases}
\]



\end{formula}


We can describe the distance bewteen any two number on the number line using the absolute value function.

Suppose $a$ and $b$ are two real numbers on the real number line.





\[
|a-b| = 
\begin{cases}
  -(a-b) & \text{if $a<b$,}  \\
  a-b & \text{if $b\ge a$}
\end{cases}
\]



$|a-b| = |b-a|$

Both are positive or $0$.

Either describes the distance between $a$ and $b$.






















\begin{onlineOnly}
\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Formulas]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/formulas/examples/exampleList}

\end{center}
\end{onlineOnly}





\end{document}
