\documentclass{ximera}

\input{../../preamble.tex}

\title{Linearization}

\begin{document}

\begin{abstract}
%%%
\end{abstract}
\maketitle







Most functions are too complex to anlayze algebraically. Therefore, we devise many ways to approximate their values. That is, we create other functions, easier functions, that are close to the original function.

That doesn't really work, because functions change too much.  But, we can localize this idea.

We can create linear functions that approximate the original function in a little interval around a single domain number.











\subsection{Learning Outcomes}


\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item Define linear approximation as an application of the tangent to a curve.
\item Find the linear approximation to a function at a point and use it to approximate the function value.
\item Identify when a linear approximation can be used.
\item Label a graph with the appropriate quantities used in linear approximation.
\item Find the error of a linear approximation.
\item Compute differentials.
\item Contrast the notation and meaning of $dy$ versus $\Delta y$.
\end{itemize}
\end{sectionOutcomes}









\begin{center}
\textbf{\textcolor{green!50!black}{ooooo=-=-=-=-=-=-=-=-=-=-=-=-=ooOoo=-=-=-=-=-=-=-=-=-=-=-=-=ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Linearization]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/linearization/examples/exampleList}

\end{center}













\end{document}
