\documentclass{ximera}

\input{../../preamble.tex}

\title{Algebraic Geometry}

\begin{document}

\begin{abstract}
%%%
\end{abstract}
\maketitle





The real number line had just one dimension, which made the idea of direction simple - two of them: left and right.  These were commonly known as the positive and negative directions.  Each number was described with two pieces of information.  Each number came with a direction and a distance.  We usually give the direction information first and then the distance.


\begin{itemize}
\item For example $-3$.  The hyphen out front means the left (negative) direction and the $3$ gives the distance.
\item For example $+5$.  The plus sign out front means the right (positive) direction and the $5$ gives the distance.
\end{itemize}
\textbf{Note:} We usually do not write the $+$, $+3 = 3$.

We represent the direction first.  A hyphen appears to the left of the distance number for a negative number. Or, the a hyphen is absent for a positive number.

The Complex Numbers have many more directions, but the idea is the same.  Each complex number can be described with a direction and a distance.  This type of representation is known as \textbf{polar form}.






\subsection*{Learning Outcomes}



\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item represent real numbers on the 2D number line in polar form.
\end{itemize}
\end{sectionOutcomes}













\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Polar Form of Complex Numbers]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/polarFormComplexNumbers/examples/exampleList}

\end{center}







\end{document}
