\documentclass{ximera}

\input{../../preamble.tex}

\title{Analyzing More Functions}

\begin{document}

\begin{abstract}
%
\end{abstract}
\maketitle











What do we want when analyzing a function?


We want \textbf{\textcolor{red!80!black}{reasoning}} on how we figure out

\begin{itemize}
\item Domain
\item Zeros 
\item Continuity (Discontinuities, Singularities, and Behavior)
\item End-behavior
\item Behavior (Increasing and Decreasing)
\item Global Maximum and Minimum
\item Local Maximums and Minimums
\item Range
\end{itemize}



Of course, we want a nice, informative graph
\begin{itemize}
	\item labeled axes
	\item intercepts
	\item endpoints
	\item vertical asymptotes
	\item horizontal asymptotes
	\item arrows
\end{itemize}


A graph is simply a communication tool.  It is inherently inaccurate by the very fact that it is drawn.  Algebra and functional reasoning are the only tools for exactness, which is our end goal. \\



$\blacktriangleright$ \textbf{\textcolor{red!70!black}{Reasoning:}} Reasoning is a logical explanation that describes our conclusions, how we arrived at those conclusions, and why we think those conclusions are correct. \\

Analysis is not a list of conclusions. We are not looking for such a list. \\

We are certainly not looking for such a list based on viewing a graph. \\

We are looking for the thought process that arrived at the list of conclusions. \\









\subsection{Learning Outcomes}



\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item analyze functions from their formula.
\end{itemize}
\end{sectionOutcomes}








\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Analyzing More Functions]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/analyzingMoreFunctions/examples/exampleList}

\end{center}





\end{document}
