\documentclass{ximera}

\input{../../preamble.tex}

\title{Describe Everything}

\begin{document}

\begin{abstract}
features
\end{abstract}
\maketitle





\begin{example}  Complete Analysis


Completely analyze   

\[   B(f) = \frac{(f+3)(2f-1)}{(f+5)(f-2)}        \]

with


\[
B'(f) = \frac{f^2 - 34f - 41}{(f+5)^(f-2)^}
\]





Categorize:  $B(f)$ is a rational function, because it is of the form $\frac{polynomial}{polynomial}$




\textbf{Domain}

The natural domain of a rational function is all real numbers except the zeros of the denominator, which are $-5$ and $2$. \\

The domain is $(-\infty, -5) \cup (-5, 2) \cup (2, \infty)$.







\textbf{Zeros}

The zeros of a rational function are the zeros of the numerator, which are not also zeros of the denominator, i.e. they are in the domain. \\





$-3$ and $\answer{\frac{1}{2}}$ are the zeros of $B$. \\





\textbf{Continuity}


Rational functions are continuous. \\

Rational functions can have singularities.  Here the zeros of the denominator are $-5$ and $2$.  These are not zeros of the numerator.  That makes them asymptotic singularities. So, we know $B$ is unbounded near them.  We just need to figure out the sign.\\


Let's map out the signs of $B$. \\



$B$ has four zeros and/or singularities. In order, they are $-5$, $-3$, $\frac{1}{2}$, and $2$.\\

They all have multiplicity $1$, which is odd.  That means $B$ switches signs across them. \\


The end-behavior will get us started on the signs.
















\textbf{End-Behavior}


$B$ is a rational function.

The degree of the numerator is $\answer{2}$. \\


The degree of the denominator is $\answer{2}$. \\


$B$ is a rational function and the degree of the denominator is \wordChoice{\choice{greater than} \choice{less than} \choice[correct]{equal to}} the degree of the numerator. \\


Since the degrees of the numerator and denominator are equal, the end-behvior is the quotient of the leading coefficients.


\[ \lim_{f \to -\infty} B(f) = \frac{2}{1} = 2   \, \text{ and }  \,  \lim_{f \to \infty} B(f) = \frac{2}{1} = 2    \]





\begin{explanation} Continuing with singularity behavior....




We now know that $B$ is positive for large negative domain numbers.  This switches at $-5$, which is the first singularity.

$B$ is positive on $(-\infty, -5)$ \\

That tells us that 

\[
\lim\limits_{f \to -5^-} B(f) = \infty 
\]

The sign of $B$ switches at $-5$, which gives us 


\[
\lim\limits_{f \to -5^+} B(f) = -\infty 
\]



$B$ switches signs at $-3$, which is the first zero. \\


Therefore, $B$ is negative on $(-5, -3)$.  Then, $B$ switches to positive on $\left( -3, \frac{1}{2} \right)$ Then, $B$ switches to negative on $\left( \frac{1}{2}, 2 \right)$.


That tells us that 

\[
\lim\limits_{f \to 2^-} B(f) = -\infty 
\]


Then, $B$ switches to positive on $(2, \infty)$.


\[
\lim\limits_{f \to 2^+} B(f) = \infty 
\]


\end{explanation}
















\textbf{Behavior (Increasing and Decreasing)}


We'll use the derivative to help us with behavior. \\


\[
B'(f) = \frac{f^2 - 34f - 41}{(f+5)^(f-2)^}
\]


$B'(f)$ is also a rational function.  So, it is continous.  It has the same singularities as $B$.  But it has different zeros. \\


To get the critical numbers, we need to factor the numerator.


\[
f = \frac{34 \pm \sqrt{(-34)^2 - 4 (1)(-41)}}{2(1)} = \frac{34 \pm \sqrt{1320}}{2} = \frac{34 \pm 2\sqrt{330}}{2} = 17 \pm \sqrt{330}
\]




Let's check with a graph.

$17 - \sqrt{330} \approx -1.165902125$

$17 + \sqrt{330} \approx 35.16590212$







\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{yjhhl3llii}{400}{300}
\end{center}





\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{le6wkkt3uz}{400}{300}
\end{center}




It seems our algebra is good.



\[
B'(f) = \frac{(f - (17 - \sqrt{330}))(f - (17 + \sqrt{330}))}{(f+5)^2(f-2)^2}
\]


$B'$ has two zeros and two singularities.

\[
17 - \sqrt{330}, 17 + \sqrt{330}, -5, 2
\]

We need their order. \\




The graph suggests that $-5 < 17 - \sqrt{330}$.  Let's see if we can show that is true. \\



\[
-5 < -1 = 17 - 18 = 17 -\sqrt{324} < 17 - \sqrt{330}
\]


Since $ \sqrt{330} > \sqrt{289} = 17$, we know that $17 - \sqrt{330} < 0$. \\


We also know that $2 < 17 + \sqrt{330} $. \\


That gives us

\[
-5 < 17 - \sqrt{330} < 2 < 17 + \sqrt{330}
\]


$17 - \sqrt{330}$ and $17 + \sqrt{330}$ are zeros of $B'$ with multiplicity $1$, which is odd.  $B'$ will change signs across them.



$-5$ and $2$ are singularities of $B'$ with multiplicity $2$, which is even.  $B'$ will not change signs across them.


We just need a starting sign. \\



For very large negative numbers (anything less than $17 - \sqrt{330}$), 

\[
B'(f) = \frac{negative \cdot negative}{positive \cdot positive} = positive
\]


Therefore, the signs of $B'$ are


\begin{itemize}
  \item $B'$ is positive on $(-\infty, -5)$
  \item $B'$ is positive on $(-5, 17 - \sqrt{330})$
  \item $B'$ is negative on $(17 - \sqrt{330}, 2)$
  \item $B'$ is negative on $(2, 17 + \sqrt{330})$
  \item $B'$ is positive on $(17 + \sqrt{330}, \infty)$
\end{itemize}



This give us the behavior of $B$.





\begin{itemize}
  \item $B$ is increasing on $(-\infty, -5)$
  \item $B$ is increasing on $(-5, 17 - \sqrt{330})$
  \item $B$ is decreasing on $(17 - \sqrt{330}, 2)$
  \item $B$ is decreasing on $(2, 17 + \sqrt{330})$
  \item $B$ is increasing on $(17 + \sqrt{330}, \infty)$
\end{itemize}



















\textbf{Global Maximums and Minimums)}


$B$ has no global maximum or minimum, because


\[
\lim\limits_{f \to 2^-} B(f) = -\infty 
\]




\[
\lim\limits_{f \to 2^+} B(f) = \infty 
\]





\textbf{Local Maximums and Minimums)}


$B$ has two critical numbers.  They are locations for local maximums or minimums.






\begin{itemize}
  \item $B$ is increasing on $(-5, 17 - \sqrt{330})$
  \item $B$ is decreasing on $(17 - \sqrt{330}, 2)$
\end{itemize}


Therefore, $B(17 - \sqrt{330})$ is a local maximum occurring at $17 - \sqrt{330}$.





\begin{itemize}
  \item $B$ is decreasing on $(2, 17 + \sqrt{330})$
  \item $B$ is increasing on $(17 + \sqrt{330}, \infty)$
\end{itemize}



Therefore, $B(17 + \sqrt{330})$ is a local minimum occurring at $17 + \sqrt{330}$.









\textbf{Range)}


To figure out the range, we need to know if there is a gap.  We need to know if $B(17 + \sqrt{330})$ or $B(17 - \sqrt{330})$ is greater. \\

The graph suggests that 

\[
B(17 + \sqrt{330})  > B(17 - \sqrt{330})
\]


That is going to be a bit of algebra. Let's ask Wolfram Alpha to simply each of those.



\[
B(17 - \sqrt{330}) = \frac{1}{49} (61 - 2 \sqrt{330})
\]




\[
B(17 + \sqrt{330})  =  \frac{1}{49} (61 + 2 \sqrt{330})
\]


Confirmed.


\[
B(17 + \sqrt{330})  > B(17 - \sqrt{330})
\]


There is a gap in the range.


The range of $B$ is 


\[
\left( -\infty, \frac{1}{49} (61 - 2 \sqrt{330}) \right) \cup  \left( \frac{1}{49} (61 + 2 \sqrt{330}), \infty \right) 
\]





\end{example}














\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Function Analysis]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/functionAnalysis/examples/exampleList}

\end{center}




\end{document}
