\documentclass{ximera}

\input{../../preamble.tex}

\title{Alternate Formulas}

\begin{document}

\begin{abstract}
many representations
\end{abstract}
\maketitle





We use algebraic expressions as descriptions of relationships and when we wish to identify specific items within the relationship, we create equations and attempt to solve them.

As it turns out, it is almost impossible to look at an equation and recognize its solutions.


There are very few forms we can maniplulate in our heads that easily.



\textbf{Simple Equations}

$\blacktriangleright$ Something like $3x = 7$ can be solved in your head.  Whatever $x$ is, it gets rid of the $3$ and inserts a $7$.  That gives us $x =\frac{7}{3}$





$\blacktriangleright$ Something like $y^2 = 11$ can be solved in your head.  Whatever $y$ is, when you square it, you get $11$.  That gives us $\sqrt{11}$ and $\-sqrt{11}$.  We can handle more complicated  quadratics with the use of the quadratic formula.




$\blacktriangleright$ Something like $4^t = 9$ can be solved in your head.  $t$ is the thing you raise $4$ to, to get $9$.  That gives us $log_4(9)$.



$\blacktriangleright$ Something like $log_3(k) = 2$ can be solved in your head.  The equation tells us that $2$ is the thing you raise $3$ to, to get $k$.  That makes $k = 3^2$.



$\blacktriangleright$ Something like $\sqrt{m} = 5$ can be solved in your head.  $m = 5^2$.


$\blacktriangleright$ Something like $sin(\theta) = \frac{1}{2}$ can be solved in your head, once you have some trigonometric facts memorized.



We are somewhat fluent in really small equations that involve familiar functions.  



Start complicating the equations and we are just out of luck.


So, our strategy is to break up our equations into these types of equations.  The only way we know how to break up equations is by factoring (distributive property).  And, factoring only works with the zero product property.



\section{Strategy}


Our strategy has two parts:


(1) If you see what the answer should look like, then start guessing and refine your guesses.\\
(2) Get everything on one side and $0$ on the other side of the equal sign. Use the distributive property to factor.  Break up according to the zero product property. \\




There you have it. Not much for $2,000$ years of investigations.

On the one hand, you know what to do.  On the other hand, doing it is not easy.













\end{document}
