\documentclass{ximera}

\input{../../preamble.tex}

\title{Equivalent Forms}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}
\maketitle




Formulas are not functions.  Graphs are not functions. 


A function is a collection of three sets: a domain, a range, and a set of pairs, along with one rule, which states that each domain item is included in exactly one pair.

Functions do not have Variables

Formulas and Graphs are tools representing those pairs. They have variables, which we use to represent pieces of the function.  


In fact, there are always many formulas representing the same function. This is very very very helpful.

For example, each formula below represents the same function on $(-\infty, \infty)$.


\begin{itemize}
\item $x^2$
\item $(x - 1)^2 + 2 x - 1$
\item $\sqrt{x^4}$
\item $ x \cdot x$
\item $ 4 \cdot \left( \frac{x}{2} \right)^2$
\item $\begin{cases}
  \frac{x^3}{x} &\text{if $x \ne 0$,}\\
  0 &\text{if $x = 0$}.
\end{cases}$
\end{itemize}




Much of our algebraic reasoning relies on stepping through equivalent formulas.




\subsection*{Lerning Outcomes}


\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item compare equivalent forms.
\item compare nearly equivalent forms.
\end{itemize}
\end{sectionOutcomes}










\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Equivalent Forms]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/equivalentForms/examples/exampleList}

\end{center}







\end{document}
