\documentclass{ximera}

\input{../../preamble.tex}

\title{Polynomials}

\begin{document}

\begin{abstract}
rational roots
\end{abstract}
\maketitle




Let $f(x)$ be a polynomial with rational coefficients.

\[   f(x) = b_n x^n + b_{n-1} x^{n-1} + \cdots + b_2 x^2 + b_1 x + b_0     \]


where $b_k \in \mathbb{Q}$:  $b_k = \frac{n_k}{d_k}$


Then we could get a common denominator of all of the coefficients and factor that out front



\[   f(x) = \frac{1}{d}(c_n x^n + c_{n-1} x^{n-1} + \cdots + c_2 x^2 + c_1 x + c_0)    \]


These are the same polynomial, so that have the same roots and factorization.


So, when we are investigating zeros and roots and factors, we can always assume that a polynomial with rational coefficients just has integer coefficents.




\section{Rational Roots}

Let $f(x)$ be a polynomial with integer coefficients.

\[   f(x) = a_n x^n + a_{n-1} x^{n-1} + \cdots + a_2 x^2 + a_1 x + a_0     \]


Suppose $f(x)$ has a rational root:  $\frac{N}{D}$ in reduced form.  That is $N$ and $D$ do not share any prime factors.



That means


\[    f \left( \frac{N}{D} \right) = a_n \left( \frac{N}{D} \right)^n + a_{n-1} \left( \frac{N}{D} \right)^{n-1} + \cdots + a_2 \left( \frac{N}{D} \right)^2 + a_1 \left( \frac{N}{D} \right) + a_0  = 0       \]


Let's factor out $\frac{1}{D^n}$





\[   \frac{1}{d^n} (a_n N^n + a_{n-1} N^{n-1} D + \cdots + a_2 N^2 D^{n-2}+ a_1 N D^{n-1} + a_0 D^n)  = 0       \]


That tells us

\[   a_n N^n + a_{n-1} N^{n-1} D + \cdots + a_2 N^2 D^{n-2}+ a_1 N D^{n-1} + a_0 D^n  = 0       \]



\[   a_n N^n  =   -(a_{n-1} N^{n-1} D + \cdots + a_2 N^2 D^{n-2}+ a_1 N D^{n-1} + a_0 D^n)      \]


$D$ is a factor of the right side, so $D$ must be a factor of the left side,  But $D$ is not a factor of $N$, because the original rational root was in reduced form.  Therefore $D$ must be a factor of $a_n$.

Similar reasoning tells us that $N$ is a factor  $a_0$.





\begin{theorem} Rational Root Theorem


Let $ f(x) = a_n x^n + a_{n-1} x^{n-1} + \cdots + a_2 x^2 + a_1 x + a_0 $ be a polynomial with integer coefficients.

Let $\frac{N}{D}$ be a rational root of $f(x)$.

Then $D$ is a factor of $a_n$, the leading coefficient and $N$ is a factor of $a_0$, the constant term.

\end{theorem}











\begin{example}  Roots and Factoring


Let $T(h) = 2 h^2 - 9h - 5$



If $T(h)$ has a rational root, $\frac{N}{D}$, then


\begin{itemize}
\item $N$ is a factor of $5$: $-5$, $-1$, $1$, or $5$
\item $D$ is a factor of $2$: $-2$, $-1$, $1$, or $2$
\end{itemize}



Possibilities are 

\[  \frac{5}{2}, \frac{-5}{2}, \frac{5}{1}=5, \frac{-5}{1}=-5, \frac{-1}{-1}=1, \frac{-1}{1}=-1, \frac{1}{2}, \frac{-1}{2}   \]


There are only two roots. If they are rational roots, then they are among these.


We can try each one.



\begin{itemize}
\item $T\left( \frac{5}{2} \right) = -15$
\item $T\left( \frac{-5}{2} \right) = 30$
\item $T(5) = 0$   : a root
\item $T(-5) = 90$
\item $T(1) = -12$   
\item $T(-1) = 6$
\item $T\left( \frac{1}{2} \right) = -9$
\item $T\left( \frac{-1}{2} \right) = 0$ : a root
\end{itemize}

We have two roots, which gives us two factors.


$T(h) = 2\left(h-\frac{1}{2} \right)(h-5)$

We could also mulitply the $2$ inside the first factor.


$T(h) = (2h - 1)(h-5)$



\end{example}







\end{document}
