\documentclass{ximera}

\input{../../preamble.tex}

\title{Visual Analysis}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}
\maketitle




Functions are packages containing three sets (one of which is a set of pairs), which satisfies one rule.  That is the struture of a function.  In Analysis, the sets are often measurements and the function is how we compare those measurements.

We would like to know things like:

\begin{itemize}
\item When one measurement increases, does the other also increase?
\item When one measurement increases, does the other decrease?
\item When one measurement changes quickly, does the other change quickly or slowly?
\item Which measurements are connected to maximums and minimums in the other measurements?
\item If one measurement is smooth, then is the other smooth or does it have sudden changes?
\end{itemize}


In this section, we will introduce this type of \textbf{function analysis} via graphs. Once we gain a graphical familiarity with this type of analysis, we will extend our thinking to include formulas (Algebra).  Our eventual goal is to be exact with our analysis.  This will require both our graphical and algebraic tools working together.









\subsection{Learning Outcomes}

\begin{sectionOutcomes}
In this section, students will...

\begin{itemize}
\item locate intervals where functions are increasing and decreasing.
\item identify maximums and minimums.
\item find zeros.
\end{itemize}
\end{sectionOutcomes}



















\begin{center}
\textbf{\textcolor{green!50!black}{ooooo=-=-=-=-=-=-=-=-=-=-=-=-=ooOoo=-=-=-=-=-=-=-=-=-=-=-=-=ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Visual Analysis]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/visualAnalysis/examples/exampleList}

\end{center}









\end{document}
