\documentclass{ximera}

\input{../../preamble.tex}

\title{Sine}

\begin{document}

\begin{abstract}
analysis
\end{abstract}
\maketitle












\textbf{\textcolor{purple!85!blue}{Analyze  $T(r) =  2 \sin(3r - \pi) - 1$}}



\begin{template}


We are going to view this as a composition of three functions.


\[
T = Out \circ \sin \circ In
\]


where


$Out(x) = 2x - 1$ \\


$In(y) = 3y - \pi$ \\




Along with $\sin(\theta)$, these three component functions, we get



\[
T(r) = (Out \circ \sin \circ In)(r) = Out(sin(In(r)))  = 2 \sin(3r - \pi) - 1
\]




\end{template}

$T$ is a periodic function, since $\sin(\theta)$ is a periodic function. \\

The principal interval of $\sin(\theta)$ is $[0, 2\pi)$.  The principal interval of $T$ will be the values of $r$ that make the values of $In(y)$ run from $0$ to $2\pi$. \\


$3y - \pi = 0$ at $y = \frac{\pi}{3}$

$3y - \pi = 2\pi$ at $y = \pi$

The principle interval of $T$ is $\left[ \frac{\pi}{3}, \pi \right)$. \\

The length of this interval is $\frac{2\pi}{3}$, which is the period of $T$. \\









\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{b9yiwlxg2n}{400}{300}
\end{center}





\begin{observation}

$T$ is a periodic function with period $\frac{2\pi}{3}$. \\


Therefore, our analysis will focus on the principal interval of $\left[ \frac{\pi}{3}, \pi \right)$. \\


All of the features and characteristics we discover will repeat wiht a period of $\frac{2\pi}{3}$.\\


\end{observation}









\textbf{\textcolor{blue!55!black}{Domain}}


The domain of $Out(x) = 2x - 1$ is $(-\infty, \infty)$, because $Out$ is a linear function.  \\

This includes any value of $\sin(\theta)$. That means we can use the entire domain of $\sin(\theta)$, which is $(-\infty, \infty)$.  \\


This includes any value of $In(y)$, which means we can use the entire domain of $In(y)$, which is $(-\infty, \infty)$, because $In$ is a linear function.  \\


The domain of $T$ is $(-\infty, \infty)$. \\








\textbf{\textcolor{blue!55!black}{Zeros}}


Since the values of $T$ come from the values of $Out$, we are first looking for zeros of $Out$.  $Out(x) = 2x - 1$ is a linear function and has only one zero, $\frac{1}{2}$. \\

We are looking for where $\sin(\theta) = \frac{1}{2}$. \\


There are two numbers in $[0, 2\pi)$ where $\sin(\theta) = \frac{1}{2}$.  They are $\frac{\pi}{6}$ and $\frac {5\pi}{6}$.

We need the values of $y$ where $In(y) = 3y - \pi = \frac{\pi}{6}$ and $In(y) = 3y - \pi = \frac{5\pi}{6}$

$y = \frac{7\pi}{18}$  and   $y = \frac{11\pi}{18}$

[ These agree with the graph. ]


And, these repeat every $\frac{2\pi}{3}$ for $T$.












\textbf{\textcolor{blue!55!black}{Continuity}}


The component functions of the composition are linear and sine, all continuous. \\

The composition of continuous function is continuous.

$T$ is continuous. It has no disontinuities. \\


Since the domain is $(-\infty, \infty)$, there are no singularities. \\








\textbf{\textcolor{blue!55!black}{End-Behavior}}


$T$ is a periodic function with period $\frac{2\pi}{3}$. \\

It either has no end-behavior or the end-behavior is that it is periodic.












\textbf{\textcolor{blue!55!black}{Behavior}}


We need the behavior of each of the three component funcitons and then we will compose them together. \\




$Out(x) = 2x - 1$ is a linear function with a positive leading coefficient, so it is an increasing function.\\


$In(y) = 3y - \pi$ is a linear function with a positive leading coefficient, so it is an increasing function.\\


$\sin(\theta)$ increases and decreases through the quadrants. For the principal interval, we have


\begin{itemize}
  \item increases on $\left( 0, \frac{\pi}{2} \right)$
  \item decreases on $\left( \frac{\pi}{2}, \pi \right)$
  \item decreases on $\left( \pi, \frac{3\pi}{2} \right)$
  \item increases on $\left( \frac{3\pi}{2}, 2\pi \right)$
\end{itemize}


Now to trace domains and ranges.


We need the values of $In(y) = 3y - \pi$ to be $0$, $\frac{\pi}{2}$, $\pi$, $\frac{3\pi}{2}$, and $2\pi$. \\



$\blacktriangleright$ $3y - \pi = 0$ when $y = \frac{\pi}{3}$ \\

$\blacktriangleright$ $3y - \pi = \frac{\pi}{2}$ when $y = \frac{\pi}{2}$ \\

$\blacktriangleright$ $3y - \pi = \pi$ when $y = \frac{2\pi}{3}$ \\

$\blacktriangleright$ $3y - \pi = \frac{3\pi}{2}$ when $y = \frac{5\pi}{6}$ \\

$\blacktriangleright$ $3y - \pi = 2\pi$ when $y = \pi$ \\





Now to glue everything together. \\



\textbf{\textcolor{purple!80!black}{On $r = y \in \left( \frac{\pi}{3}, \frac{\pi}{2} \right)$, }}



$In(y) = 3y - \pi$ is increasing.  The range of $In(y)$ is $\theta = In(y) \in \left(0, \frac{\pi}{2} \right)$, where $\sin(\theta)$ is increasing. \\



The range of $\sin(\theta)$ on $\theta \in \left(0, \frac{\pi}{2} \right)$ is $x = \sin(\theta) \in (0, 1)$, where $Out(x)$ is increasing, because $Out$ is always increasing. \\


\[
T(r) = increasing \circ increasing \circ increasing = increasing
\]








\textbf{\textcolor{purple!80!black}{On $r = y \in \left( \frac{\pi}{2}, \frac{2\pi}{3} \right)$, }}




$In(y) = 3y - \pi$ is increasing.  The range of $In(y)$ is $\theta = In(y) \in \left(\frac{\pi}{2}, \pi \right)$, where $\sin(\theta)$ is decreasing. \\



The range of $\sin(\theta)$ on $\theta \in \left( \frac{\pi}{2}, \pi \right)$ is $x = \sin(\theta) \in (0, 1)$, where $Out(x)$ is increasing, because $Out$ is always increasing. \\


\[
T(r) = increasing \circ decreasing \circ increasing = decreasing
\]












\textbf{\textcolor{purple!80!black}{On $r = y \in \left( \frac{2\pi}{3}, \frac{5\pi}{6} \right)$, }}




$In(y) = 3y - \pi$ is increasing.  The range of $In(y)$ is $\theta = In(y) \in \left(\pi, \frac{3\pi}{2} \right)$, where $\sin(\theta)$ is decreasing. \\



The range of $\sin(\theta)$ on $\theta \in \left( \pi, \frac{3\pi}{2} \right)$ is $x = \sin(\theta) \in (-1, 0)$, where $Out(x)$ is increasing, because $Out$ is always increasing. \\


\[
T(r) = increasing \circ decreasing \circ increasing = decreasing
\]










\textbf{\textcolor{purple!80!black}{On $r = y \in \left( \frac{5\pi}{6}, \pi \right)$, }}




$In(y) = 3y - \pi$ is increasing.  The range of $In(y)$ is $\theta = In(y) \in \left( \frac{3\pi}{2}, 2\pi \right)$, where $\sin(\theta)$ is increasing. \\



The range of $\sin(\theta)$ on $\theta \in \left( \frac{3\pi}{2}, 2\pi \right)$ is $x = \sin(\theta) \in (-1, 0)$, where $Out(x)$ is increasing, because $Out$ is always increasing. \\


\[
T(r) = increasing \circ increasing \circ increasing = increasing
\]







This behavior repeats with a period of $\frac{2\pi}{3}$. \\



[ This agrees with the graph. ]





\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{bntu7gqkni}{400}{300}
\end{center}



















\textbf{\textcolor{blue!55!black}{Local Maximum and Minimum}}



$T$ is continuous on $(-\infty, \infty)$.


$\blacktriangleright$ $T$ switches from increasing to decreasing at $\frac{\pi}{2}$, which makes $\frac{\pi}{2}$ a crtical number and the location of a local maximum. \\




$\blacktriangleright$ $T$ switches from decreasing to increasing at $\frac{5\pi}{6}$, which makes $\frac{5\pi}{6}$ a crtical number and the location of a local minimum. \\


These repeat with a period of $\frac{2\pi}{3}$. \\



\[
T\left( \frac{\pi}{2} \right) = 2 \sin\left(3 \cdot \frac{\pi}{2} \right) - 1 = -3
\]




\[
T\left( \frac{5\pi}{6} \right) = 2 \sin\left(3 \cdot \frac{5\pi}{6} \right) - 1 = 1
\]












\textbf{\textcolor{blue!55!black}{Global Maximum and Minimum}}


Since $T$ is continuous on $(-\infty, \infty)$, $\frac{\pi}{2}$ and $\frac{5\pi}{6}$ are the only critical numbers, we also have global extrema here. \\




\[
T\left( \frac{\pi}{2} \right) = 2 \sin\left(3 \cdot \frac{\pi}{2} \right) - 1 = -3
\]




\[
T\left( \frac{5\pi}{6} \right) = 2 \sin\left(3 \cdot \frac{5\pi}{6} \right) - 1 = 1
\]




These repeat with a period of $\frac{2\pi}{3}$. \\









\textbf{\textcolor{blue!55!black}{Range}}


Since $T$ is continuous with a global maximum of $1$ and a global minimum of $-1$, the range is $[-3, 1]$.  \\ 



[ This agrees with the graph. ]
























\subsection*{Periodic}

$T$ is periodic with period $\frac{2\pi}{3}$. \\



We have zeros in the principal interval as $\frac{7\pi}{18}$ and $\frac{11\pi}{18}$. \\


To describe all of the zeros of $T$, we need to include all numbers which are these two numbers plus or minus and whole number of $\frac{2\pi}{3}$. \\



\[
\left\{  \frac{7\pi}{18} + k  \cdot \frac{2\pi}{3}  \, | \,  k \in \mathbb{Z}         \right\}
\]



\[
\left\{  \frac{11\pi}{18} + k  \cdot \frac{2\pi}{3}  \, | \,  k \in \mathbb{Z}         \right\}
\]




Similarly, we could describe the intervals by including $k  \cdot \frac{2\pi}{3} $. \\














\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Trigonometric Functions]{https://ximera.osu.edu/csccmathematics/precalculus/precalculus/trigFunctions/examples/exampleList}

\end{center}


\end{document}

