\documentclass{ximera}

\input{../../preamble.tex}

\title{Sequences}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}

\maketitle





The elementary functions are but a drop in the bucket of the collection of all functions.  We just don't have formulas for most functions.  Therefore, we invent new ways of describing them. \\


One method of describing a function is to mimic how we use decimal approximations to describe irrational numbers. \\


For example $\sqrt{17}$. \\

Follow this list of decimal numbers:


\begin{itemize}
\item $4.1$
\item $4.12$
\item $4.123$
\item $4.1231$
\item $4.12310$
\item $4.123105$
\item $4.1231056$
\item $4.12310562$
\item $4.123105$626
\end{itemize}


This lists keeps going.  There is no end to the list.  $\sqrt{17}$ is the number that this list is getting closer and closer to. \\



We can give a list of elementary functions, explain the pattern in the list, and then say the functions in the list are getting closer and closer to the function being described. \\

These lists are called \textbf{sequences}. \\







\subsection{Expectations}


\begin{sectionOutcomes}

In this section, students will

\begin{itemize}
\item Define a sequence.
\item Write the first several terms of a sequence using an explicit formula.
\item Write the first several terms of a sequence using a recurrence
  relation.
\item Find an explicit formula for a sequence given recursively.
\item Find a recurrence relation for a sequence given explicitly.
\end{itemize}

\end{sectionOutcomes}

\end{document}
