\documentclass{ximera}

\input{../../preamble.tex}

\title{Core Analysis}

\begin{document}

\begin{abstract}
Part 1
\end{abstract}
\maketitle




As a starting point for function analysis, let's list characteristics of our Core Functions. A full analysis would also provide reasoning.


\begin{explanation} \textbf{\textcolor{red!75!green}{Constant Functions}} 


\[ Constant(x) = C  \]



\textbf{Domain:} The domain of every constant function is always $(-\infty, \infty)$.

\textbf{Zeros:} If the constant funciton just happens to be the zero function and every value is $0$, then every number in the domain is a zero of the function.

Otherwise, constant functions do not have zeros.

\textbf{Continuity:} Constant functions are always contiuous.

\textbf{End-Behavior:} The end-behavior is just the value of constant function.


\textbf{Behavior (Inc/Dec):} Constant functions do not increase and they do not decrease.

\textbf{Global Maximum and Minimum:} Since a constant function has only one value, that value is the global maxmum and global minimum and they both occur at every domain number.

\textbf{Local Maximum and Minimum:}  Since a constant function has only one value, that value is a local maxmum and local minimum and they both occur at every domain number.

\textbf{Range:} The range of a constant function consists of just one number, the value of the function.

\end{explanation}









\begin{explanation} \textbf{\textcolor{red!75!green}{The Identity Function}} 


\[ Id(x) = x  \]



\textbf{Domain:} The domain of the identity funciton is $(-\infty, \infty)$.

\textbf{Zeros:} $0$ is the only zero of the identity function.

\textbf{Continuity:} The identity function is a continuous function.

\textbf{End-Behavior:} 

As the domain becomes unbounded positively, the identity also becomes unbounded positively.  Our notation for this will look like

\[
\lim\limits_{x \to \infty} Id(x) = \infty
\]

As the domain becomes unbounded negatively, the identity also becomes unbounded negatively.  Our notation for this will look like

\[
\lim\limits_{x \to -\infty} Id(x) = -\infty
\]




\textbf{Behavior (Inc/Dec):} The identity is an increasing function.

\textbf{Global Maximum and Minimum:} The identity function has no global maximum or global minimum.  This is explained by the end-behavior.




\textbf{Local Maximum and Minimum:}  The identity function has no local maximums or global minimums. 

\textbf{Range:} The range of the identity function is $(-\infty, \infty)$.

\end{explanation}













\begin{explanation} \textbf{\textcolor{red!75!green}{Power Functions}} 


The power makes a big difference here.  As a beginning, we will only consider two types of powers: positive integers and negative integers,


\[ Power(x) = x^n  \]


\textbf{Domain:} 

If $n$ is a whole number, then the domain is $(-\infty, \infty)$.

If $n$ is a negative integer, then the domain is $(-\infty, 0) \cup (0, \infty)$.


\textbf{Zeros:} 

If $n$ is a whole number, then $0$ is the only zero.

If $n$ is a negative integer, then the power function has no zeros.


\textbf{Continuity:} Power functions are continuous functions

\textbf{End-Behavior:} 

If $n$ is a whole number, then there are two different cases.

\begin{itemize}
\item If $n$ is odd, then the power function is unbounded positively as the domain becomes unbounded positively or negative.
\item If $n$ is even, then the power function is unbounded positively in both directions.
\end{itemize}

If $n$ is a negative integer, then the power function tends to $0$ as the domain becomes unbounded positively or negatively.



\textbf{Behavior (Inc/Dec):} 

If $n$ is a whole number, then there are two different cases.

\begin{itemize}
\item If $n$ is odd, then the power function is an increasing function.
\item If $n$ is even, then the power function decreases on $(-\infty, 0)$ and increases on $(0, \infty)$.
\end{itemize}

If $n$ is a negative integer, then there are two different cases.

\begin{itemize}
\item If $n$ is odd, then the power function is a decreasing function.
\item If $n$ is even, then the power function increases on $(-\infty, 0)$ and decreases on $(0, \infty)$.
\end{itemize}



\textbf{Global Maximum and Minimum:} 

A power function has a global minimums only if the power is an even whole number. In that case, $0$ is the global minimum, which occurs at $0$.

\textbf{Local Maximum and Minimum:} 

\textbf{Range:} 

\end{explanation}










\begin{explanation} \textbf{\textcolor{red!75!green}{Root Functions}} 


\[ Root(x) = \sqrt[n]{x}  \]

To begin, we are only considering $n$ to be a natural number.


\textbf{Domain:} 


\begin{itemize}
\item If $n$ is odd, then the domain is $(-\infty, \infty)$.
\item If $n$ is even, then the domain is $[0. \infty)$.
\end{itemize}


\textbf{Zeros:} Core root functions only have $0$ as their only zero.

\textbf{Continuity:}  Core root functions are continuous functions.

\textbf{End-Behavior:} 


\begin{itemize}
\item If $n$ is odd, then the root function is unbounded positively as the domain becomes unbounded positively and unbounded negatively as the domain becomes unbounded negatively.

Our notation for this will look like 


\[
\lim\limits_{x \to \infty} \sqrt[n]{x} = \infty
\]


\[
\lim\limits_{x \to -\infty} \sqrt[n]{x} = -\infty
\]


\item If $n$ is even, then there is only end-behavior in the positive direction.

\[
\lim\limits_{x \to \infty} \sqrt[n]{x} = \infty
\]



\end{itemize}


\textbf{Behavior (Inc/Dec):}  Core root functions are increasing functions.

\textbf{Global Maximum and Minimum:}  

\begin{itemize}
\item If $n$ is even, then the root function has $0$ as its global minimum value and no maximum global value.
\item If $n$ is odd, then there is neither a global minimum nor global maximum.
\end{itemize} 

\textbf{Local Maximum and Minimum:} 


\begin{itemize}
\item If $n$ is even, then the root function has $0$ as its local minimum value and no local maximum values.
\item If $n$ is odd, then there are no local extrema.
\end{itemize} 


\textbf{Range:} 


\begin{itemize}
\item If $n$ is even, then the range is $[0, \infty)$.
\item If $n$ is odd, then the range is $(-\infty, \infty)$.
\end{itemize} 

\end{explanation}































































\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Visual Behavior]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/visualBehavior/examples/exampleList}

\end{center}





\end{document}
