\documentclass{ximera}

\input{../../preamble.tex}

\title{C{\&}M Trucking}

\begin{document}

\begin{abstract}
least cost
\end{abstract}
\maketitle


[From Calculus\&\textit{Mathematica}]




You are the chief dispatcher for the C{\&}M Trucking Company, which sends Mack trucks on the straight shot between Chicago and New Orleans on Interstate 57.  You know that:



\begin{itemize}
\item The run between the two cities is $750$ miles.
\item Running at a steady $50$ miles per hour, the Mack gets $4$ miles per gallon.
\item For each mile per hour increase in speed, the big Mack loses $\frac{1}{10}$ of a mile per gallong in its mileage.
\item The driver team gets $27$ dollars per hour.
\item Keeping the truck on the road costs an extra $12$ dollars per hour over and above the cost of the fuel.
\item Diesel fule for the Mack costs $\$1.19$ per gallon. 
\end{itemize}

Approximately, what steady speed should you tell your drivers to hold in order to make the run at least cost? \\





\begin{question} \boxdot

Which of the following are components of the total cost for the Chicago to New Orleans run?

\begin{selectAll}
\choice[correct] {Fuel}
\choice {Food}
\choice[correct] {Driving Team}
\choice[correct] {Truck Maintanence} 
\choice {Tolls}
\end{selectAll}
\end{question}


\begin{question} \boxdot
If the truck travels at a steady speed of $v$, then how long with it take to complete the trip? \\

time = $\answer{\frac{750}{v}}$ hours.
\end{question}




\begin{question} \boxdot
If the truck travels at a steady speed of $v$, then what will be the cost of the driving team \\

Driving Team Cost = $\answer{\frac{27*750}{v}}$ dollars.
\end{question}




\begin{question} \boxdot
If the truck travels at a steady speed of $v$, then what will be the cost of maintaining the truck? \\

Truck Maintanence Cost = $\answer{\frac{12*750}{v}}$ dollars.
\end{question}



The final cost to calculate is for fuel.

Fuel cost $\$3.09$ per gallon, therefore we need to know how many gallons the trip will take. To calculate galons, we need the mileage for the truck.


\begin{explanation}

If the truck travels at a steady speed of $v$, then it is $\answer{v-50}$ miles per hours over $50$. For each mile per hour over $50$, the truck loses $\frac{1}{10}$ of a mile per gallon from $4$ miles per gallon..

Therefore, it will lose $\answer{(v-50)/10}$ miles per gallon when it is traveling at $v$ miles per hour.

That gives a mileage of $\answer{4-(v-50)/10} \frac{miles}{gallon}$.

\end{explanation}


The truck is travelling $750$ miles and gets $\frac{4-(v-50)}{10} \frac{miles}{gallon}$. From this we can calculate the total number of gallons needed, which will then give us the cost of fuel.






\begin{question} \boxdot
If the truck travels at a steady speed of $v$, then what will be the cost of fuel? \\

Fuel Cost = $ 3.09 \cdot \answer{\frac{750}{4-(v-50)/10}}$ dollars.
\end{question}



That gives a total cost based on speed

\[
Cost(v) = \frac{27 \cdot 750}{v} + \frac{12 \cdot 750}{v} + \frac{3.09 \cdot 750}{4-(v-50)/10}
\]





\begin{center}
\desmos{byvuwohbqw}{400}{300}
\end{center}



From the graph, we can see that there is a minimum cost of \$1161 when the truck is travelling at a speed of approximately $47.62$ miles per hour.




However, we would like an exact speed. \\





Calculus tells us that the derivative of $Cost(v)$ is


\[
Cost'(v) = \frac{23175}{(90-v)^2} - \frac{29250}{v^2}
\]



We can obtain the critical number by setting $Cost'(v)$ equal to $0$ and solving.



By clearing denominators, we get

\[
\answer{23175} \, v^2 - \answer{29250} \, \answer{(90-v)^2}
\]


We can multiply this out and gather like terms to get the quadratic equation

\[
\answer{6075} \, v^2 - \answer{5265000} \, v + 236925000 = 0
\]



There are two solutions, but only one is near $47.62$.


\[
v = \frac{10}{3} \cdot (\answer{130} - \sqrt{13390}) \approx 47.62
\]














\end{document}
