\documentclass{ximera}

\input{../preamble}

\outcome{To be able to use the method of substitution to solve some ``simple'' integrals, with an emphasis on being able to correctly identify what to substitute for.}
\outcome{Undo the Chain Rule.}
\outcome{Calculate indefinite integrals (antiderivatives) using basic substitution.}
\outcome{Calculate definite integrals using basic substitution.}
\title[Dig-In:]{The idea of substitution}
\begin{document}
\begin{abstract}
  We learn a new technique, called substitution, to help us solve
  problems involving integration.
\end{abstract}
\maketitle


Computing antiderivatives is not as easy as computing derivatives.
One issue is that the chain rule can be difficult to ``undo.''  We
have a general method called ``integration by substitution'' that will
somewhat help with this difficulty. 

If the functions are differentiable, we can apply the chain rule and obtain
\[
\frac{d}{dx} f(g(x)) = f'(g(x))g'(x)
\]
If the derivatives are continuous, we can use this equality to evaluate a definite integral. Namely,
\begin{align*}
  \int_a^b f'(g(x))g'(x) dx &= \bigg[ f(g(x)) \bigg]_a^b \\
  &= f(g(b)) - f(g(a)). \\
 \end{align*}
On the other hand, it is also true that
\begin{align*}
  \int_{g(a)}^{g(b)} f'(u) du &= \bigg[ f(u) \bigg]_{g(a)}^{g(b)}\\
  &= f(g(b)) - f(g(a)). \\
 \end{align*}
 Since the right hand sides of these equalities are equal,  the left hand sides must be equal, too. So, it follows that
 
\[
  \int_a^b f'(g(x))g'(x) dx =  \int_{g(a)}^{g(b)} f'(u) du .
\]

 
This simple observation  leads to the following theorem. 


\begin{theorem}[Integral Substitution Formula]\index{integral substitution formula} 
Let $g'$ be the derivative of a  function $g$ and let $f'$ be the derivative of a function $f$. If $g'$ is continuous on the interval $[a,b]$ and  $f'$ is
continuous on the interval $[g(a),g(b)]$, then
\[
\int_a^b f'(g(x)) g'(x) dx =\int_{g(a)}^{g(b)} f'(u) du.
\]
\end{theorem}
The integral on the right appears to be much simpler than the original integral.
So, when evaluating a difficult integral,  we try to apply this theorem and replace the original integral with a simpler one.
 We can do this as long as the conditions of the theorem are satisfied.

We will apply the Integral Substitution Formula (ISF)  to the following example.

\begin{example}
Compute:
\[
\int_0^2 x e^{x^2} dx
\]
\begin{explanation}
Why do we think that the ISF can be useful in this example?

Does this integral have the structure
\[
\int_a^b f'(g(x)) g'(x)  dx ,  
\]
so that we can apply the ISF?

We set $g(x) =x^2$, so $g'(x)dx =2xdx$, and note that

\[
\int_0^2 x e^{x^2} dx
=\frac{1}{2}\int_0^{2} e^{\underbrace{x^2}}\underbrace{2x dx}=\frac{1}{2}\int_0^{2} e^{g(x)}g'(x) dx
\]
So we wrote the integral in a form suitable for application of the ISF. Therefore, we can substitute $u$ for $g(x)$ and $ du$ for $g'(x) dx$ and write

\[
\int_0^2 x e^{x^2} dx=\frac{1}{2}\int_0^{2} e^{g(x)}g'(x) dx=\frac{1}{2}\int_{g(0)}^{g(2)} e^{u} du
\]
Now, compute the endpoints: 
\begin{align*}
g(0) &= 0^2 = 0  \\
g(2) &=2^2 = 4.
\end{align*}
We can easily evaluate the last integral
\[
\frac{1}{2}\int_{g(0)}^{g(2)} e^{u} du=
\frac{1}{2}\int_{0}^{4} e^{u} du=\frac{1}{2}\bigg[ e^u \bigg]_0^4=\frac{1}{2}\left(\answer[given]{e^4}-e^0\right)=\frac{\answer[given]{e^4}-1}{2}
\]
In summary,

\[
\int_0^2 x e^{x^2} dx=\frac{1}{2}\int_{0}^{4} e^{u} du
\]
  The figure below illustrates both integrals.
\begin{image}
  \begin{tikzpicture}
    \begin{axis}[
        xmin=-0.3,xmax=4.2,ymin=-0.3,ymax=3.5,
        clip=true,
        unit vector ratio*=2 2 2,
        axis lines=center,
        grid = major,
         width=3in,
        height=3in,
        ytick={0,3.125},
        xtick={0,2,4},
         yticklabels={0,100},
        xlabel=$x$, ylabel=$y$,
        every axis y label/.style={at=(current axis.above origin),anchor=south},
        every axis x label/.style={at=(current axis.right of origin),anchor=west},
      ]
     \addplot[fill=penColor,fill opacity=0.3,draw=none,domain=0:{2},samples=50]  {((1/32)*x*e^(x^2)} \closedcycle;;   
      \addplot[thick,penColor,domain=0:{2},samples=150] {((1/32)*x*e^(x^2)};   
  % \addplot [draw=none,fill=fillp,domain=8:9, smooth] {f(x)} \closedcycle;
    \node at (axis cs:1.75,0.55) { $A$};  
     \node at (axis cs:3,2.55) { $A=\int_0^2 xe^{x^2} dx$};       
      \node at (axis cs:1.2,1.75) {$y=xe^{x^2}$};
      \end{axis}`
  \end{tikzpicture}
  \begin{tikzpicture}
    \begin{axis}[
        xmin=-0.3,xmax=4.2,ymin=-0.3,ymax=3.5,
        clip=true,
        unit vector ratio*=2 2 2,
        axis lines=center,
        grid = major,
         width=3in,
        height=3in,
        ytick={0,3.125},
        xtick={0,2,4},
         yticklabels={0,100},
        xlabel=$u$, ylabel=$y$,
        every axis y label/.style={at=(current axis.above origin),anchor=south},
        every axis x label/.style={at=(current axis.right of origin),anchor=west},
      ]
      \addplot[fill=penColor,fill opacity=0.3,draw=none,domain=0:{4},samples=50] {(1/64)*e^x} \closedcycle;
      \addplot[thick,draw=penColor,domain=0:{4},samples=50] {(1/64)*e^x};

    \node at (axis cs:3.6,0.3) { $B$}; 
     \node at (axis cs:1,2.55) {$B=\int_0^4 \frac{1}{2}e^{u} du$};           
      \node at (axis cs:2.8,0.7) {$y=\frac{1}{2}e^u$};
      \end{axis}`
  \end{tikzpicture}
\end{image}
The two shaded areas in the figure, $A= \int_0^2 x e^{x^2} dx$ and $B=\frac{1}{2}\int_{0}^{4} e^{u} du$, appear to be equal,
and this confirms what was established by computation before.
  
  
Therefore, whenever we are faced with a problem of evaluating a difficult integral, we try to  replace it with  a simpler one, as long as  both  integrals represent the same net area.


\end{explanation}
\end{example}

We can solve a problem like this in a slightly different
way. Let's do the same example again, this time we will think in terms
of differentials.

\begin{example}\label{main example}
Compute:
\[
\int_0^2 x e^{x^2} dx
\]
\begin{explanation}
Here we will set
\[ u=g(x) = x^2.
\]
 Then 
 \[
 du =g'(x) dx=2x dx.
 \]
Here, we are thinking in terms of differentials.
We can solve for $dx$ to get 
\[
dx = \frac{du}{2x}.
\]
Using this equality, we can write

\begin{align*}
\int_0^2 x e^{x^2} dx &= \int_{g(0)}^{g(2)} x  e^{u} \frac{du}{2x}\\
  &= \int_{0}^{4} \frac{e^u}{2} du .
\end{align*}
At this point, we can continue as we did before and write
\[
 \int_{0}^{4} \frac{e^u}{2} du= \frac{\answer[given]{e^4} -e^0}{2}.
\]
\end{explanation}
\end{example}


















\end{document}
