\documentclass{ximera}

\input{../../preamble.tex}

\title{Sums of sequences}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}

\maketitle




We have noted that decimal approximations or expansions can be viewed as a sequence of numbers, where a new digit is added for each term.


\begin{itemize}
\item $4.1$
\item $4.12$
\item $4.123$
\item $4.1231$
\item $4.12310$
\item $4.123105$
\item $4.1231056$
\item $4.12310562$
\item $4.123105626$
\end{itemize}


For decimal expansions, this pattern continues with no end. \\

We can view this decimal expansion in a different way: as a sum. \\


\[
4 + \frac{1}{10^1} + \frac{2}{10^2} + \frac{3}{10^3} + \frac{1}{10^4} + \frac{0}{10^5} + \frac{5}{10^6} + \frac{6}{10^7} + \frac{2}{10^8} + \frac{6}{10^9} + \cdots
\]



We would like to connect these two views.


One way is to make a sequence from the sum by adding on the next place value digit for each term:  

\[
\frac{4}{10^0}, \frac{41}{10^1},  \frac{412}{10^2}, \frac{4123}{10^3}, \frac{41231}{10^4}, \frac{412310}{10^5}, \frac{4123105}{10^6}, \frac{41231056}{10^7}, \frac{412310562}{10^8}, \frac{4123105626}{10^9}, \cdots
\]


Each term of the sequence is a part of the sum:



\[
\frac{41231}{10^4} = 4 + \frac{1}{10^1} + \frac{2}{10^2} + \frac{3}{10^3} + \frac{1}{10^4} 
\]



When summing up the terms in a sequence we obtain a \textbf{series}, a new way to represent numbers. \\


$\blacktriangleright$  $\sqrt{17} = 0.123105626\cdots$ \\



$\blacktriangleright$ $\sqrt{17} =  \lim\limits_{\infty} \, \{ \frac{4}{10^0}, \frac{41}{10^1},  \frac{412}{10^2}, \frac{4123}{10^3}, \frac{41231}{10^4}, \frac{412310}{10^5}, \frac{4123105}{10^6}, \frac{41231056}{10^7}, \frac{412310562}{10^8}, \frac{4123105626}{10^9}, \cdots \}$ \\



$\blacktriangleright$ $\sqrt{17} = \lim\limits_{\infty} \, 4 + \frac{1}{10^1} + \frac{2}{10^2} + \frac{3}{10^3} + \frac{1}{10^4} + \frac{0}{10^5} + \frac{5}{10^6} + \frac{6}{10^7} + \frac{2}{10^8} + \frac{6}{10^9} + \cdots$ \\











\subsection*{Learning Outcomes}








\begin{sectionOutcomes}

In this section, students will explore

\begin{itemize}
\item defining a series.
\item recognizing a geometric series.
\item recognizing a telescoping series.
\item computing the sum of a geometric series.
\item computing the sum of a telescoping series.
\end{itemize}

\end{sectionOutcomes}












\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Sums of Sequences]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/sumsOfSequences/examples/exampleList}

\end{center}






\end{document}
