\documentclass{ximera}

\input{../../preamble.tex}

\title{Analysis}

\begin{document}

\begin{abstract}
8 characteristics
\end{abstract}
\maketitle












Analyze $H(x) = 5 \sqrt{x^2-3}$ with its natural domain. \\



\subsction*{Analysis}




Our plan is to view this as a composition. \\






We are looking for functions $f$ and $g$ such that $(f \circ g)(x) = f(g(x)) = 5 \sqrt{x^2-3}$ \\


First, identify ``insides'' of functions.  In this case, $x^2 - 3$ is the inside of $5 \sqrt{t}$. \\



\begin{itemize}
\item Let $f(t) = 5 \sqrt{t}$ \\
\item Let $g(y) = y^2 - 3$ \\
\end{itemize}

Algebraically, this produces the formula want, $(f \circ g)(x) = f(g(x)) = 5 \sqrt{x^2-3}$.

To get us started, we are thinking that each component function comes with its natural domain.

\begin{itemize}
\item Let $f(t) = 5 \sqrt{t}$ with domain $[0, \infty)$ \\
\item Let $g(y) = y^2 - 3$ with domain $(-\infty, \infty)$ \\
\end{itemize}

We are also ready to revise and modify and restrict these domains to fit the composition together. \\





We are ready to analyze. \\




\textbf{\textcolor{blue!55!black}{Domain:}} \\


$H$ is not a square root function, since the inside is not a linear function.  However, the formula includes a square root and the inside must not be negative. \\

We need $x^2-3 \geq 0$.  Therefore, the domain of $H$ is is the part of the domain of $g$ that is incommon (intersectis) with $(-\infty, -\sqrt{3} ] \cup [ \sqrt{3}, \infty)$. \\

Since, the natural domainm of $g$ is $(-\infty, \infty)$, the domain of $H$ is also $(-\infty, -\sqrt{3} ] \cup [ \sqrt{3}, \infty)$. \\



\textbf{\textcolor{blue!55!black}{Zeros:}} \\


The zeros of $H$ are the domain numbers of $H$ where $H$ equals $0$.  This would be where $f = 0$, since the values of $H$ are values of $f$. \\



$f(t) = \sqrt{t}$ equals $0$ when $t = 0$. Therefore, we are looking for when the output of $g$ equals $0$. \\



$g(y) = y^2 - 3 = 0$ at $-\sqrt{3}$ and $\sqrt{3}$. \\



Or, we could just work with the entire formula for $H$. \\


\[
5 \sqrt{x^2-3} = 0
\]


\[
\sqrt{x^2-3} = 0
\]


\[
x^2-3 = 0
\]

\[
x^2 = 3
\]


We have two zeros, $-\sqrt{3}$ and $\sqrt{3}$. \\

We just have to make sure these are actually in the domain of $H$.  They are. \\






\textbf{\textcolor{blue!55!black}{Continuity:}} \\


$H$ is a composition of a root function and a quadratic funciton.  Both of these are continuous, which means their composition is continuous. \\

Neither root functions nor quadratics have discontinuites or singularities. \\





\textbf{\textcolor{blue!55!black}{End-Behavior:}} \\



We'll think of this in two steps. First, we'll determine the end-behavior of $g$, then we'll determine the end-behavior of $f$ with respect to the end-behavior of $g$.\\


Since $g$ is a quadratic function with a negative lading coeeficient, we know that 

\[
\lim\limits_{y \to -\infty} g(y) = \lim\limits_{x \to -\infty} g(y)  = \infty
\]


\[
\lim\limits_{y \to \infty} g(y) = \lim\limits_{x \to \infty} g(y)  = \infty
\]

This means the limit of $H$ is the limit of $f$ as the domain of $f$ approaches $\infty$. \\


$f$ is a sqaure root function with positive leading coefficients inside and out. \\

\[
\lim\limits_{x \to \infty} H(x) =  \lim\limits_{t \to \infty} f(t)  = \lim\limits_{t \to \infty} 5\sqrt{t} = \infty
\]


\[
\lim\limits_{-x \to \infty} H(x) =  \lim\limits_{t \to \infty} f(t)  = \lim\limits_{t \to \infty} 5\sqrt{t} = \infty
\]









\textbf{\textcolor{blue!55!black}{Behavior:}} \\


We want to figure out where $H$ is increasing and decreasing. \\

$H$ is a composition, so we need to know how the component functions behave. \\





$f(t) = 5 \sqrt{t}$ is an increasing function, since it is a root function with positive leading coefficients inside and outside. \\

$g(y) = y^2 - 3$ is a quadratic with a positive leading coefficient.  Therefore, it decreases and then increases.   It switches at its criticial number.\\



\[
\frac{-b}{2 a} = \frac{0}{2} = 0
\]


$g$ decreases on $(-\infty, 0)$ and increases on $(0, \infty)$.  However, $0$ is not in the domain of $H$. We need to restrict this information to the actual domain of $H$. \\





On $(-\infty, -\sqrt{3}]$, 


$g$ is decreasing. Here, the range of $g$ is positive and $f$ is increasing for positive domain numbers. \\ 



\begin{center}
$H$ is $increasing \circ decreasing$, which is decreasing. \\
\end{center}





On $[\sqrt{3}, \infty)$, 


$g$ is increasing. Here, the range of $g$ is positive and $f$ is increasing for positive domain numbers. \\ 


\begin{center}
$H$ is $increasing \circ increasing$, which is increasing. \\
\end{center}













\textbf{\textcolor{blue!55!black}{Local Maximum and Minimum:}} \\





On $(-\infty, -\sqrt{3}]$, $H$ is $decreasing$, which makes $H(-\sqrt{3}) = 0$ a local minimum.  \\


On $[\sqrt{3}, \infty)$, $H$ is $increasing$, which makes $H(\sqrt{3}) = 0$ a local minimum.  \\



There are no other critical numbers or endpoints in the domain.









\textbf{\textcolor{blue!55!black}{Global Maximum and Minimum:}} \\



On $(-\infty, -\sqrt{3}]$, $H$ is $decreasing$, which makes $H(-\sqrt{3}) = 0$ a local minimum. \\


On $[\sqrt{3}, \infty)$, $H$ is $increasing$, which makes $H(\sqrt{3}) = 0$ a local minimum  \\




Since $5 \sqrt{x^2-3} \geq 0$, $0$ is the global minimum. \\



$0$ is the global minimum value of $H$. It occurs at two places in the domain, $-\sqrt{3}$ and $\sqrt{3}$. \\



Since, $\lim\limits_{x \to -\infty} H(x) = \infty$, there is no global maximum. 














\textbf{\textcolor{blue!55!black}{Range:}} \\



$H$ is continuous and decreasing on $( -\infty, -\sqrt{3} ]$. \\


$\lim\limits_{x \to -\infty} H(x) = \infty$ \\


$H(-\sqrt{3}) = 0$


Therefore, the range of $H$ on $( -\infty, -\sqrt{3} ]$ is $[0, \infty)$. \\





$H$ is continuous and increasing on $[ \sqrt{3} ,\infty)$. \\


$\lim\limits_{x \to \infty} H(x) = \infty$ \\


$H(\sqrt{3}) = 0$


Therefore, the range of $H$ on $[ \sqrt{3}, \infty )$ is $[0, \infty)$. \\








The range of $H$ is $[0, \infty) \cup [0, \infty) = [0, \infty)$.




This all agrees with the graph. \\





\begin{image}
\begin{tikzpicture}
    \begin{axis}[name = sinax, domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$x$, ylabel={$y=f(x)$}, grid = major, grid style={dashed},
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            yticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$}, 
            xticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
          \addplot [line width=2, penColor, smooth,samples=100,domain=(-2.5:-1.7),<-] ({x},{5*sqrt(x^2 - 3)});
          \addplot [color=penColor,only marks,mark=*] coordinates{(-1.7,0)};


          \addplot [line width=2, penColor, smooth,samples=100,domain=(1.7:2.5),->] ({x},{5*sqrt(x^2 - 3)});
          \addplot [color=penColor,only marks,mark=*] coordinates{(1.7,0)};

           

    \end{axis}

\end{tikzpicture}
\end{image}

































\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Composition]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/composition/examples/exampleList}

\end{center}







\end{document}
