\documentclass{ximera}

\input{../../preamble.tex}

\title{Composition}

\begin{document}

\begin{abstract}
%
\end{abstract}
\maketitle








Our goal in Calculus is to analyze functions.


This means describing the function's behavior.

Calculus will provide procedures for obtaining the derivative of any function.  However, for compositions, we can get the same behavior information right now.






\begin{idea} \textbf{\textcolor{blue!55!black}{Composition}}  


Let $f$ be a function with domain $D_f$. 


Let $g$ be a function with domain $D_g$.




Then the \textbf{composition of f and g, $f \circ g$,} is defined as

\[
(f \circ g)(x) = f(g(x))
\] 


The value of $g$ becomes a domain number for $f$.


The composition is defined on a subset of the domain of $g$.  The composition is defined at those numbers in the domain of $g$ where the value of $g$ is in the domain of $f$.




\end{idea}


The Chain Rule is our tool of choice when it comes to the behavior of a composition.

The Chain Rule tells us how to decipher the behavior of a composition function from the behavior of its component functions.






\subsection*{Learning Outcomes}


\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item view functions as compositions.
\item deduce behavior of compositions.
\end{itemize}
\end{sectionOutcomes}












\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Composition]{https://ximera.osu.edu/csccmathematics/precalculus2/precalculus2/composition/examples/exampleList}

\end{center}







\end{document}
