\documentclass{ximera}

\input{../../preamble.tex}

\title{Tangent Lines}

\begin{document}

\begin{abstract}
zooming in
\end{abstract}
\maketitle


 
 


\subsection*{Tangent Lines}



Suppose we have the quadratic function $Q(x) = \frac{1}{2} (x - 2)^2 + 1$. 


Then its graph is a parabola. The point $\left( 3, \frac{3}{2} \right)$ is on the parabola.


The line $y=x-\frac{3}{2}$  goes through the point $\left( 3, \frac{3}{2} \right)$.

The parabola and line have a special relationship, which we can see by zooming in.





\begin{image}
\begin{tikzpicture}
     \begin{axis}[
                domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
                axis lines =center, xlabel=$x$, ylabel=$y$,
                ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
                xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
                ticklabel style={font=\scriptsize},
                every axis y label/.style={at=(current axis.above origin),anchor=south},
                every axis x label/.style={at=(current axis.right of origin),anchor=west},
                axis on top,
                ]


          \addplot [draw=penColor, very thick, smooth, domain=(-2:6),<->] {0.5*(x-2)^2 + 1};
        
          \addplot [draw=penColor2, very thick, smooth, domain=(0:8),<->] {x-1.5};

          \addplot [color=penColor,only marks,mark=*] coordinates{(3,1.5)};



    \end{axis}
\end{tikzpicture}
\end{image}





\begin{image}
\begin{tikzpicture}
     \begin{axis}[
                domain=2:3.5, ymax=2, xmax=3.5, ymin=0.5, xmin=2,
                axis lines =center, xlabel=$x$, ylabel=$y$,
                ytick={0.6,0.8,1,1.2,1.4,1.6,1.8},
                xtick={2.2,2.4,2.6,2.8,3,3.2,3.4},
                ticklabel style={font=\scriptsize},
                every axis y label/.style={at=(current axis.above origin),anchor=south},
                every axis x label/.style={at=(current axis.right of origin),anchor=west},
                axis on top,
                ]


          \addplot [draw=penColor, very thick, smooth, domain=(2.2:3.4),<->] {0.5*(x-2)^2 + 1};
        
          \addplot [draw=penColor2, very thick, smooth, domain=(2.2:3.4),<->] {x-1.5};

          \addplot [color=penColor,only marks,mark=*] coordinates{(3,1.5)};
        



    \end{axis}
\end{tikzpicture}
\end{image}









\begin{image}
\begin{tikzpicture}
     \begin{axis}[
                domain=2.7:3.2, ymax=1.7, xmax=3.2, ymin=1.2, xmin=2.7,
                axis lines =center, xlabel=$x$, ylabel=$y$,
                ytick={1.3,1.4,1.5,1.6},
                xtick={2.8,2.9,3,3.1,3.2},
                ticklabel style={font=\scriptsize},
                every axis y label/.style={at=(current axis.above origin),anchor=south},
                every axis x label/.style={at=(current axis.right of origin),anchor=west},
                axis on top,
                ]


          \addplot [draw=penColor, very thick, smooth, domain=(2.7:3.2),<->] {0.5*(x-2)^2 + 1};
        
          \addplot [draw=penColor2, very thick, smooth, domain=(2.7:3.2),<->] {x-1.5};

          \addplot [color=penColor,only marks,mark=*] coordinates{(3,1.5)};
        



    \end{axis}
\end{tikzpicture}
\end{image}










\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{bd3xxzj4rl}{400}{300}
\end{center}






As you zoom in, the graph slowly looks more and more like the line. 


The line does the best job of approximating the graph at the point $\left( 3, \frac{3}{2} \right)$. 


And, this is the only line that will work for the point $\left( 3, \frac{3}{2} \right)$.  Any other line, besides this one, will have a permanent (nonzero) angle between the graph and the line.  

The graph and this particular line have the same ``slope'' at the point. 


This line is called the \textbf{\textcolor{blue!55!black}{tangent line}} for $Q(x) = \frac{1}{2} (x - 2)^2 + 1$ at the point $\left( 3, \frac{3}{2} \right)$.




\begin{definition} \textbf{\textcolor{green!50!black}{Tangent Line}}


Suppose $C$ is a curve, like the graph of a function. 

Let $(a, b)$ be a point on the curve.

The line tangent to the curve, $C$, at the point $(a, b)$ is the line that goes through the point $(a, b)$ and does the best job of approximating the curve at the point $(a, b)$.

This line is called a \textbf{tangent line} at the point $(a, b)$.

This line is \textbf{tangent} to the curve at the point $(a, b)$.


\end{definition}


Since ``slope'' is a characteristic of a line and most curves are not lines, curves do not have a slope as we know that word. 

However, it is easy to see that curves have something that seems a lot like a slope at points. 


\begin{center}
\textbf{\textcolor{blue!55!black}{We will adopt the slope of the tangent line as the slope of the curve at the tangent point. }}  
\end{center}

As the parabola curve above demonstrates, curves might have different slopes at different points. 

We want to quantize this idea of a slope for a curve at a point. 

If we can quantize it, then we can build a function for it and use that to measure slopes for our curve. This is the first step toward Calculus. It will take us a while to quantize this idea of slope for a curve into a logical tool. 



\begin{center}
\textbf{\textcolor{blue!55!black}{A tool we will call the \textbf{derivative}.}}  
\end{center}



We will first investigate this idea with quadratic functions and parabolas. 








\textbf{\textcolor{blue!55!black}{$\blacktriangleright$ desmos graph}} 
\begin{center}
\desmos{tdkxeewwcj}{400}{300}
\end{center}


















\begin{center}
\textbf{\textcolor{green!50!black}{ooooo-=-=-=-ooOoo-=-=-=-ooooo}} \\

more examples can be found by following this link\\ \link[More Examples of Quadratic Behavior]{https://ximera.osu.edu/csccmathematics/precalculus1/precalculus1/quadraticBehavior/examples/exampleList}

\end{center}






\end{document}




