\documentclass{ximera}

\input{../../preamble.tex}

\title{Indeterminate}

\begin{document}

\begin{abstract}
equally fast
\end{abstract}
\maketitle





Functions can dominate other functions, or they can approach infinity similarly.  If the two functions approach infinity at the same rate, then the limit of their quotient equals a number.  The graph of the quotient approaches a horizontal asymptote.






\begin{example}

Consider the function $B(t) = \frac{3e^t - 5}{e^t + 1}$. \\

As $t \to \infty$, the exponential pieces will dominate the constant pieces.  

\[   B(t) = \frac{3e^t - 5}{e^t + 1} \sim \frac{3 e^t}{e^t} = 3   \]




As $t \to -\infty$, we have a different story.  $\lim\limits_{t \to -\infty} e^t = 0$  In this case, the constant terms dominate.


\[   B(t) = \frac{3e^t - 5}{e^t + 1} \sim \frac{-5}{1} = -5   \]


The graph has two different horizontal asymptotes.






\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-10:10, ymax=10, xmax=10, ymin=-10, xmin=-10,
            axis lines =center, xlabel=$t$, ylabel={$y=B(t)$}, grid = major,
            ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            yticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$}, 
            xticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          
            \addplot [line width=1, gray, dashed,samples=200,domain=(-9:9),<->] {3};
            \addplot [line width=1, gray, dashed,samples=200,domain=(-9:9),<->] {-5};

            \addplot [line width=2, penColor2, smooth,samples=100,domain=(-8:8),<->] {(3*(e^x) - 5)/(e^x + 1))};
 

  \end{axis}
\end{tikzpicture}
\end{image}






\end{example}








\section{Indeterminate Forms}


We have been examining functions that approach infinity at the same rate. When combined into a new quotient funciton, this new funciton might have a contant end-behavior.  The problem is that a fraction of the form $\frac{very \, big}{very \, big}$ could equal anything.



\[  \frac{10000000000}{2000000000} = 50      \]

\[  \frac{1000000000}{2000000000} = \frac{1}{2}      \]

\[  \frac{1000000000}{20000000000} = 0.2      \]





The same thing happens with values near $0$


\[  \frac{0.000000001}{0.0000000002} = 50      \]

\[  \frac{0.000000001}{0.000000002} = \frac{1}{2}      \]

\[  \frac{0.0000000001}{0.0000000002} = 0.2      \]





Formulas whose values seem like they are approaching $\frac{\infty}{\infty}$ or $\frac{0}{0}$ are called \textbf{indeterminate forms}, because you can't easily determine their values.






\begin{example}  Indeterminate Forms

Both the functions $S(t) = \sqrt{2t^3 - 5}$ and $L(t) = t\sqrt{9t+4}$ approach $\infty$ as $t \to \infty$. \\



$R(t) = \frac{S(t)}{L(t)} = \frac{\sqrt{2t^3 - 5}}{t\sqrt{9t+4}}$ is an indeterminate form. It is approaching the form $\frac{\infty}{\infty}$ \\


If we examine the quotient in terms of dominance, then we think like...

\[
R(t) =  \frac{\sqrt{2t^3 - 5}}{t \sqrt{9t+4}} \sim \frac{\sqrt{2t^3}}{t \sqrt{9t}} = \frac{t \sqrt{2t}}{t \sqrt{9t}} = \frac{\sqrt{2}}{\sqrt{9}}
\]



\[
\lim\limits_{t \to \infty} R(t) = \frac{\sqrt{2}}{3}
\]

\end{example}






























\section{Calculus}


We are introducing the term \textbf{indeterminate form} here in connection to dominance and horizontal asymptotes.  In Calculus, this idea will be expanded significantly to include other targets besides $\infty$.




\begin{example} Indeterminate Forms

Both the functions $S(t) = sin(t)$ and $L(t) = t$ approach $0$ as $t \to 0$. \\

Do they approach $0$ at the same rate? \\

Let's investigate $R(t) = \frac{sin(t)}{t}$










\begin{image}
\begin{tikzpicture}
  \begin{axis}[
            domain=-1:1, ymax=2, xmax=1, ymin=0, xmin=-1,
            axis lines =center, xlabel=$t$, ylabel={$y=R(t)$}, grid = major,
            %ytick={-10,-8,-6,-4,-2,2,4,6,8,10},
            %xtick={-10,-8,-6,-4,-2,2,4,6,8,10},
            %yticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$}, 
            %xticklabels={$-10$,$-8$,$-6$,$-4$,$-2$,$2$,$4$,$6$,$8$,$10$},
            ticklabel style={font=\scriptsize},
            every axis y label/.style={at=(current axis.above origin),anchor=south},
            every axis x label/.style={at=(current axis.right of origin),anchor=west},
            axis on top
          ]
          

            \addplot [line width=2, penColor2, smooth,samples=100,domain=(-8:8),<->] {sin(deg(x))/x};





           

  \end{axis}
\end{tikzpicture}
\end{image}







$sin(t)$ and $t$ approach $0$ at exactly the same rate.  Their quotient approaches $1$.



\end{example}
















\begin{example} Indeterminate Forms

We have seen an even weirder indeterminate form.

Everyone knows that $1$ to any power equals $1$. Everyone knows that numbers greater than $1$ to any power get bigger.


What about something like $\left( 1 + \frac{1}{x} \right)^x$ ?

The base is getting closer to $1$, but they are always greater than $1$.  The power is growing larger.  

We are heading to something like $1^{\infty}$.

And we have seen this function actually approaches the value $e \approx 2.71828$. \\

How far out in the domain do you need to go for the graph to be inside the interval $[2.7, 2.8]$?








\begin{center}
\desmos{ttk7altjct}{400}{300}
\end{center}









\end{example}












\end{document}
