\documentclass{ximera}

\input{../../preamble.tex}

\title{Implicit Functions}

\begin{document}

\begin{abstract}
%Stuff can go here later if we want!
\end{abstract}
\maketitle








We have been investigating functions.  More specifically, we have been investigating functions of one variable - all of our formulas have one variable.

We have investigated the graphs of these functions.  The graph of a function is a collection of points in the Cartesian plane.  If $f$ is the function, then each point belonging to the graph is of the form $(d, f(d))$, where $d$ is a member of the domain of $f$.

Our functions have domains which are subsets of the real numbers and they take on real number values.  Our notation for this looks like 

\[  f : \mathbb{R} \mapsto \mathbb{R}      \]



\begin{center}
People say, "$f$ maps $\mathbb{R}$ to $\mathbb{R}$."
\end{center}




We can view all of this in a different light. \\




Rather than thinking that the function pairs real numbers with real number, let's thinking of a function that takes real numbers to points in the Cartesian plane.


We might think of the function $F$: $F(d) = (d, f(d))$.  It maps a single real number to an ordered paired of real numbers.



\[  F : \mathbb{R} \mapsto \mathbb{R^2}      \]



When $f$ is a continuous function, then the image of $F$ is called a \textbf{curve}.


These curves look a lot like the graphs of functions, because graphs of functions are special types of curves.  However, generally speaking, curves are not graphs of functions.  They do not pass the vertical line test. \\

However, we might be able to snip off a piece of the graph, which would serve as a graph of a function.  When we are thinking this way, people say we are thinking \textbf{implicitly}.







\begin{sectionOutcomes}
In this section, students will 

\begin{itemize}
\item compare curves and functions.
\item investigate functions implicitly.
\end{itemize}
\end{sectionOutcomes}

\end{document}
